---
title: "Data Filtering"
author: "Alex Depew"
date: "2025-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is for filtering the overall database for subsequent analyses
Files used described below:

- `Comprehensive_Bison_Database_Metadata_18.12.2025.csv`
   - Contains all bison microsatellite data from wood and plains bison with added metadata for year collected


#### Clear environment and load required libraries
```{r, warning = FALSE, message = FALSE}
rm(list = ls())
library(tidyverse)
library(adegenet)
library(strataG)
```

#### Load required data
```{r}
fulldf <- read.csv("Comprehensive_Bison_Database_Metadata_18.12.2025.csv",
                         check.names = F)
```

## Global data cleaning and organization

#### Filter out sex/X-linked loci
Loci BM6017 and BMS911 are X-linked, and must be removed for analyses
```{r}
# Add a name to the first column name
colnames(fulldf)[1] <- "Spreadsheet_ID"

# Remove loci 
fulldf_50 <- fulldf %>% 
  dplyr::select(-BM6017, -BM6017_2, -BMS911, -BMS911_2, -SE4748, -SE4748_2)
```

#### Herd naming consistency
Personal correspondence with Greg Wilson indicated that WBNP Carlson is likely Hay Camp, and WBNP Lake One is likely Garden River.
Other herds must have their names made to be consistent with the remainder of the samples.
```{r}
fulldf_50 <- fulldf_50 %>%
  mutate(HERD = case_when(
    HERD == "WBNPCarlson" ~ "WBNPHC",
    HERD == "WBNPLakeOne" ~ "WBNPGR",
    HERD == "NORD/Liard" ~ "NORD",
    HERD == "Ronald Lake" ~ "RL",
    HERD == "Delta" & SUBSPECIES == "Wood Bison" ~ "WBNPPAD",
    HERD == "AB?" ~ "RL",
    HERD == "Yukon" ~ "AISH",
    TRUE ~ HERD
))
```

#### Profile completeness metric
Downstream stringency filtering will select only samples with well amplified profiles. 
```{r Profile Completeness}
  fulldf_50 <- fulldf_50 %>%
  # Convert "", "**", or "0" to NA for all character columns
  mutate(across(
    where(is.character),
    ~ replace(.x, .x %in% c("", "**", "0"), NA)
  )) %>%
  
  # Calculates a row sum of non-NA values (% of profile amplified)
  mutate(COMPLETE = rowSums(!is.na(across(16:115)))) %>%
  
  # Moves newly created column with profile completeness (%) after AMPLIFICATION
  relocate(COMPLETE, .after = AMPLIFICATION)
```

#### Wood Bison Subsetting
```{r}
# Vector containing herds for analysis
target_herds <- c("EINPP",
                  "EINPW",
                  "WBNP",
                  "WBNPHC",
                  "WBNPGR",
                  "WBNPNY",
                  "WBNPPAD",
                  "WBNPLB",
                  "NORD",
                  "NAH",
                  "ETH",
                  "MBS",
                  "HZ",
                  "WENT",
                  "AISH",
                  "RL",
                  "WAB",
                  "CL")

wb_df <- fulldf_50 %>% 
  filter(HERD %in% target_herds)

```

## Duplicate ID Removal
Samples genotyped at Trent University have had duplicate genotypes identified usingS GENECAP. In the data file, samples identified as originating from the same individual are denoted with unique alphabetic identifiers per individual in the "INDIVIDUAL" column. Samples containing these pre-identified alphabetic individual identifier must be filtered, so the dataset contains just one instance of each individual. However, profiles will be manually checked to ensure:
- A. Maximum heterozygosity (fecal samples are more prone to null alleles and allelic dropout)
- B. Profile completeness (fuller profiles are more likley to yield better quality data)
- C. Sample type (prefer tissue samples over hair/fecal)
```{r Duplicate Individual Removal}
# Dataframe with individuals pre-identified as duplicate profiles
duplicatedid <- wb_df[!is.na(wb_df$INDIVIDUAL), ]

# Write out to csv for manual selection
write.csv(duplicatedid, "dup.sample.selection.csv", row.names = F)

# Dataframe without pre-identified duplicate profiles 
non_duplicatedid <- wb_df[is.na(wb_df$INDIVIDUAL), ]

# Import the dataframe with manually selected samples to keep
sample_selection <- read.csv("selected_dupes.csv",
         check.names = F)

# Make a dataframe containing the samples that have been chosen to keep
sample_keep <- sample_selection %>% 
  filter(KEEP == "TRUE")

# Vector containing positions in the spreadsheet (samples) to keep. This is done as some of the sample IDs themselves are duplicates, and they would not be identical matches
keep_vec <- sample_keep$Spreadsheet_ID

# Filter the duplicated ID dataframe to just contain the samples as identified
dup_fix <- duplicatedid %>% 
  filter(Spreadsheet_ID %in% keep_vec)

# Bind the fixed duplicated and non-duplicated ID dataframes back together
wb_fixed_ids <- rbind(dup_fix, non_duplicatedid)

which(duplicated(dup_fix$INDIVIDUAL))
```
#### Profile Coverage Filtering
As per Zimmerman et al. 2025, remove samples containing missing data at more than 4 loci (ie. >= 46 loci, or 92%). Filtering to include "usuable" high-amplification samples at this step allow for accurate duplicate filtering by genotypes in later steps.
## Plot

```{r}
summary(wb_fixed_ids$COMPLETE)
```


```{r}
p <- ggplot(wb_fixed_ids, aes(x = COMPLETE)) +
  geom_bar(fill = "grey40", width = 1.5) +
  labs(
    x = "% Profile Amplified",
    y = "Number of Samples"
  ) +
  theme_bw(base_size = 12) +
  theme(

    axis.title = element_text(size = 10)
  )

ggsave(
  filename = "plots/wb_profile_completeness_distribution.png",
  plot = p,
  width = 100,      
  height = 70,
  units = "mm",
  dpi = 600
)
```

```{r Profile Coverage Stringency}
wb_qual <- wb_fixed_ids %>% 
  filter(COMPLETE >= 92)

wb_low_qual <- wb_fixed_ids %>% 
  filter(COMPLETE <92)

table(wb_low_qual$HERD)
table(wb_low_qual$YEAR.COLLECTED, wb_low_qual$HERD)
sum(grepl("Thesis", wb_low_qual$YEAR))
wb_low_noM <- wb_low_qual %>% 
  filter(!grepl("Thesis", wb_low_qual$YEAR))
table(wb_low_noM$YEAR.COLLECTED, wb_low_noM$HERD)
```

## Year Assignments
```{r}
sum(is.na(wb_qual$YEAR.COLLECTED))
```
#### Identify patterns in year
```{r}
# Make a new database containing columns of YEAR.ASSIGN, which (at first) is identical to YEAR.COLLECTED, and will be populated with inferences of patterns in sample IDs
year_manipulation <- wb_qual %>% 
  mutate(
    YEAR.ASSIGN = YEAR.COLLECTED,
    YEAR.BIN = "", 
  ) %>% 
  relocate(YEAR.ASSIGN, YEAR.BIN, .after = YEAR.COLLECTED)

# Write out to csv to identify patterns in sample naming
write.csv(year_manipulation, "year_pattern_assign.csv", row.names = F)
```

Almost 1/3 (318) of samples do not have year collection metadata. A file containing samples used in the Ball et al. 2016 publication provides information on what was collected, but only as specific as to pre-2016.

#### Ball et al. 2016 Sample List
A list of samples used in the Ball et al. 2016 publication was provided for insight into collection year
```{r}
# Import list of genotypes used in Ball 2016 - 
ball_samples <- read.table("GWilson_goldgenotypes_AB sample info list.txt",
                           sep = "\t", header = T)

ball_samples <- ball_samples %>% 
  
  # Fix the ID names to match what is present in the database
  mutate(
    IDrework = ID %>%
      gsub("^Bison-", "", .) %>%
      gsub("_", "-", .) %>% 
      gsub("RainbowLake1", "rainbow lake 1", .) %>% 
      gsub("Chinchaga-1", "chinchaga 1", .) %>% 
      gsub("Chinchaga-2", "chinchaga 2", .) %>% 
      gsub("RonaldLakeFetus", "ronald lake fetus", .)
  ) %>% 
  relocate(IDrework, .after = ID) %>% 
  
  # Make herd names consistent with the abbreviations in the main database
  mutate(
    HERD = case_when(
      Population == "RonaldLake"   ~ "RL",
      Population == "NWT"          ~ "MBS",
      Population == "Rainbow Lake" ~ "HZ",
      Population == "Wentzel"      ~ "WENT",
      Population == "Harpercreek"  ~ "WAB",
      Population == "Hay zama"     ~ "HZ",
      TRUE ~ Population
    )
  ) %>% 
  relocate(HERD, .after = Population)
```

The goal is to bin these samples pre-2016
```{r}
# How many samples can be successfully matched between the Ball et al. 2016 sample list, and the dataset to determine collection year bin in samples with unknown collection years?

years_unknown <- year_manipulation %>% 
  filter(is.na(YEAR.COLLECTED))

sum(years_unknown$SAMPLE.ID %in% ball_samples$IDrework & years_unknown$HERD %in% ball_samples$HERD)
```
Therefore 72 of the samples with unknown collection years can be assigned a pre-2016 collection bin.

From the Ball et al. 2016 sample list, the arbitrary year "2004" may be assigned to their year assignment, as no other samples originate from 2004. 
```{r}
year_manipulation_added_ball <- year_manipulation %>% 
  mutate(
    YEAR.ASSIGN = replace(
      YEAR.ASSIGN,
      is.na(YEAR.ASSIGN) &
      SAMPLE.ID %in% ball_samples$IDrework &
      HERD %in% ball_samples$HERD,
      2015    
    )
  )

# Confirm 72 samples were changed
sum(year_manipulation_added_ball$YEAR.ASSIGN == 2015, na.rm = TRUE)
```
# Ball plains bison
```{r}
ball.EINPW.WBNP <- read.csv("goldgenotypes_greg_Parks.csv",
                      sep = ",",
                      header = T)

ball.EINPP <- read.csv("plains bison einp.csv",
                      sep = ",",
                      header = T)

ball.EINPP <- ball.EINPP %>% 
  mutate(
    ID_rework = gsub("-", "", ID)
  ) %>% 
  relocate(ID_rework, .after = ID)


sum(year_manipulation_added_ball$SAMPLE.ID %in% ball.EINPP$ID_rework)
```


#### Sample Naming Schemes
"W" and "WR" naming conventions used at EINP for both wood and plains bison indicate tagging year
```{r}
year_manipulation_added_GW <- year_manipulation_added_ball %>%
  mutate(
    # Handle WR##### IDs
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & grepl("^WR\\d{5}", SAMPLE.ID),
      {
        two_digits <- str_sub(str_remove(SAMPLE.ID, "^WR"), 1, 2)
        two_digits <- as.numeric(two_digits)
        ifelse(two_digits >= 50, 1900 + two_digits, 2000 + two_digits)
      },
      YEAR.ASSIGN
    ),
    
    # Handle W##### IDs
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & grepl("^W\\d{5}", SAMPLE.ID),
      {
        two_digits <- str_sub(str_remove(SAMPLE.ID, "^W"), 1, 2)
        two_digits <- as.numeric(two_digits)
        ifelse(two_digits >= 50, 1900 + two_digits, 2000 + two_digits)
      },
      YEAR.ASSIGN
    )
  )

sum(grepl("^WR\\d{5}$|^W\\d{5}$", year_manipulation_added_ball$SAMPLE.ID))
```
#### Sample Naming Patterns
When reviewing the whole dataset, samples without collection year information sometimes have SAMPLE.ID which is consistent with samples containing collection year information. It is reasonable to assume that these samples may originate from the same sampling event, and therefore can be assigned a year. Some samples have indicators of collection year within their SAMPLE.ID ex.
- 25WBNP_19BF1 likely was collected in 2025
- 2009-2238 likely was collected in 2009

```{r}
year_manipulation_added_IDpatterns <- year_manipulation_added_GW %>% 
  mutate(
    # Add year collection for 25WBNP samples
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & grepl("^25WBNP", SAMPLE.ID),
      2025,
      YEAR.ASSIGN
    ),
    
    # Add year collection for 2009 samples
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & grepl("^2009", SAMPLE.ID),
      2009,
      YEAR.ASSIGN
    ),
    
    # Add year collection to specific samples based on 
    YEAR.ASSIGN = case_when(
      SAMPLE.ID == "YUK-223-Fe" ~ 2022,
      SAMPLE.ID == "P86064" ~ 1986,
      SAMPLE.ID == "P85033" ~ 1985,
      SAMPLE.ID == "P85030" ~ 1985,
      SAMPLE.ID == "P84030" ~ 1984,
      SAMPLE.ID == "86063" ~ 1986,
      SAMPLE.ID == "85029" ~ 1985,
      TRUE ~ YEAR.ASSIGN
    ),
    
    # Add years from Ball et al. 2016
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) &
      HERD == "EINPW" &
      between(
        as.numeric(str_extract(SAMPLE.ID, "\\d+")),
        88, 149
      ),
      2015,
      YEAR.ASSIGN
    ),
    # Add years from Wilson 1999
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) &
      HERD == "EINPP" &
      between(
        as.numeric(str_extract(SAMPLE.ID, "\\d+")),
        414, 478
      ),
      2015,
      YEAR.ASSIGN
    ),
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) &
      grepl("^Mackenzie Bison Sanctuary", SAMPLE.ID),
      1999,
      YEAR.ASSIGN
    ),
      # Add year collection for Wentzel 2025 samples
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & HERD == "WENT" & grepl("^BISO25", SAMPLE.ID),
      2025,
      YEAR.ASSIGN
    ),
    YEAR.ASSIGN = ifelse(
      is.na(YEAR.ASSIGN) & HERD == "EINPP" &grepl("2v", SAMPLE.ID, ignore.case = TRUE),
      1990,
      YEAR.ASSIGN
    )
)

wb_years_added <- year_manipulation_added_IDpatterns
```

## Sample Removal
Some samples have been identified as incompatible for downstream analyses. These include duplicate genotypes and samples containing nonsense data

#### Problem Samples
Wab24_011 needs to be removed (3 alleles)
```{r}
wb_years_added <- wb_years_added %>% 
  filter(SAMPLE.ID != "Wab24_011")
```

#### Duplicate Genotypes
Use package adegenet and strataG to identify duplicates across the whole dataset by their genotypes. Before using these packages, reformatting of the allele data must occur.

```{r}
# Need to import dataframe as a gtypes object. For this, columns need to be merged.
locus_cols <- names(wb_years_added)[19:ncol(wb_years_added)]

# Get base locus names without "_2"
base_loci <- unique(gsub("_2$", "", locus_cols))

# Loop through each locus and combine alleles
for (locus in base_loci) {
  
  col1 <- locus
  col2 <- paste0(locus, "_2")
  
  if (col1 %in% names(wb_years_added) && col2 %in% names(wb_years_added)) {
    wb_years_added[[col1]] <- paste(wb_years_added[[col1]], wb_years_added[[col2]], sep = "/")
  }
}

# Remove all the _2 columns
wb_years_added <- wb_years_added[, !grepl("_2$", names(wb_years_added))]

# Fix missing data values
wb_years_added[wb_years_added == "0/0"] <- NA
wb_years_added[wb_years_added == "**/**"] <- NA
wb_years_added[wb_years_added == "NA/NA"] <- NA

# Import as a gtypes object with strataG package
wb_dup.genind <- df2genind(X=wb_years_added[,c(19:ncol(wb_years_added))],
                              sep = "/",
                              ind.names = wb_years_added$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = wb_years_added$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Convert to a gtypes object
bison.gtype <- genind2gtypes(wb_dup.genind)
```

Determine genotype duplicates with an 80% profile concordance or greater
```{r}
dupGenotypes(bison.gtype, num.shared = 0.8)
```

We can see that strataG called 16 samples which share over 80% profile similarity - ie. 8 pairs of profiles
```{r}
# Isolate samples identified as potential duplicates
dup_check <- wb_years_added[wb_years_added$Spreadsheet_ID %in% c("1080", "892",
                                                               "2371", "2388", 
                                                               "2368", "983",
                                                               "566", "894",
                                                               "533", "536",
                                                               "948", "959",
                                                               "2484", "2491",
                                                               "2486", "2595",
                                                               "1029", "2418",
                                                               "2508", "2521"), ]
# Write out to csv to manually inspect genotypes
write.csv(dup_check, "dup_check.csv", row.names= F)


# Following selection of best quality genotypes (as per same criteria used previously), reimport dataset with additional "KEEP" Column
dup_genotype_filter <- read.csv("dup_genotype_filter.csv", check.names = F)

# Make dataframe containing samples identified to keep
dup_removals <- dup_genotype_filter %>% 
  filter(
    is.na(KEEP)
  )

# Make vector containing IDs we want to remove
dup_removals_ids <- dup_removals$Spreadsheet_ID

# Now filter out the duplicate genotype samples from the original dataset
wb_cleaned_samples <- wb_years_added %>% 
  filter(
    !Spreadsheet_ID %in% dup_removals_ids
  )
```
At this stage, all samples have been cleaned and the dataset should have no duplicates.

#### Year Filtering
Following the year assignments, the removal of samples without collection year information is necessary
```{r}
# Remove any samples without year metadata
wb_cleaned_years <- wb_cleaned_samples %>% 
  filter(
    !is.na(YEAR.ASSIGN)
  )
```

Write the cleaned product out to a csv
```{r}
write.csv(wb_cleaned_years, "Cleaned_samples.06.01.2026.csv", row.names = F)
```



