---
title: "Genetic Diversity"
author: "Alex Depew"
date: "2026-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# File information
This script is used to create descriptive statistic tables for population genetic analyses of all wood bison herds in Canada. 

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```

## Load Required Data
```{r}
bison.data <- read.csv("./datafiles/Cleaned_locus_data_20.02.2026.csv",
                       check.names = F)
```

Merge WBNP Subherds
```{r}
mod.data <- bison.data %>% 
  mutate(HERD.2 = case_when(
    grepl("WBNPLB|WBNPNY", HERD) ~ "WBNPC1",
    grepl("WBNPPAD|WBNPGR|WENT", HERD) ~ "WBNPC2",
    TRUE                         ~ HERD  # This acts as the "else" (keep original name)
  )) %>% 
  relocate(HERD.2, .after = HERD)
```

#### Convert data to Genind object
```{r}
# Make genind object
merge.genind <- df2genind(X=mod.data[,c(20:ncol(mod.data))],
                              sep = "/",
                              ind.names = mod.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = mod.data$HERD.2,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(merge.genind), mod.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
merge.genind@other$YEAR.ASSIGN <- mod.data$YEAR.ASSIGN[idx]

# Make genind object
bison.genind <- df2genind(X=bison.data[,c(20:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]
```


#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Merged WBNP herds
m1 <- merge.genind[merge.genind@other$YEAR.ASSIGN <=1999,]
m1.sum <- summary(m1) # Create summary of genind object

m2 <- merge.genind[merge.genind@other$YEAR.ASSIGN >= 2000 & merge.genind@other$YEAR.ASSIGN <= 2015, ]
m2.sum <- summary(m2)

m3 <- merge.genind[merge.genind@other$YEAR.ASSIGN >=2016,]
m3.sum <- summary(m3)


# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)
```

# Descriptive Statistics 
Functions requiring the package *strataG* require inputs to be in "gtypes" objects. Programs such as HP-RARE require data to be in GENEPOP file format. 
```{r}
# Convert genind to gtypes
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)
temp3.g <- genind2gtypes(temp3)

# Merged data
m1.g <- genind2gtypes(m1)
m2.g <- genind2gtypes(m2)
m3.g <- genind2gtypes(m3)

# Convert genind to genepop
library(graph4lg)
genind_to_genepop(temp1, output = "genepop_files/temp1.genepop.txt")
genind_to_genepop(temp2, output = "genepop_files/temp2.genepop.txt")
genind_to_genepop(temp3, output = "genepop_files/temp3.genepop.txt")

# Hierfstat summary format
temp1.stats <- basic.stats(temp1, diploid = TRUE, digits = 3)
temp2.stats <- basic.stats(temp2, diploid = TRUE, digits = 3)
temp3.stats <- basic.stats(temp3, diploid = TRUE, digits = 3)

# Merged data
m1.stats <- basic.stats(m1, diploid = TRUE, digits = 3)
m2.stats <- basic.stats(m2, diploid = TRUE, digits = 3)
m3.stats <- basic.stats(m3, diploid = TRUE, digits = 3)
```

## Temporal Group 1
```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity
temp1.Ho <- round(colMeans(temp1.stats$Ho), 2) # Observed (Ho)
# HO SD
temp1.Ho.sd <- round(apply(temp1.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp1.He <- round(colMeans(temp1.stats$Hs), 2) # Expected (He)
# HE SD
temp1.He.sd <- round(apply(temp1.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp1.Ar <- allelic.richness(temp1, min.n = 10, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)
# AR SD
temp1.Ar.sd <- round(apply(temp1.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df1 <- enframe(temp1.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp1.He[Population], "±", temp1.He.sd[Population]), # Combine SD with ±
        HO = paste(temp1.Ho[Population], "±", temp1.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp1.Ar.mean[Population], "±", temp1.Ar.sd[Population]), # Combine with ±
        FIS = temp1.Fis
        )

print(descr.df1)
```
```{r}
# Number of samples per population
m1.N = m1.sum$n.by.pop
m1.N

# Heterozygosity
m1.Ho <- round(colMeans(m1.stats$Ho), 2) # Observed (Ho)
# HO SD
m1.Ho.sd <- round(apply(m1.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

m1.He <- round(colMeans(m1.stats$Hs), 2) # Expected (He)
# HE SD
m1.He.sd <- round(apply(m1.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
m1.Ar <- allelic.richness(m1, min.n = 10, diploid = TRUE)
m1.Ar.mean <- round(colMeans(m1.Ar$Ar), 2)
# AR SD
m1.Ar.sd <- round(apply(m1.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
m1.Fis <- round(colMeans(m1.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df.m1 <- enframe(m1.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(m1.He[Population], "±", m1.He.sd[Population]), # Combine SD with ±
        HO = paste(m1.Ho[Population], "±", m1.Ho.sd[Population]), # Combine SD with ±
        AR = paste(m1.Ar.mean[Population], "±", m1.Ar.sd[Population]), # Combine with ±
        FIS = m1.Fis
        )

print(descr.df.m1)
```

## Temporal Group 2
```{r}
# Number of samples per population
temp2.N = temp2.sum$n.by.pop
temp2.N

# Heterozygosity
temp2.Ho <- round(colMeans(temp2.stats$Ho), 2) # Observed (Ho)
# HO SD
temp2.Ho.sd <- round(apply(temp2.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp2.He <- round(colMeans(temp2.stats$Hs), 2) # Expected (He)
# HE SD
temp2.He.sd <- round(apply(temp2.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp2.Ar <- allelic.richness(temp2, min.n = 10, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)
# AR SD
temp2.Ar.sd <- round(apply(temp2.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df2 <- enframe(temp2.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp2.He[Population], "±", temp2.He.sd[Population]), # Combine SD with ±
        HO = paste(temp2.Ho[Population], "±", temp2.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp2.Ar.mean[Population], "±", temp2.Ar.sd[Population]), # Combine with ±
        FIS = temp2.Fis
        )

print(descr.df2)
```



## Temporal Group 3
```{r}
# Number of samples per population
temp3.N = temp3.sum$n.by.pop
temp3.N

# Heterozygosity
temp3.Ho <- round(colMeans(temp3.stats$Ho), 2) # Observed (Ho)
# HO SD
temp3.Ho.sd <- round(apply(temp3.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp3.He <- round(colMeans(temp3.stats$Hs), 2) # Expected (He)
# HE SD
temp3.He.sd <- round(apply(temp3.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp3.Ar <- allelic.richness(temp3, min.n = 10, diploid = TRUE)
temp3.Ar.mean <- round(colMeans(temp3.Ar$Ar), 2)
# AR SD
temp3.Ar.sd <- round(apply(temp3.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp3.Fis <- round(colMeans(temp3.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df3 <- enframe(temp3.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp3.He[Population], "±", temp3.He.sd[Population]), # Combine SD with ±
        HO = paste(temp3.Ho[Population], "±", temp3.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp3.Ar.mean[Population], "±", temp3.Ar.sd[Population]), # Combine with ±
        FIS = temp3.Fis
        )

print(descr.df3)
```

# Merged 3
```{r}
# Number of samples per population
m3.N = m3.sum$n.by.pop
m3.N

# Heterozygosity
m3.Ho <- round(colMeans(m3.stats$Ho), 2) # Observed (Ho)
# HO SD
m3.Ho.sd <- round(apply(m3.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

m3.He <- round(colMeans(m3.stats$Hs), 2) # Expected (He)
# HE SD
m3.He.sd <- round(apply(m3.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
m3.Ar <- allelic.richness(m3, min.n = 10, diploid = TRUE)
m3.Ar.mean <- round(colMeans(m3.Ar$Ar), 2)
# AR SD
m3.Ar.sd <- round(apply(m3.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
m3.Fis <- round(colMeans(m3.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df.m3 <- enframe(m3.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(m3.He[Population], "±", m3.He.sd[Population]), # Combine SD with ±
        HO = paste(m3.Ho[Population], "±", m3.Ho.sd[Population]), # Combine SD with ±
        AR = paste(m3.Ar.mean[Population], "±", m3.Ar.sd[Population]), # Combine with ±
        FIS = m3.Fis
        )

print(descr.df.m3)
```

Join Tables (Long)
```{r}
df1_v <- descr.df1 %>% mutate(Period = "T1")
df2_v <- descr.df2 %>% mutate(Period = "T2")
df3_v <- descr.df3 %>% mutate(Period = "T3")

dfm1_v <- descr.df.m1 %>% 
  filter(Population %in% c("WBNPC1", "WBNPC2")) %>% 
  mutate(Period = "T1")

dfm3_v <- descr.df.m3 %>% 
  filter(Population %in% c("WBNPC1", "WBNPC2")) %>% 
  mutate(Period = "T3")

full.descr.long <- bind_rows(df1_v, df2_v, df3_v, dfm1_v, dfm3_v) %>% 
  arrange(Population == "EINPP", Population) %>% 
  relocate(Period, .after = Population)

# Write to csv
write.table(full.descr.long, "./tables/Temporal_GD_Table_long.csv", 
            sep = ",",
            fileEncoding = "UTF-8", 
            row.names = FALSE)
```


```{r}
mRatio(temp1.g, by.strata = TRUE)
```


# Genetic Distance
#### Merge WBNP C1 and C2 to the temp1 and temp2 datasets
```{r}
WBNP_Clusters <- mod.data %>% 
  filter(HERD.2 %in% c("WBNPC1", "WBNPC2")) %>% 
  mutate(HERD = HERD.2) %>% 
  select(-HERD.2)

# Bind rows to full dataset for comparison
wbnp_clust.merged <- bison.data %>% 
  bind_rows(WBNP_Clusters)


# Make genind object
full.gen <- df2genind(X=wbnp_clust.merged[,c(20:ncol(wbnp_clust.merged))],
                              sep = "/",
                              ind.names = wbnp_clust.merged$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = wbnp_clust.merged$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
full.gen@other$YEAR.ASSIGN <- wbnp_clust.merged$YEAR.ASSIGN

# Subset into temporal groups
temp1_merge <- full.gen[full.gen@other$YEAR.ASSIGN <=1999]

temp2_merge <- full.gen[full.gen@other$YEAR.ASSIGN %in% 2000:2015]
# Remove Wentzel from this group
temp2_merge <- temp2_merge[!temp2_merge@pop == "WBNPC2"]

temp3_merge <- full.gen[full.gen@other$YEAR.ASSIGN>= 2016]
```


#### Weir and Cockerham (1984) Fst
```{r}
# Temporal group 1
## Order herds 
temp1_fst.mat <- as.matrix(pairwise.WCfst(temp1_merge,
                                        diploid = TRUE))

fst1 <- data.frame(round(temp1_fst.mat, 3))

# Temporal group 2
temp2_fst.mat <- as.matrix(pairwise.WCfst(temp2_merge,
                                          diploid = TRUE))

fst2 <- data.frame(round(temp2_fst.mat, 3))
write.csv(fst2, "./tables/fst2.csv", row.names = F)

# Temporal group 3
temp3_fst.mat <- as.matrix(pairwise.WCfst(temp3_merge, 
                                          diploid = TRUE))
fst3 <- data.frame(round(temp3_fst.mat, 3))
write.csv(fst3, "./tables/fst3.csv", row.names = F)


```


```{r}
# Reorder
current_names <- colnames(temp1_fst.mat)

# 2. Define the new order manually (swapping the last two)
new_order <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD", 
                "EINPW", "MBS", "EINPP", "WBNPC1", "WBNPC2")

# 3. Apply the swap to both rows and columns
temp1_fst.ordered <- temp1_fst.mat[new_order, new_order]

# View the result
round(temp1_fst.ordered, 3)
write.csv(temp1_fst.ordered, "./tables/fst1.csv", row.names = T)

# Reorder
current_names2 <- colnames(temp2_fst.mat)
current_names2

# 2. Define the new order manually (swapping the last two)
new_order2 <- c("HZ","MBS", "RL", "WAB", "WENT")

# 3. Apply the swap to both rows and columns
temp2_fst.ordered <- temp2_fst.mat[new_order2, new_order2]

# View the result
round(temp2_fst.ordered, 3)
write.csv(temp2_fst.ordered, "./tables/fst2.csv", row.names = T)

# Reorder
current_names3 <- colnames(temp3_fst.mat)
current_names3

# 2. Define the new order manually (swapping the last two)
new_order3 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD", "WENT", "AISH","CL", "EINPW", "ETH", "HZ", "NAH", "NORD", "RL", "WAB", "EINPP", "WBNPC1", "WBNPC2")

# 3. Apply the swap to both rows and columns
temp3_fst.ordered <- temp3_fst.mat[new_order3, new_order3]

# View the result
round(temp3_fst.ordered, 3)
write.csv(temp3_fst.ordered, "./tables/fst3.csv", row.names = T)


```

## Bootstraps of Fst
```{r}
boot1.mat <- boot.ppfst(temp1_merge, nboot = 1000, quant=c(0.025,0.975))
boot1.mat$ll

boot2.mat <- boot.ppfst(temp2_merge, nboot = 1000, quant=c(0.025,0.975))
boot2.mat$ll

boot3.mat <- boot.ppfst(temp3_merge, nboot = 1000, quant=c(0.025,0.975))
boot3.mat$ll

```
Fis Sig
```{r}
temp1_fis_boot <- boot.ppfis(dat=temp1_merge,nboot=1000,quant=c(0.025,0.975))
temp1_fis_boot$fis.ci

temp1_merge.stat <- basic.stats(temp1_merge)
colMeans(temp1_merge.stat$Fis)

temp2_fis_boot <- boot.ppfis(dat=temp2_merge,nboot=1000,quant=c(0.025,0.975))
temp2_fis_boot$fis.ci

temp2_merge.stat <- basic.stats(temp2_merge)
colMeans(temp2_merge.stat$Fis)

temp3_fis_boot <- boot.ppfis(dat=temp3_merge,nboot=1000,quant=c(0.025,0.975))
temp3_fis_boot$fis.ci

temp3_merge.stat <- basic.stats(temp3_merge)
colMeans(temp3_merge.stat$Fis)
```

```{r}
# HE
# Put your objects into a named list
all_temps <- list("T1" = temp1_merge.stat, "T2" = temp2.stats, "T3" = temp3_merge.stat)

# Define the metric to compare (Hs is Expected Heterozygosity)
metric <- "Hs"

# Define the possible pairs of time periods
period_pairs <- list(c("T1", "T2"), c("T2", "T3"), c("T1", "T3"))

# List to store results
wilcox_results <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Find herds common to both time periods
  common_herds <- intersect(colnames(all_temps[[p1]][[metric]]), 
                            colnames(all_temps[[p2]][[metric]]))
  
  for (herd in common_herds) {
    # Extract per-locus values
    v1 <- all_temps[[p1]][[metric]][, herd]
    v2 <- all_temps[[p2]][[metric]][, herd]
    
    # Run the paired test
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # Store results in a data frame
    wilcox_results[[paste(herd, p1, p2, sep="_")]] <- data.frame(
      Herd = herd,
      Comparison = paste(p1, "vs", p2),
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine into one master table
temporal_summary_table <- do.call(rbind, wilcox_results)
print(temporal_summary_table)

```

```{r}
# Change the metric to Ho
metric_ho <- "Ho"

# List to store Ho results
wilcox_results_ho <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  common_herds <- intersect(colnames(all_temps[[p1]][[metric_ho]]), 
                            colnames(all_temps[[p2]][[metric_ho]]))
  
  for (herd in common_herds) {
    v1 <- all_temps[[p1]][[metric_ho]][, herd]
    v2 <- all_temps[[p2]][[metric_ho]][, herd]
    
    # Paired test evaluates if v2 is consistently lower/higher than v1 across loci
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    wilcox_results_ho[[paste(herd, p1, p2, sep="_")]] <- data.frame(
      Herd = herd,
      Comparison = paste(p1, "vs", p2),
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine and apply FDR correction to account for multiple herd comparisons
temporal_summary_ho <- do.call(rbind, wilcox_results_ho)
temporal_summary_ho$adj_p_value <- p.adjust(temporal_summary_ho$p_value, method = "fdr")
temporal_summary_ho$Significant_Adj <- ifelse(temporal_summary_ho$adj_p_value < 0.05, "Yes", "No")

print(temporal_summary_ho)
```

# Allelic Richness
```{r}
temp1_merge.Ar <- allelic.richness(temp1_merge, min.n = 10, diploid = TRUE)
temp3_merge.Ar <- allelic.richness(temp3_merge, min.n = 10, diploid = TRUE)

# Define the objects (ensure you include all relevant temporal groups)
# Note: Rarefaction ensures Ar is comparable across different sample sizes (n)
all_temps_ar <- list("T1" = temp1_merge.Ar, "T2" = temp2.Ar, "T3" = temp3_merge.Ar)

# Define the metric (Ar represents Allelic Richness)
metric_ar <- "Ar"

# Pairs of time periods to compare
period_pairs <- list(c("T1", "T2"), c("T2", "T3"), c("T1", "T3"))

# List to store Ar results
wilcox_results_ar <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Find herds common to both time periods
  common_herds <- intersect(colnames(all_temps_ar[[p1]][[metric_ar]]), 
                            colnames(all_temps_ar[[p2]][[metric_ar]]))
  
  for (herd in common_herds) {
    # Extract per-locus Ar values
    v1 <- all_temps_ar[[p1]][[metric_ar]][, herd]
    v2 <- all_temps_ar[[p2]][[metric_ar]][, herd]
    
    # Run the paired test
    # This evaluates if Ar consistently decreases/increases across loci
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # Store results
    wilcox_results_ar[[paste(herd, p1, p2, sep="_")]] <- data.frame(
      Herd = herd,
      Comparison = paste(p1, "vs", p2),
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine and apply False Discovery Rate (FDR) correction
temporal_summary_ar <- do.call(rbind, wilcox_results_ar)
temporal_summary_ar$adj_p_value <- p.adjust(temporal_summary_ar$p_value, method = "fdr")
temporal_summary_ar$Significant_Adj <- ifelse(temporal_summary_ar$adj_p_value < 0.05, "Yes", "No")

print(temporal_summary_ar)
```




```{r}
# Choose the temporal group to test (e.g., temp3)
current_obj <- temp3_merge
herd_list <- levels(pop(current_obj))

# Storage for results
one_vs_all_results <- list()

for (target_herd in herd_list) {
  
  # 1. Create a copy of the object to avoid overwriting original data
  temp_genind <- current_obj
  
  # 2. Recode: Target herd stays the same, everyone else becomes "Others"
  new_pop_vector <- ifelse(pop(temp_genind) == target_herd, target_herd, "Others")
  pop(temp_genind) <- as.factor(new_pop_vector)
  
  # 3. Split the new 2-population genind
  split_gen <- seppop(temp_genind)
  
  # 4. Run the permutation test (Hs.test)
  # Comparing Target vs. Pooled Others
  test_out <- Hs.test(split_gen[[target_herd]], split_gen[["Others"]], n.sim = 999)
  
  # 5. Store data
  one_vs_all_results[[target_herd]] <- data.frame(
    Herd = target_herd,
    Observed_Diff = test_out$obs, # Negative means herd is LESS diverse than others
    P_Value = test_out$pvalue
  )
}

# Combine into a master table and apply correction
final_one_vs_all <- do.call(rbind, one_vs_all_results)
final_one_vs_all$adj_p_value <- p.adjust(final_one_vs_all$P_Value, method = "fdr")

print(final_one_vs_all)

write.csv(final_one_vs_all, "./tables/temp3_sig_hs.csv", row.names = F)
```


