---
title: "Genetic Diversity"
author: "Alex Depew"
date: "2026-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# File information
This script is used to create descriptive statistic tables for population genetic analyses of all wood bison herds in Canada. 

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026.csv",
                       check.names = F)
```

Merge WBNP Subherds
```{r}
bison.data <- bison.data %>% 
  filter(HERD != "WBNP") %>% 
  mutate(HERD.2 = if_else(grepl("^WBNP|WENT", HERD), "WBNP", HERD)) %>% 
  relocate(HERD.2, .after = HERD)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(20:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD.2,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```


#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

# Remove herds from temporal group 2 
remove_herds <- c("WBNP", "ETH", "NORD", "WENT", "WBNPPAD", "WBNPNY")
temp2 <- temp2[pop = setdiff(popNames(temp2), remove_herds), drop = TRUE]
temp2.sum <- summary(temp2) # Recalculate summary
temp2.sum
```
# Descriptive Statistics 
Functions requiring the package *strataG* require inputs to be in "gtypes" objects. Programs such as HP-RARE require data to be in GENEPOP file format. 
```{r}
# Convert genind to gtypes
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)
temp3.g <- genind2gtypes(temp3)

# Convert genind to genepop
library(graph4lg)
genind_to_genepop(temp1, output = "genepop_files/temp1.genepop.txt")
genind_to_genepop(temp2, output = "genepop_files/temp2.genepop.txt")
genind_to_genepop(temp3, output = "genepop_files/temp3.genepop.txt")

# Hierfstat summary format
temp1.stats <- basic.stats(temp1, diploid = TRUE, digits = 3)
temp2.stats <- basic.stats(temp2, diploid = TRUE, digits = 3)
temp3.stats <- basic.stats(temp3, diploid = TRUE, digits = 3)
```

## Temporal Group 1
```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity
temp1.Ho <- round(colMeans(temp1.stats$Ho), 2) # Observed (Ho)
# HO SD
temp1.Ho.sd <- round(apply(temp1.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp1.He <- round(colMeans(temp1.stats$Hs), 2) # Expected (He)
# HE SD
temp1.He.sd <- round(apply(temp1.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp1.Ar <- allelic.richness(temp1, min.n = NULL, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)
# AR SD
temp1.Ar.sd <- round(apply(temp1.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df1 <- enframe(temp1.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp1.He[Population], "±", temp1.He.sd[Population]), # Combine SD with ±
        HO = paste(temp1.Ho[Population], "±", temp1.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp1.Ar.mean[Population], "±", temp1.Ar.sd[Population]), # Combine with ±
        FIS = temp1.Fis
        )

print(descr.df1)
```
## Temporal Group 2
```{r}
# Number of samples per population
temp2.N = temp2.sum$n.by.pop
temp2.N

# Heterozygosity
temp2.Ho <- round(colMeans(temp2.stats$Ho), 2) # Observed (Ho)
# HO SD
temp2.Ho.sd <- round(apply(temp2.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp2.He <- round(colMeans(temp2.stats$Hs), 2) # Expected (He)
# HE SD
temp2.He.sd <- round(apply(temp2.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp2.Ar <- allelic.richness(temp2, min.n = NULL, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)
# AR SD
temp2.Ar.sd <- round(apply(temp2.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df2 <- enframe(temp2.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp2.He[Population], "±", temp2.He.sd[Population]), # Combine SD with ±
        HO = paste(temp2.Ho[Population], "±", temp2.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp2.Ar.mean[Population], "±", temp2.Ar.sd[Population]), # Combine with ±
        FIS = temp2.Fis
        )

print(descr.df2)
```

## Temporal Group 3
```{r}
# Number of samples per population
temp3.N = temp3.sum$n.by.pop
temp3.N

# Heterozygosity
temp3.Ho <- round(colMeans(temp3.stats$Ho), 2) # Observed (Ho)
# HO SD
temp3.Ho.sd <- round(apply(temp3.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp3.He <- round(colMeans(temp3.stats$Hs), 2) # Expected (He)
# HE SD
temp3.He.sd <- round(apply(temp3.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp3.Ar <- allelic.richness(temp3, min.n = NULL, diploid = TRUE)
temp3.Ar.mean <- round(colMeans(temp3.Ar$Ar), 2)
# AR SD
temp3.Ar.sd <- round(apply(temp3.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp3.Fis <- round(colMeans(temp3.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df3 <- enframe(temp3.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp3.He[Population], "±", temp3.He.sd[Population]), # Combine SD with ±
        HO = paste(temp3.Ho[Population], "±", temp3.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp3.Ar.mean[Population], "±", temp3.Ar.sd[Population]), # Combine with ±
        FIS = temp3.Fis
        )

print(descr.df3)
```

Join Tables (Wide)
```{r}
full.descr.wide <- descr.df1 %>% 
  full_join(descr.df2, by = "Population") %>% 
  full_join(descr.df3, by = "Population") %>% 
  
  # Relocate columns for merging in excel
  relocate(c("N.y", "N"), .after = "N.x") %>% # Number of samples
  relocate(c("HE.y", "HE"), .after = "HE.x") %>% 
  relocate(c("HO.y", "HO"), .after = "HO.x") %>%
  relocate(c("AR.y", "AR"), .after = "AR.x") %>% 
  relocate(c("FIS.y", "FIS"), .after = "FIS.x") %>% 
  arrange(Population == "EINPP", Population)

# Export to csv
write.table(full.descr.wide, "./tables/Temporal_GD_Table_wide.csv", 
            sep = ",",
            fileEncoding = "UTF-8", 
            row.names = FALSE)
```

Join Tables (Long)
```{r}
df1_v <- descr.df1 %>% mutate(Period = "T1")
df2_v <- descr.df2 %>% mutate(Period = "T2")
df3_v <- descr.df3 %>% mutate(Period = "T3")

full.descr.long <- bind_rows(df1_v, df2_v, df3_v) %>% 
  arrange(Population == "EINPP", Population) %>% 
  relocate(Period, .after = Population)

# Write to csv
write.table(full.descr.long, "./tables/Temporal_GD_Table_long.csv", 
            sep = ",",
            fileEncoding = "UTF-8", 
            row.names = FALSE)
```

# Temporal Comparison
```{r}
#WAB
MBS_T1 <- temp2.Ar$Ar[,3]
MBS_T1

MBS_T2 <- temp3.Ar$Ar[,6]
MBS_T2

temporal_test <- wilcox.test(WAB_T1, WAB_T2, paired = TRUE)

temporal_test
```


