---
title: "Descriptive Stats"
author: "Alex Depew"
date: "2026-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to create descriptive statistic tables for population genetic analyses of all wood bison herds in Canada. 

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_locus_data_20.02.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)

```
#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

```
# Descriptive Statistics 
Functions requiring the package *strataG* require inputs to be in "gtypes" objects. Programs such as HP-RARE require data to be in GENEPOP file format. 
```{r}
# Convert genind to gtypes
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)
temp3.g <- genind2gtypes(temp3)

# Convert genind to genepop
library(graph4lg)
genind_to_genepop(temp1, output = "genepop_files/temp1.genepop.txt")
genind_to_genepop(temp2, output = "genepop_files/temp2.genepop.txt")
genind_to_genepop(temp3, output = "genepop_files/temp3.genepop.txt")

# Hierfstat summary format
temp1.stats <- basic.stats(temp1, diploid = TRUE, digits = 3)
temp2.stats <- basic.stats(temp2, diploid = TRUE, digits = 3)
temp3.stats <- basic.stats(temp3, diploid = TRUE, digits = 3)
```

## Temporal Group 1
```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity
temp1.Ho <- round(colMeans(temp1.stats$Ho), 3) # Observed (Ho)
temp1.He <- round(colMeans(temp1.stats$Hs), 3) # Expected (He)

# Mean Number of Alleles
temp1.pf <- temp1.stats$pop.freq

# Count alleles per population per locus, then mean across loci
temp1.mean_Na <- sapply(temp1.pf, function(temp1.mat) {
  if(is.null(dim(temp1.mat))) temp1.mat <- matrix(temp1.mat, ncol = length(temp1.mat))  # handle monomorphic loci
  colSums(temp1.mat > 0)
})

# Take the mean across loci and round
temp1.Na <- round(rowMeans(temp1.mean_Na), 2)
temp1.Na

# Allelic Richness
temp1.Ar <- allelic.richness(temp1, min.n = NULL, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)

# Effective number of alleles
temp1.AE <- round(1/(1 - temp1.He), 2)

# Private Alleles
temp1.NP <- privateAlleles(temp1.g)
temp1.NP.sums <- data.frame(NP = colSums(temp1.NP)) %>% 
  rownames_to_column(var = "Population")

# Private Allelic Richness
temp1.PR <- read.table("hprare_output/temp1.hprare.txt", 
                       header = FALSE,
                       skip = 3)
colnames(temp1.PR) <- c("Population", "RP")
temp1.PR.short <- temp1.PR %>% 
  select(1:2) %>% 
  mutate(Population = str_remove(Population, "_.*")) # Note: rarefied to minimum genotypes = 18

# Join NP and PR by Herd
temp1.priv <- left_join(temp1.NP.sums, temp1.PR.short, by = "Population")

# Fis
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 3)
temp1.Fis

# Table
temp1.descr <- data.frame(
  Population = colnames(temp1.stats$Hs),
  n = temp1.N,
  HEXP = temp1.He,
  HOBS = temp1.Ho,
  Na = temp1.Na,
  AR = temp1.Ar.mean,
  AE = temp1.AE,
  FIS = temp1.Fis
) %>% 
left_join(temp1.priv)

temp1.descr

# Write to csv
write.csv(temp1.descr,"tables/Descriptive Stats Temp1.csv")

colMeans(temp1.descr[2:9])
```

## Temporal Group 2
```{r}
# Number of samples per population
temp2.N = temp2.sum$n.by.pop
temp2.N

# Heterozygosity
temp2.Ho <- round(colMeans(temp2.stats$Ho), 3) # Observed (Ho)
temp2.He <- round(colMeans(temp2.stats$Hs), 3) # Expected (He)

# Mean Number of Alleles
temp2.pf <- temp2.stats$pop.freq

# Count alleles per population per locus, then mean across loci
temp2.mean_Na <- sapply(temp2.pf, function(temp2.mat) {
  if(is.null(dim(temp2.mat))) temp2.mat <- matrix(temp2.mat, ncol = length(temp2.mat))  # handle monomorphic loci
  colSums(temp2.mat > 0)
})

# Take the mean across loci and round
temp2.Na <- round(rowMeans(temp2.mean_Na), 2)
temp2.Na

# Allelic Richness
temp2.Ar <- allelic.richness(temp2, min.n = NULL, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)

# Effective number of alleles
temp2.AE <- round(1/(1 - temp2.He), 2)

# Private Alleles
temp2.NP <- privateAlleles(temp2.g)
temp2.NP.sums <- data.frame(NP = colSums(temp2.NP)) %>% 
  rownames_to_column(var = "Population")

# Private Allelic Richness
temp2.PR <- read.table("hprare_output/temp2.hprare.txt", 
                       header = FALSE,
                       skip = 3)
colnames(temp2.PR) <- c("Population", "RP")
temp2.PR.short <- temp2.PR %>% 
  select(1:2) %>% 
  mutate(Population = str_remove(Population, "_.*")) # Note: rarefied to minimum genotypes = 20

# Join NP and PR by Herd
temp2.priv <- left_join(temp2.NP.sums, temp2.PR.short, by = "Population")

# Fis
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 3)
temp2.Fis

# Table
temp2.descr <- data.frame(
  Population = colnames(temp2.stats$Hs),
  n = temp2.N,
  HEXP = temp2.He,
  HOBS = temp2.Ho,
  Na = temp2.Na,
  AR = temp2.Ar.mean,
  AE = temp2.AE,
  FIS = temp2.Fis
) %>% 
left_join(temp2.priv)

temp2.descr

# Write to csv
write.csv(temp2.descr,"tables/Descriptive Stats Temp2.csv")
```

## Temporal Group 3
```{r}
# Number of samples per population
temp3.N = temp3.sum$n.by.pop
temp3.N

# Heterozygosity
temp3.Ho <- round(colMeans(temp3.stats$Ho), 3) # Observed (Ho)
temp3.He <- round(colMeans(temp3.stats$Hs), 3) # Expected (He)

# Mean Number of Alleles
temp3.pf <- temp3.stats$pop.freq

# Count alleles per population per locus, then mean across loci
temp3.mean_Na <- sapply(temp3.pf, function(temp3.mat) {
  if(is.null(dim(temp3.mat))) temp3.mat <- matrix(temp3.mat, ncol = length(temp3.mat))  # handle monomorphic loci
  colSums(temp3.mat > 0)
})

# Take the mean across loci and round
temp3.Na <- round(rowMeans(temp3.mean_Na), 2)
temp3.Na

# Allelic Richness
temp3.Ar <- allelic.richness(temp3, min.n = NULL, diploid = TRUE)
temp3.Ar.mean <- round(colMeans(temp3.Ar$Ar), 2)

# Effective number of alleles
temp3.AE <- round(1/(1 - temp3.He), 2)

# Private Alleles
temp3.NP <- privateAlleles(temp3.g)
temp3.NP.sums <- data.frame(NP = colSums(temp3.NP)) %>% 
  rownames_to_column(var = "Population")

# Private Allelic Richness
temp3.PR <- read.table("hprare_output/temp3.hprare.txt", 
                       header = FALSE,
                       skip = 3)
colnames(temp3.PR) <- c("Population", "RP")
temp3.PR.short <- temp3.PR %>% 
  select(1:2) %>% 
  mutate(Population = str_remove(Population, "_.*")) # Note: rarefied to minimum genotypes = 6

# Join NP and PR by Herd
temp3.priv <- left_join(temp3.NP.sums, temp3.PR.short, by = "Population")

# Fis
temp3.Fis <- round(colMeans(temp3.stats$Fis, na.rm = TRUE), 3)
temp3.Fis

# Table
temp3.descr <- data.frame(
  Population = colnames(temp3.stats$Hs),
  n = temp3.N,
  HEXP = temp3.He,
  HOBS = temp3.Ho,
  Na = temp3.Na,
  AR = temp3.Ar.mean,
  AE = temp3.AE,
  FIS = temp3.Fis
) %>% 
left_join(temp3.priv)

temp3.descr

# Write to csv
write.csv(temp3.descr,"tables/Descriptive Stats Temp3.csv")
```


# Visualize Expected Heterozygosity
```{r}
# Hs dataframes
temp1.He.df <- Hs(temp1) %>% as.data.frame()
colnames(temp1.He.df) <- c("He")
temp1.He.df <- rownames_to_column(temp1.He.df, var = "Herd")

temp2.He.df <- Hs(temp2) %>% as.data.frame()
colnames(temp2.He.df) <- c("He")
temp2.He.df <- rownames_to_column(temp2.He.df, var = "Herd")


temp3.He.df <- Hs(temp3) %>% as.data.frame()
colnames(temp3.He.df) <- c("He")
temp3.He.df <- rownames_to_column(temp3.He.df, var = "Herd")


# Plotting
ggplot(temp1.He.df, aes(x = Herd, y = He)) + 
  geom_bar(stat = "identity", fill = "#93E371") + 
  theme_minimal() + 
  labs(title ="Expected Heterozygosity in Temporal Group 1") +
  theme(axis.text.x = element_text(angle = 45))

ggplot(temp2.He.df, aes(x = Herd, y = He)) + 
  geom_bar(stat = "identity", fill = "#E5A1CC") + 
  theme_minimal() + 
  labs(title ="Expected Heterozygosity in Temporal Group 2") + 
  theme(axis.text.x = element_text(angle = 45))

ggplot(temp3.He.df, aes(x = Herd, y = He)) + 
  geom_bar(stat = "identity", fill = "#393b79") + 
  theme_minimal() + 
  labs(title ="Expected Heterozygosity in Temporal Group 3") + 
  theme(axis.text.x = element_text(angle = 45))
```
# Map plots on top of each other
```{r}
# Prepare each dataframe with a Group column
temp1.He.df <- rownames_to_column(Hs(temp1) %>% as.data.frame(), var = "Herd")
colnames(temp1.He.df)[2] <- "He"
temp1.He.df$Group <- "Temporal 1"

temp2.He.df <- rownames_to_column(Hs(temp2) %>% as.data.frame(), var = "Herd")
colnames(temp2.He.df)[2] <- "He"
temp2.He.df$Group <- "Temporal 2"

temp3.He.df <- rownames_to_column(Hs(temp3) %>% as.data.frame(), var = "Herd")
colnames(temp3.He.df)[2] <- "He"
temp3.He.df$Group <- "Temporal 3"

# Combine all dataframes
all.He.df <- bind_rows(temp1.He.df, temp2.He.df, temp3.He.df)

# Make Herd a factor so ggplot keeps all herds in order
all.He.df$Herd <- factor(all.He.df$Herd, levels = unique(all.He.df$Herd))

# Plot
ggplot(all.He.df, aes(x = Herd, y = He, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Expected Heterozygosity Across Temporal Groups") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Temporal 1" = "#93E371",
                               "Temporal 2" = "#E5A1CC",
                               "Temporal 3" = "#393b79"))
```
# Pairwise Fst 
```{r}
# Subset data to reduce computation time
temp1.sub <- popsub(temp1, sublist = c("WBNPGR",
                                        "WBNPPAD",
                                        "WBNPNY",
                                        "WBNPLB",
                                        "WBNPHC",
                                        "MBS",
                                        "EINPP",
                                        "EINPW"))

temp2.sub <- popsub(temp2, sublist = c("HZ",
                                       "RL",
                                       "WAB",
                                       "EINPP",
                                       "EINPW",
                                       "MBS"))
                    
temp3.sub <- popsub(temp3, sublist = c("CL",
                                       "ETH",
                                       "HZ",
                                       "NAH",
                                       "RL",
                                       "WAB",
                                       "WBNPGR",
                                       "WBNPPAD",
                                       "WBNPNY",
                                       "WBNPLB",
                                       "WBNPHC",
                                       "EINPP",
                                       "EINPW",
                                       "NORD",
                                       "WENT",
                                       "AISH"))

# Calculate pairwise Fst
temp1.fst <- genet.dist(temp1.sub, method = "WC84") %>% 
  round(digits = 3)

temp2.fst <- genet.dist(temp2.sub, method = "WC84") %>% 
  round(digits = 3)

temp3.fst <- genet.dist(temp3.sub, method = "WC84") %>% 
  round(digits = 3)

# Visualize with heatmaps ----

# Label order
temp1.order <- c("WBNPGR",
                 "WBNPPAD",
                 "WBNPNY",
                 "WBNPLB",
                 "WBNPHC",
                 "EINPP",
                 "EINPW",
                 "MBS")

temp2.order <- c("HZ",
                 "RL",
                 "WAB",
                 "EINPP",
                 "EINPW",
                 "MBS")

temp3.order <- c("CL",
                 "ETH",
                 "HZ",
                 "NAH",
                 "RL",
                 "WAB",
                 "WBNPGR",
                 "WBNPPAD",
                 "WBNPNY",
                 "WBNPLB",
                 "WBNPHC",
                 "EINPP",
                 "EINPW",
                 "NORD",
                 "WENT",
                 "AISH")

# Change order of rows and cols in matrices
library(reshape2)
temp1.fst.mat <- as.matrix(temp1.fst)[temp1.order, temp1.order]
temp1.fst.df <- melt(temp1.fst.mat, varnames = c("Site1", "Site2"), value.name = "Fst")
temp1.fst.df <- temp1.fst.df[as.numeric(factor(temp1.fst.df$Site1, levels=temp1.order)) > 
                             as.numeric(factor(temp1.fst.df$Site2, levels=temp1.order)), ]

temp1.fst.df$Site1 <- factor(temp1.fst.df$Site1, levels = temp1.order)
temp1.fst.df$Site2 <- factor(temp1.fst.df$Site2, levels = temp1.order)

temp1.fst.df$Fst[temp1.fst.df$Fst < 0] <- 0

# Group 2
temp2.fst.mat <- as.matrix(temp2.fst)[temp2.order, temp2.order]
temp2.fst.df <- melt(temp2.fst.mat, varnames = c("Site1", "Site2"), value.name = "Fst")
temp2.fst.df <- temp2.fst.df[as.numeric(factor(temp2.fst.df$Site1, levels=temp2.order)) > 
                             as.numeric(factor(temp2.fst.df$Site2, levels=temp2.order)), ]

temp2.fst.df$Site1 <- factor(temp2.fst.df$Site1, levels = temp2.order)
temp2.fst.df$Site2 <- factor(temp2.fst.df$Site2, levels = temp2.order)

temp2.fst.df$Fst[temp2.fst.df$Fst < 0] <- 0

# Group3 
temp3.fst.mat <- as.matrix(temp3.fst)[temp3.order, temp3.order]
temp3.fst.df <- melt(temp3.fst.mat, varnames = c("Site1", "Site2"), value.name = "Fst")
temp3.fst.df <- temp3.fst.df[as.numeric(factor(temp3.fst.df$Site1, levels=temp3.order)) > 
                             as.numeric(factor(temp3.fst.df$Site2, levels=temp3.order)), ]

temp3.fst.df$Site1 <- factor(temp3.fst.df$Site1, levels = temp3.order)
temp3.fst.df$Site2 <- factor(temp3.fst.df$Site2, levels = temp3.order)

temp3.fst.df$Fst[temp3.fst.df$Fst < 0] <- 0



# Italic label
fst.label <-  expression(italic("F")[ST])

# Set middle Fst value to set gradient argument
temp1.mid = max(temp1.fst.df$Fst) / 2

temp2.mid = max(temp2.fst.df$Fst) / 2

temp3.mid = max(temp3.fst.df$Fst) / 2

# Plot ----
temp1.fst.plot <- ggplot(data = temp1.fst.df, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = Fst), color="black", size = 3)+
                  scale_fill_gradient2(low = "blue", mid = "pink", high = "red", midpoint = temp1.mid, name = fst.label, 
                  limits = c(0, max(temp1.fst.df$Fst)), breaks = c(0, 0.05, 0.10, 0.15))+
                  scale_x_discrete(expand = c(0,0))+
                  scale_y_discrete(expand = c(0,0), position = "right")+
                  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
                        axis.title = element_blank(),
                        panel.grid = element_blank(),
                        panel.background = element_blank(),
                        legend.position = "right",
                        legend.title = element_text(size = 14, face = "bold"),
                        legend.text = element_text(size = 10)
                        )
temp1.fst.plot
ggsave("plots/pre1999fst.pdf",
       width = 8,
       height = 4,
       units = "in")


# Temporal group 2
temp2.fst.plot <- ggplot(data = temp2.fst.df, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = Fst), color="black", size = 3)+
                  scale_fill_gradient2(low = "blue", mid = "pink", high = "red", midpoint = temp2.mid, name = fst.label, 
                  limits = c(0, max(temp2.fst.df$Fst)), breaks = c(0, 0.05, 0.10, 0.15))+
                  scale_x_discrete(expand = c(0,0))+
                  scale_y_discrete(expand = c(0,0), position = "right")+
                  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
                        axis.title = element_blank(),
                        panel.grid = element_blank(),
                        panel.background = element_blank(),
                        legend.position = "right",
                        legend.title = element_text(size = 14, face = "bold"),
                        legend.text = element_text(size = 10)
                        )

temp2.fst.plot
ggsave("plots/pre2016fst.pdf",
       width = 8,
       height = 4,
       units = "in")

# Temporal Group 3
temp3.fst.plot <- ggplot(data = temp3.fst.df, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = Fst), color="black", size = 3)+
                  scale_fill_gradient2(low = "blue", mid = "pink", high = "red", midpoint = temp3.mid, name = fst.label, 
                  limits = c(0, max(temp3.fst.df$Fst)), breaks = c(0, 0.05, 0.10, 0.15))+
                  scale_x_discrete(expand = c(0,0))+
                  scale_y_discrete(expand = c(0,0), position = "right")+
                  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
                        axis.text.x = element_text(angle = 45, hjust = +1),
                        axis.title = element_blank(),
                        panel.grid = element_blank(),
                        panel.background = element_blank(),
                        legend.position = "right",
                        legend.title = element_text(size = 14, face = "bold"),
                        legend.text = element_text(size = 10)
                        )
temp3.fst.plot

ggsave("plots/post2016fst.pdf",
       width = 8,
       height = 4,
       units = "in")


```

