---
title: "GD_Revised"
author: "Alex Depew"
date: "2026-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```


## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_07.03.2026.csv",
                         check.names = F)


# Merge WBNP Subherds
mod.data <- cleaned.data %>% 
  mutate(HERD = case_when(
    grepl("WBNPLB|WBNPNY", HERD) ~ "WBNPC1",
    grepl("WBNPPAD|WBNPGR|WENT", HERD) ~ "WBNPC2",
    TRUE                         ~ HERD  
  )) %>% 
  relocate(HERD, .after = YEAR)

# Merge new WBNP clusters back onto the dataset
wbnp.data <- mod.data %>% 
  filter(HERD %in% c("WBNPC1", "WBNPC2"))

merged.data <- rbind(cleaned.data, wbnp.data)

```

#### Convert data to Genind object
```{r}
# Make genind object
merge.genind <- df2genind(X=merged.data[,c(19:ncol(merged.data))],
                              sep = "/",
                              ind.names = merged.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = merged.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

merge.genind@other$YEAR.ASSIGN <- merged.data$YEAR.ASSIGN
```


#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 2: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- merge.genind[merge.genind@other$YEAR.ASSIGN <=2015,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- merge.genind[merge.genind@other$YEAR.ASSIGN >=2016,]
temp2.sum <- summary(temp2)
```

# Descriptive Statistics 
Functions requiring the package *strataG* require inputs to be in "gtypes" objects. Programs such as HP-RARE require data to be in GENEPOP file format. 
```{r}
# Convert genind to gtypes
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)

# Convert genind to genepop
library(graph4lg)
genind_to_genepop(temp1, output = "genepop_files/temp1_revised.genepop.txt")
genind_to_genepop(temp2, output = "genepop_files/temp2_revised.genepop.txt")

# Hierfstat summary format
temp1.stats <- basic.stats(temp1, diploid = TRUE, digits = 3)
temp2.stats <- basic.stats(temp2, diploid = TRUE, digits = 3)
```

## Temporal Group 1
```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity
temp1.Ho <- round(colMeans(temp1.stats$Ho), 2) # Observed (Ho)
# HO SD
temp1.Ho.sd <- round(apply(temp1.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp1.He <- round(colMeans(temp1.stats$Hs), 2) # Expected (He)
# HE SD
temp1.He.sd <- round(apply(temp1.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp1.Ar <- allelic.richness(temp1, min.n = 10, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)
# AR SD
temp1.Ar.sd <- round(apply(temp1.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df1 <- enframe(temp1.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp1.He[Population], "±", temp1.He.sd[Population]), # Combine SD with ±
        HO = paste(temp1.Ho[Population], "±", temp1.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp1.Ar.mean[Population], "±", temp1.Ar.sd[Population]), # Combine with ±
        FIS = temp1.Fis
        )

print(descr.df1)
```

## Temporal Group 2
```{r}
# Number of samples per population
temp2.N = temp2.sum$n.by.pop
temp2.N

# Heterozygosity
temp2.Ho <- round(colMeans(temp2.stats$Ho), 2) # Observed (Ho)
# HO SD
temp2.Ho.sd <- round(apply(temp2.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp2.He <- round(colMeans(temp2.stats$Hs), 2) # Expected (He)
# HE SD
temp2.He.sd <- round(apply(temp2.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp2.Ar <- allelic.richness(temp2, min.n = 10, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)
# AR SD
temp2.Ar.sd <- round(apply(temp2.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df2 <- enframe(temp2.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp2.He[Population], "±", temp2.He.sd[Population]), # Combine SD with ±
        HO = paste(temp2.Ho[Population], "±", temp2.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp2.Ar.mean[Population], "±", temp2.Ar.sd[Population]), # Combine with ±
        FIS = temp2.Fis
        )

print(descr.df2)
```


Join Tables (Long)
```{r}
df1_v <- descr.df1 %>% mutate(Period = "T1")
df2_v <- descr.df2 %>% mutate(Period = "T2")

full.descr.long <- bind_rows(df1_v, df2_v) %>% 
  arrange(Population == "EINPP", Population) %>% 
  relocate(Period, .after = Population)

# Write to csv
write.table(full.descr.long, "./tables/Temporal_GD_Table.csv", 
            sep = ",",
            fileEncoding = "UTF-8", 
            row.names = FALSE)
```

# Genetic Distance 

#### Weir and Cockerham (1984) Fst
```{r}
# Temporal group 1
## Order herds 
temp1_fst.mat <- as.matrix(pairwise.WCfst(temp1,
                                        diploid = TRUE))

fst1 <- data.frame(round(temp1_fst.mat, 3))
write.csv(fst1, "./tables/fst1.csv", row.names = T)

# Temporal group 2
temp2_fst.mat <- as.matrix(pairwise.WCfst(temp2,
                                          diploid = TRUE))

fst2 <- data.frame(round(temp2_fst.mat, 3))
write.csv(fst2, "./tables/fst2.csv", row.names = T)
```

#### Reorder FST Tables
```{r}
# Reorder
current_names <- colnames(temp1_fst.mat)

# 2. Define the new order manually (swapping the last two)
new_order <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "WBNPC1", "WBNPC2",
                "EINPW", "HZ", "MBS", "RL", "WAB", "EINPP") 

# 3. Apply the swap to both rows and columns
temp1_fst.ordered <- temp1_fst.mat[new_order, new_order]

# View the result
round(temp1_fst.ordered, 3)
write.csv(temp1_fst.ordered, "./tables/fst1.csv", row.names = T)

# Reorder
current_names2 <- colnames(temp2_fst.mat)
current_names2

# 2. Define the new order manually (swapping the last two)
new_order2 <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "WBNPC1", "WBNPC2", "AISH","CL", "EINPW", "ETH", "HZ", "NAH", "NORD", "RL", "WAB", "EINPP")

# 3. Apply the swap to both rows and columns
temp2_fst.ordered <- temp2_fst.mat[new_order2, new_order2]

# View the result
round(temp2_fst.ordered, 3)
write.csv(temp2_fst.ordered, "./tables/fst2.csv", row.names = T)
```


#### Bootstraps of Fst - Test for significance (Null hypothesis Fst = 0)
```{r}
boot1.mat <- boot.ppfst(temp1, nboot = 1000, quant=c(0.025,0.975))
boot1.mat$ll
boot2.mat <- boot.ppfst(temp2, nboot = 1000, quant=c(0.025,0.975))
boot2.mat$ll
```

## Fis Sig
```{r}
temp1_fis_boot <- boot.ppfis(dat=temp1,nboot=1000,quant=c(0.025,0.975))
temp1_fis_boot$fis.ci$ll # EINPW is the only herd with a significant FIS value

temp1.stat <- basic.stats(temp1)
colMeans(temp1.stats$Fis)

temp2_fis_boot <- boot.ppfis(dat=temp2,nboot=1000,quant=c(0.025,0.975))
temp2_fis_boot$fis.ci$ll # WBNPPAD, WBNPC2

temp2.stat <- basic.stats(temp2)
colMeans(temp2.stat$Fis)
```



### He
```{r}
# Both T1 and T2 hierfstat objects into list
all_temps <- list("T1" = temp1.stats, "T2" = temp2.stats)

# Metric to compare (Hs/He)
metric <- "Hs"

# Comparison time periods
period_pairs <- list(c("T1", "T2"))

# List to store results of comparison tests
wilcox_results <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Both time periods have different herds - this identifies which are alike
  common_herds <- intersect(colnames(all_temps[[p1]][[metric]]), 
                            colnames(all_temps[[p2]][[metric]]))
  
  for (herd in common_herds) {
    # Extract per-locus values
    v1 <- all_temps[[p1]][[metric]][, herd]
    v2 <- all_temps[[p2]][[metric]][, herd]
    
    # Run test
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # df output
    wilcox_results[[paste(herd)]] <- data.frame(
      Herd = herd,
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine results
temporal_summary_table <- do.call(rbind, wilcox_results)
temporal_summary_table
```


### Ho
```{r}
# Both T1 and T2 hierfstat objects into list
all_temps <- list("T1" = temp1.stats, "T2" = temp2.stats)

# Metric to compare (Hs/He)
metric <- "Ho"

# Comparison time periods
period_pairs <- list(c("T1", "T2"))

# List to store results of comparison tests
wilcox_results <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Both time periods have different herds - this identifies which are alike
  common_herds <- intersect(colnames(all_temps[[p1]][[metric]]), 
                            colnames(all_temps[[p2]][[metric]]))
  
  for (herd in common_herds) {
    # Extract per-locus values
    v1 <- all_temps[[p1]][[metric]][, herd]
    v2 <- all_temps[[p2]][[metric]][, herd]
    
    # Run test
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # df output
    wilcox_results[[paste(herd)]] <- data.frame(
      Herd = herd,
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine results
temporal_summary_table <- do.call(rbind, wilcox_results)
temporal_summary_table
```

# Allelic Richness
```{r}
temp1.Ar <- allelic.richness(temp1, min.n = 10, diploid = TRUE)
temp2.Ar <- allelic.richness(temp2, min.n = 10, diploid = TRUE)

# Put each temporal cohort into list form
all_temps_ar <- list("T1" = temp1.Ar, "T2" = temp2.Ar)

# Metric (Ar)
metric_ar <- "Ar"

# Cohort comparisons
period_pairs <- list(c("T1", "T2"))

# Results list
wilcox_results_ar <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Herds present in both periods
  common_herds <- intersect(colnames(all_temps_ar[[p1]][[metric_ar]]), 
                            colnames(all_temps_ar[[p2]][[metric_ar]]))
  
  for (herd in common_herds) {
    # Per-locus Ar
    v1 <- all_temps_ar[[p1]][[metric_ar]][, herd]
    v2 <- all_temps_ar[[p2]][[metric_ar]][, herd]
    
    # Run test
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # df output
    wilcox_results_ar[[paste(herd)]] <- data.frame(
      Herd = herd,
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

temporal_summary_ar <- do.call(rbind, wilcox_results_ar)
temporal_summary_ar
```


```{r}
# Temp1 He comparisons
current_obj <- temp1
herd_list <- levels(pop(current_obj))

# List results
one_vs_all_results <- list()

for (target_herd in herd_list) {
  
  temp_genind <- current_obj
  
  # Remove the target herd to compare against others
  new_pop_vector <- ifelse(pop(temp_genind) == target_herd, target_herd, "Others")
  pop(temp_genind) <- as.factor(new_pop_vector)
  
  # Split the genind
  split_gen <- seppop(temp_genind)
  
  # Run Hs.test
  test_out <- Hs.test(split_gen[[target_herd]], split_gen[["Others"]], n.sim = 999)
  
  # Results
  one_vs_all_results[[target_herd]] <- data.frame(
    Herd = target_herd,
    Observed_Diff = test_out$obs, # Negative means herd is LESS diverse than others
    P_Value = test_out$pvalue
  )
}

# make df and apply BH correction (permutation test requires)
final_one_vs_all <- do.call(rbind, one_vs_all_results)
final_one_vs_all$adj_p_value <- p.adjust(final_one_vs_all$P_Value, method = "fdr")

print(final_one_vs_all)

write.csv(final_one_vs_all, "./tables/temp1_sig_hs_v2.csv", row.names = F)
```
```{r}
# Temp2 He comparisons
current_obj <- temp2
herd_list <- levels(pop(current_obj))

# List results
one_vs_all_results <- list()

for (target_herd in herd_list) {
  
    temp_genind <- current_obj
  
  # Remove the target herd to compare against others
  new_pop_vector <- ifelse(pop(temp_genind) == target_herd, target_herd, "Others")
  pop(temp_genind) <- as.factor(new_pop_vector)
  
  # Split the genind
  split_gen <- seppop(temp_genind)
  
  # Run Hs.test
  test_out <- Hs.test(split_gen[[target_herd]], split_gen[["Others"]], n.sim = 999)
  
  # Results
  one_vs_all_results[[target_herd]] <- data.frame(
    Herd = target_herd,
    Observed_Diff = test_out$obs, # Negative means herd is LESS diverse than others
    P_Value = test_out$pvalue
  )
}

# make df and apply BH correction (permutation test requires)
final_one_vs_all <- do.call(rbind, one_vs_all_results)
final_one_vs_all$adj_p_value <- p.adjust(final_one_vs_all$P_Value, method = "fdr")

print(final_one_vs_all)

write.csv(final_one_vs_all, "./tables/temp2_sig_hs_v2.csv", row.names = F)
```

```{r}
btlneck <- mRatio(temp1.g, by.strata = T)

btlneck.sum <- btlneck %>% 
  group_by(stratum) %>% 
  summarize(
    avg.m = mean(m.ratio, na.rm = TRUE),
    sd.m = sd(m.ratio, na.rm = TRUE),
    n_loci = sum(!is.na(m.ratio))
  )

```

