---
title: "GD_Revised"
author: "Alex Depew"
date: "2026-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```


## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_07.03.2026.csv",
                         check.names = F)


# Merge WBNP Subherds
mod.data <- cleaned.data %>% 
  mutate(HERD = case_when(
    grepl("WBNPLB|WBNPNY", HERD) ~ "WBNPC1",
    grepl("WBNPPAD|WBNPGR|WENT", HERD) ~ "WBNPC2",
    TRUE                         ~ HERD  
  )) %>% 
  relocate(HERD, .after = YEAR)

# Merge new WBNP clusters back onto the dataset
wbnp.data <- mod.data %>% 
  filter(HERD %in% c("WBNPC1", "WBNPC2"))

merged.data <- rbind(cleaned.data, wbnp.data)

```

#### Convert data to Genind object
```{r}
# Make genind object
merge.genind <- df2genind(X=merged.data[,c(19:ncol(merged.data))],
                              sep = "/",
                              ind.names = merged.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = merged.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

merge.genind@other$YEAR.ASSIGN <- merged.data$YEAR.ASSIGN
```


#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 2: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- merge.genind[merge.genind@other$YEAR.ASSIGN <=2015,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- merge.genind[merge.genind@other$YEAR.ASSIGN >=2016,]
temp2.sum <- summary(temp2)
```

# Descriptive Statistics 
Functions requiring the package *strataG* require inputs to be in "gtypes" objects. Programs such as HP-RARE require data to be in GENEPOP file format. 
```{r}
# Convert genind to gtypes
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)

# Convert genind to genepop
library(graph4lg)
genind_to_genepop(temp1, output = "genepop_files/temp1_revised.genepop.txt")
genind_to_genepop(temp2, output = "genepop_files/temp2_revised.genepop.txt")

# Hierfstat summary format
temp1.stats <- basic.stats(temp1, diploid = TRUE, digits = 3)
temp2.stats <- basic.stats(temp2, diploid = TRUE, digits = 3)
```

## Temporal Group 1
```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity
temp1.Ho <- round(colMeans(temp1.stats$Ho), 2) # Observed (Ho)
# HO SD
temp1.Ho.sd <- round(apply(temp1.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp1.He <- round(colMeans(temp1.stats$Hs), 2) # Expected (He)
# HE SD
temp1.He.sd <- round(apply(temp1.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp1.Ar <- allelic.richness(temp1, min.n = 10, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)
# AR SD
temp1.Ar.sd <- round(apply(temp1.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df1 <- enframe(temp1.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp1.He[Population], "±", temp1.He.sd[Population]), # Combine SD with ±
        HO = paste(temp1.Ho[Population], "±", temp1.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp1.Ar.mean[Population], "±", temp1.Ar.sd[Population]), # Combine with ±
        FIS = temp1.Fis
        )

print(descr.df1)
```

## Temporal Group 2
```{r}
# Number of samples per population
temp2.N = temp2.sum$n.by.pop
temp2.N

# Heterozygosity
temp2.Ho <- round(colMeans(temp2.stats$Ho), 2) # Observed (Ho)
# HO SD
temp2.Ho.sd <- round(apply(temp2.stats$Ho, 2, sd, na.rm = TRUE), 2) # SD Ho

temp2.He <- round(colMeans(temp2.stats$Hs), 2) # Expected (He)
# HE SD
temp2.He.sd <- round(apply(temp2.stats$Hs, 2, sd, na.rm = TRUE), 2) # SD Ho


# Allelic Richness
temp2.Ar <- allelic.richness(temp2, min.n = 10, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)
# AR SD
temp2.Ar.sd <- round(apply(temp2.Ar$Ar, 2, sd, na.rm = TRUE), 2)

# FIS
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 2)

# Create Table
descr.df2 <- enframe(temp2.N, name = "Population", value = "N") %>%
  mutate(
        HE = paste(temp2.He[Population], "±", temp2.He.sd[Population]), # Combine SD with ±
        HO = paste(temp2.Ho[Population], "±", temp2.Ho.sd[Population]), # Combine SD with ±
        AR = paste(temp2.Ar.mean[Population], "±", temp2.Ar.sd[Population]), # Combine with ±
        FIS = temp2.Fis
        )

print(descr.df2)
```


Join Tables (Long)
```{r}
df1_v <- descr.df1 %>% mutate(Period = "T1")
df2_v <- descr.df2 %>% mutate(Period = "T2")

full.descr.long <- bind_rows(df1_v, df2_v) %>% 
  arrange(Population == "EINPP", Population) %>% 
  relocate(Period, .after = Population)

# Write to csv
write.table(full.descr.long, "./tables/Temporal_GD_Table.csv", 
            sep = ",",
            fileEncoding = "UTF-8", 
            row.names = FALSE)
```

# Genetic Distance 

#### Weir and Cockerham (1984) Fst
```{r}
# Temporal group 1
## Order herds 
temp1_fst.mat <- as.matrix(pairwise.WCfst(temp1,
                                        diploid = TRUE))

fst1 <- data.frame(round(temp1_fst.mat, 3))
write.csv(fst1, "./tables/fst1.csv", row.names = T)

# Temporal group 2
temp2_fst.mat <- as.matrix(pairwise.WCfst(temp2,
                                          diploid = TRUE))

fst2 <- data.frame(round(temp2_fst.mat, 3))
write.csv(fst2, "./tables/fst2.csv", row.names = T)
```

## Bootstraps of Fst
```{r}
boot1.mat <- boot.ppfst(temp1, nboot = 1000, quant=c(0.025,0.975))
boot1.mat$ll

boot2.mat <- boot.ppfst(temp2, nboot = 1000, quant=c(0.025,0.975))
boot2.mat$ll
```

## Fis Sig
```{r}
temp1_fis_boot <- boot.ppfis(dat=temp1,nboot=1000,quant=c(0.025,0.975))
temp1_fis_boot$fis.ci

temp1_merge.stat <- basic.stats(temp1)
colMeans(temp1.stat$Fis)

temp2_fis_boot <- boot.ppfis(dat=temp2,nboot=1000,quant=c(0.025,0.975))
temp2_fis_boot$fis.ci

temp2_merge.stat <- basic.stats(temp2)
colMeans(temp2.stat$Fis)
```




```{r}
# HE
# Put your objects into a named list
all_temps <- list("T1" = temp1.stats, "T2" = temp2.stats)

# Define the metric to compare (Hs is Expected Heterozygosity)
metric <- "Hs"

# Define the possible pairs of time periods
period_pairs <- list(c("T1", "T2"))

# List to store results
wilcox_results <- list()

for (pair in period_pairs) {
  p1 <- pair[1]
  p2 <- pair[2]
  
  # Find herds common to both time periods
  common_herds <- intersect(colnames(all_temps[[p1]][[metric]]), 
                            colnames(all_temps[[p2]][[metric]]))
  
  for (herd in common_herds) {
    # Extract per-locus values
    v1 <- all_temps[[p1]][[metric]][, herd]
    v2 <- all_temps[[p2]][[metric]][, herd]
    
    # Run the paired test
    test <- wilcox.test(v1, v2, paired = TRUE)
    
    # Store results in a data frame
    wilcox_results[[paste(herd, p1, p2, sep="_")]] <- data.frame(
      Herd = herd,
      Comparison = paste(p1, "vs", p2),
      V_stat = test$statistic,
      p_value = round(test$p.value, 5),
      Significant = ifelse(test$p.value < 0.05, "Yes", "No")
    )
  }
}

# Combine into one master table
temporal_summary_table <- do.call(rbind, wilcox_results)
print(temporal_summary_table)

```