---
title: "FST"
author: "Alex Depew"
date: "2026-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to assess population differentiation via pairwise FST.

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering

## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(dplyr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026.csv",
                       check.names = F)
```

Merge WBNP Subherds
```{r}
bison.data <- bison.data %>% 
  filter(HERD != "WBNP") %>% 
  mutate(HERD.2 = if_else(grepl("^WBNP|WENT", HERD), "WBNP", HERD)) %>% 
  relocate(HERD.2, .after = HERD)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(20:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD.2,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

# Remove herds from temporal group 2 
remove_herds <- c("WBNP", "ETH", "NORD", "WENT")
temp2 <- temp2[pop = setdiff(popNames(temp2), remove_herds), drop = TRUE]
temp2.sum <- summary(temp2) # Recalculate summary
temp2.sum
```
Weir and Cockerham (1984 Fst)
```{r}
# Temporal group 1
temp1_fst.mat <- as.matrix(pairwise.WCfst(temp1,
                            diploid = TRUE))

round(temp1_fst.mat, 3)

# Temporal group 2
temp2_fst.mat <- as.matrix(pairwise.WCfst(temp2,
                                          diploid = TRUE))

round(temp2_fst.mat, 3)

# Temporal group 3
temp3_fst.mat <- as.matrix(pairwise.WCfst(temp3, 
                                          diploid = TRUE))
round(temp3_fst.mat, 3)
```
Jost's D
```{r}
library(mmod)

# Temporal Group 1
temp1.D.mat <- as.matrix(pairwise_D(temp1, hsht_mean = "arithmetic"))
round(temp1.D.mat, 3)

# Temporal Group 2
temp2.D.mat <- as.matrix(pairwise_D(temp2, hsht_mean = "arithmetic"))
round(temp2.D.mat, 3)

# Temporal Group 3
temp3.D.mat <- as.matrix(pairwise_D(temp3, hsht_mean = "arithmetic"))
round(temp3.D.mat, 3)
```
Hedrick G'st
```{r}
library(mmod)
# Temporal Group 1
temp1.Gst.mat <- pairwise_Gst_Hedrick(temp1,
                                      linearized = T)

round(temp1.Gst.mat, 3)

# Temporal Group 2
temp2.Gst.mat <- pairwise_Gst_Hedrick(temp2,
                                      linearized = T)

round(temp2.Gst.mat, 3)

# Temporal Group 3
temp3.Gst.mat <- pairwise_Gst_Hedrick(temp3,
                                      linearized = T)

round(temp3.Gst.mat, 3)

?mmod
```


