---
title: "PCA_STRUCT_2"
author: "Alex Depew"
date: "2026-03-05"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Revised_cleandata_11.03.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
library(ggplot2)
library(ggpubr)
library(grid)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_11.03.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=2015,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp2.sum <- summary(temp2)
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 

```{r}
# List of all herds
all_herds <- sort(unique(cleaned.data$HERD))

# Colour Palette selection for PCA
herd_palette <- c("#0026FF", 
                  "#FF5F00",
                  "#FF00FF", 
                  "#005000",
                  "#333333", 
                  "#660099", 
                  "#00E5FF", 
                  "#A52A2A", 
                  "#DAA520", 
                  "#1E90FF", 
                  "#228B22", 
                  "#cecece", 
                  "#a559aa", 
                  "#59a89c", 
                  "#f0c571", 
                  "#e02b35", 
                  "#082a54"
                  )

# 3. Assign names to the colors so they are "locked" to the herd ID
herd_cols <- setNames(as.character(herd_palette), all_herds)


# STRUCTURE colours
master_cols <- c("#ee964c", "#3b96cf", "#7d2a7c", "#e39bc1", "#e73f4e", 
                 "#4b9b46", "#f5df93", "#858434", "#ba552b", "#81c7c7", 
                 "#2d4a5e", "#44c29a", "#542e44", "#ff7f32", "#607d8b", "#1b4332")

```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp1 <- 100*pca.temp1$eig/sum(pca.temp1$eig)

x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp2 <- 100*pca.temp2$eig/sum(pca.temp2$eig)

##### -------
# Strictly for plotting (legend)
x.global <- tab(bison.genind, freq = TRUE, NA.method = "mean")
pca.global <- dudi.pca(x.global, center = T, scale = F, scannf = F, nf =2)
# Axes % Variance
eig.perc.global <- 100*pca.global$eig/sum(pca.global$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-3, 2.1)
```


## PCA plot with ggplot
```{r}
temp1@pop <- factor(temp1@pop, levels = sort(unique(as.character(temp1@pop))))
levels_temp1 <- levels(temp1@pop)
temp1_cols <- herd_cols[levels_temp1]


# Plot PCA object
pca1 <- fviz_pca_ind(pca.temp1,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp1@pop,
             repel = TRUE,
             palette = temp1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = "(1986 - 2015)",
       x = paste0("PC1 (", round(eig.perc.temp1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pca1
```

```{r}
temp2@pop <- factor(temp2@pop, levels = sort(unique(as.character(temp2@pop))))
levels_temp2 <- levels(temp2@pop)
temp2_cols <- herd_cols[levels_temp2]

# Invert Y axis
pca.temp2$li[ , 2] <- pca.temp2$li[ , 2] * -1

# Plot PCA object
pca2 <- fviz_pca_ind(pca.temp2,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp2@pop,
             repel = TRUE,
             palette = temp2_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = "(2016 - 2025)",
       x = paste0("PC1 (", round(eig.perc.temp2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pca2
```

```{r}
bison.genind@pop <- factor(bison.genind@pop, levels = sort(unique(as.character(bison.genind@pop))))
sorted_levels <- levels(bison.genind@pop)
sorted_cols <- herd_cols[sorted_levels]

global_cols <- herd_cols[sorted_levels]

# Plot PCA object
pcag <- fviz_pca_ind(pca.global,
             axes = c(1, 2),
             geom = "blank",
             habillage = bison.genind@pop,
             repel = TRUE,
             palette = global_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = bison.genind@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = bison.genind@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 
  
  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.global[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.global[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 6, shape = 19)))

```

### Arrange Plots together
```{r}
pca1_nl <- pca1 + theme(legend.position = "none")
pca2_nl <- pca2 + theme(legend.position = "none")

legend1 <- get_legend(pcag)

global_pcas <- grid.arrange(pca1_nl, legend1, pca2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/global_pcas.svg",
       plot = global_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
Remove Elk Island Plains for Structure Analyses
```{r}
temp1.noEINPP <- temp1[temp1@pop != "EINPP"]

temp2.noEINPP <- temp2[temp2@pop != "EINPP"]

bison.genind.noEINPP <- bison.genind[bison.genind@pop != "EINPP"]
```


### Perform PCA on both datasets (No EINPP)
```{r}
# Perform PCA
x.temp1.noEINPP <- tab(temp1.noEINPP, freq = TRUE, NA.method = "mean")
pca.temp1.noEINPP <- dudi.pca(x.temp1.noEINPP, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp1.noEINPP <- 100*pca.temp1.noEINPP$eig/sum(pca.temp1.noEINPP$eig)

x.temp2.noEINPP <- tab(temp2.noEINPP, freq = TRUE, NA.method = "mean")
pca.temp2.noEINPP <- dudi.pca(x.temp2.noEINPP, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp2.noEINPP <- 100*pca.temp2.noEINPP$eig/sum(pca.temp2.noEINPP$eig)

##### -------
# Strictly for plotting (legend)
x.global.noEINPP <- tab(bison.genind.noEINPP, freq = TRUE, NA.method = "mean")
pca.global.noEINPP <- dudi.pca(x.global.noEINPP, center = T, scale = F, scannf = F, nf =2)
# Axes % Variance
eig.perc.global.noEINPP <- 100*pca.global.noEINPP$eig/sum(pca.global.noEINPP$eig)
```

#### Reset xlim and ylim
```{r}
xlims.noEINPP <- c(-2.5, 3.5)
ylims.noEINPP <- c(-3, 2.1)
```


#### Temp1 (No EINPP)
```{r}
temp1.noEINPP@pop <- factor(temp1.noEINPP@pop, levels = sort(unique(as.character(temp1.noEINPP@pop))))
levels_temp1.noEINPP <- levels(temp1.noEINPP@pop)
temp1.noEINPP_cols <- herd_cols[levels_temp1.noEINPP]


# Plot PCA object
pca1.noEINPP <- fviz_pca_ind(pca.temp1.noEINPP,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp1.noEINPP@pop,
             repel = TRUE,
             palette = temp1.noEINPP_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp1.noEINPP@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp1.noEINPP@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims.noEINPP, ylim = ylims.noEINPP) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = "(1986 - 2015)",
       x = paste0("PC1 (", round(eig.perc.temp1.noEINPP[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp1.noEINPP[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 6, shape = 19)))

       

pca1.noEINPP
```

#### Temp2 (No EINPP)
```{r}
temp2.noEINPP@pop <- factor(temp2.noEINPP@pop, levels = sort(unique(as.character(temp2.noEINPP@pop))))
levels_temp2.noEINPP <- levels(temp2.noEINPP@pop)
temp2.noEINPP_cols <- herd_cols[levels_temp2.noEINPP]

pca.temp2.noEINPP$li[ ,2] <- pca.temp2.noEINPP$li[ , 2] *-1

# Plot PCA object
pca2.noEINPP <- fviz_pca_ind(pca.temp2.noEINPP,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp2.noEINPP@pop,
             repel = TRUE,
             palette = temp2.noEINPP_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp2.noEINPP@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp2.noEINPP@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims.noEINPP, ylim = ylims.noEINPP) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = "(2016 - 2025)",
       x = paste0("PC1 (", round(eig.perc.temp2.noEINPP[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp2.noEINPP[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 6, shape = 19)))

       

pca2.noEINPP
```

#### Global Dataset (No EINPP)
```{r}
bison.genind.noEINPP@pop <- factor(bison.genind.noEINPP@pop, levels = sort(unique(as.character(bison.genind.noEINPP@pop))))
sorted_levels.noEINPP <- levels(bison.genind.noEINPP@pop)
sorted_cols.noEINPP <- herd_cols[sorted_levels.noEINPP]

global_cols.noEINPP <- herd_cols[sorted_levels.noEINPP]

# Plot PCA object
pcag.noEINPP <- fviz_pca_ind(pca.global.noEINPP,
             axes = c(1, 2),
             geom = "blank",
             habillage = bison.genind.noEINPP@pop,
             repel = TRUE,
             palette = global_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = bison.genind.noEINPP@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = bison.genind.noEINPP@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 
  
  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.global.noEINPP[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.global.noEINPP[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 6, shape = 19)))

```

### Arrange Plots together (No EINPP)
```{r}
pca1_nl.noEINPP <- pca1.noEINPP + theme(legend.position = "none")
pca2_nl.noEINPP <- pca2.noEINPP + theme(legend.position = "none")

legend2 <- get_legend(pcag.noEINPP)

global_pcas.noEINPP <- grid.arrange(pca1_nl.noEINPP, legend2, pca2_nl.noEINPP, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
ggsave("./plots/global_pcas.noEINPP.svg",
       plot = global_pcas.noEINPP, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
# STRUCTURE Analysis 
```{r}
# Temp1 structure
temp1.str <- temp1
pop(temp1.str) <- as.numeric(as.factor(pop(temp1.str)))

genind_to_structure(temp1.str,
                    output = "STRUCTURE_files/temp1.w_plains.structure.txt")
# Temp2 structure
temp2.str <- temp2
pop(temp2.str) <- as.numeric(as.factor(pop(temp2.str)))

genind_to_structure(temp2.str,
                    output = "STRUCTURE_files/temp2.w_plains.structure.txt")

# Temp1.noEINPP
temp1.noEINPP.str <- temp1.noEINPP
pop(temp1.noEINPP.str) <- as.numeric(as.factor(pop(temp1.noEINPP.str)))

genind_to_structure(temp1.noEINPP.str,
                    output = "STRUCTURE_files/temp1.no_plains.structure.txt")

# Temp2.noEINPP
temp2.noEINPP.str <- temp2.noEINPP
pop(temp2.noEINPP.str) <- as.numeric(as.factor(pop(temp2.noEINPP.str)))

genind_to_structure(temp2.noEINPP.str,
                    output = "STRUCTURE_files/temp2.no_plains.structure.txt")

```

#### Load Data
```{r}
# List files
# Full datasets
sfiles_temp1 <- list.files("./STRUCTURE_files/results/temp1_w_plains/", full.names = TRUE)
sfiles_temp2 <- list.files("./STRUCTURE_files/results/temp2_w_plains/", full.names = TRUE)

# Qlist (pophelper)
slist_temp1 <- readQ(files = sfiles_temp1, filetype = "structure")
slist_temp2 <- readQ(files = sfiles_temp2, filetype = "structure")

# Summarise runs
struct_sum_temp1 <- summariseQ(tabulateQ(slist_temp1))
struct_sum_temp2 <- summariseQ(tabulateQ(slist_temp2))
```

#### Likelihood Plots
```{r}
## Temp1 With plains ----
# K plots
p1 <- evannoMethodStructure(data = struct_sum_temp1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp1_K",
                           exportpath = "./plots/K_plots/",
                           imgtype = "pdf")

p_lik <- p1$grobs[[1]]
p_dk <- p1$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_w1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./plots/K_plots/temp1.k.svg", 
       plot = evanno_w1, 
       width = 27, 
       height = 11)

# Temp2 With Plains ----
# K plots
p2 <- evannoMethodStructure(data = struct_sum_temp2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp2_K",
                           exportpath = "./plots/K_plots/",
                           imgtype = "pdf")

p_lik <- p2$grobs[[1]]
p_dk <- p2$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_w2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./plots/K_plots/temp2.k.svg", 
       plot = evanno_w2, 
       width = 27, 
       height = 11)
```

## STRUCTURE Plots (All K)
#### Temp 1 (With Plains)
```{r}
temp1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID))
# Align runs
slist_temp1 <- alignK(slist_temp1, "auto")

# Merge runs (average over 10 replicates)
temp1.mk2 <- mergeQ(slist_temp1[11:20])
temp1.mk3 <- mergeQ(slist_temp1[21:30])
temp1.mk4 <- mergeQ(slist_temp1[31:40])
temp1.mk5 <- mergeQ(slist_temp1[41:50])
temp1.mk6 <- mergeQ(slist_temp1[51:60])
temp1.mk7 <- mergeQ(slist_temp1[61:70])
temp1.mk8 <- mergeQ(slist_temp1[71:80])
temp1.mk9 <- mergeQ(slist_temp1[81:90])
temp1.mk10 <- mergeQ(slist_temp1[91:100])
temp1.mk11 <- mergeQ(slist_temp1[101:110])
temp1.mk12 <- mergeQ(slist_temp1[111:120])


temp1.merged.list <- as.qlist(list(temp1.mk2[[1]], temp1.mk3[[1]], temp1.mk4[[1]], 
                          temp1.mk5[[1]], temp1.mk6[[1]], temp1.mk7[[1]], temp1.mk8[[1]],
                          temp1.mk9[[1]], temp1.mk10[[1]], temp1.mk11[[1]], temp1.mk12[[1]]))

ind_label_temp1 <- readQ(files = sfiles_temp1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp1 <- rownames_to_column(ind_label_temp1, "ID")
locations_temp1 <- left_join(ind_label_temp1, temp1.data, by = "ID")

loc_labels_temp1 <- locations_temp1$HERD %>% as.data.frame()

herd_order_temp1 <- c("EINPP", "EINPW", "HZ", "MBS", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")
struct_temp1_cols <- c("#ee964c", "#3b96cf", "#7d2a7c","#e39bc1", "#e73f4e", "#4b9b46",  "#f5df93","#858434", "#ba552b", "#81c7c7",  "#2d4a5e", "#44c29a")

plotQ(temp1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            clustercol = master_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8", "K = 9", "K = 10", "K = 11", "K = 12"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "Temp1_w_plains",
            exportpath = "./plots/STRUCTURE_PLOTS")
```

#### Temp 2 (With Plains)
```{r}
temp2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID))
# Align runs
slist_temp2 <- alignK(slist_temp2, type = "across")

# Merge runs (average over 10 replicates)
temp2.mk2 <- mergeQ(slist_temp2[11:20])
temp2.mk3 <- mergeQ(slist_temp2[21:30])
temp2.mk4 <- mergeQ(slist_temp2[31:40])
temp2.mk5 <- mergeQ(slist_temp2[41:50])
temp2.mk6 <- mergeQ(slist_temp2[51:60])
temp2.mk7 <- mergeQ(slist_temp2[61:70])
temp2.mk8 <- mergeQ(slist_temp2[71:80])
temp2.mk9 <- mergeQ(slist_temp2[81:90])
temp2.mk10 <- mergeQ(slist_temp2[91:100])
temp2.mk11 <- mergeQ(slist_temp2[101:110])
temp2.mk12 <- mergeQ(slist_temp2[111:120])
temp2.mk13 <- mergeQ(slist_temp2[121:130])
temp2.mk14 <- mergeQ(slist_temp2[131:140])
temp2.mk15 <- mergeQ(slist_temp2[141:150])
temp2.mk16 <- mergeQ(slist_temp2[151:160])

temp2.merged.list <- as.qlist(list(temp2.mk2[[1]], temp2.mk3[[1]], temp2.mk4[[1]], 
                          temp2.mk5[[1]], temp2.mk6[[1]], temp2.mk7[[1]], temp2.mk8[[1]],
                          temp2.mk9[[1]], temp2.mk10[[1]], temp2.mk11[[1]], temp2.mk12[[1]],
                          temp2.mk13[[1]], temp2.mk14[[1]], temp2.mk15[[1]], temp2.mk16[[1]]))

ind_label_temp2 <- readQ(files = sfiles_temp2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp2 <- rownames_to_column(ind_label_temp2, "ID")
locations_temp2 <- left_join(ind_label_temp2, temp2.data, by = "ID")

loc_labels_temp2 <- locations_temp2$HERD %>% as.data.frame()

herd_order_temp2 <- c("EINPP", "AISH", "CL", "EINPW", "ETH", "HZ", "NAH", "NORD", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

struct_temp2_cols <- c("#3b96cf", 
  "#ee964c", "#7d2a7c","#e39bc1", "#e73f4e", 
  "#4b9b46",  "#f5df93","#858434", "#ba552b", 
  "#81c7c7",  "#2d4a5e", "#44c29a",
  "#542e44", "#ff7f32", "#607d8b", "#1b4332"
)

plotQ(alignK(temp2.merged.list), 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp2, 
            subsetgrp = herd_order_temp2, 
            clustercol = master_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", 
                      "K = 8", "K = 9", "K = 10", "K = 11", "K = 12", "K = 13",
                      "K = 14", "K = 15", "K = 16"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "Temp2_with_plains",
            exportpath = "./plots/STRUCTURE_PLOTS")
```

# STRUCTURE Plots (No Elk Island Plains outgroup)
#### Load Data
```{r}
# List files
# Full datasets
sfiles_temp1_no_plains <- list.files("./STRUCTURE_files/results/temp1_no_plains/", full.names = TRUE)
sfiles_temp2_no_plains <- list.files("./STRUCTURE_files/results/temp2_no_plains/", full.names = TRUE)

# Qlist (pophelper)
slist_temp1_no_plains <- readQ(files = sfiles_temp1_no_plains, filetype = "structure")
slist_temp2_no_plains <- readQ(files = sfiles_temp2_no_plains, filetype = "structure")

# Summarise runs
struct_sum_temp1_no_plains <- summariseQ(tabulateQ(slist_temp1_no_plains))
struct_sum_temp2_no_plains <- summariseQ(tabulateQ(slist_temp2_no_plains))
```

#### Likelihood Plots
```{r}
## Temp1 With plains ----
# K plots
t1 <- evannoMethodStructure(data = struct_sum_temp1_no_plains, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp1_no_plains_K",
                           exportpath = "./plots/K_plots/",
                           imgtype = "pdf")

p_lik <- t1$grobs[[1]]
p_dk <- t1$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_t1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./plots/K_plots/temp1_no_plains.k.svg", 
       plot = evanno_t1, 
       width = 27, 
       height = 11)

# Temp2 With Plains ----
# K plots
t2 <- evannoMethodStructure(data = struct_sum_temp2_no_plains, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp2_no_plains_K",
                           exportpath = "./plots/K_plots/",
                           imgtype = "pdf")

p_lik <- t2$grobs[[1]]
p_dk <- t2$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_t2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./plots/K_plots/temp2_no_plains.k.svg", 
       plot = evanno_t2, 
       width = 27, 
       height = 11)
```

## STRUCTURE Plots (All K)
#### Temp 1 (Without Plains)
```{r}
temp1_no_plains.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(HERD != "EINPP") %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID))
# Align runs
slist_temp1_no_plains <- alignK(slist_temp1_no_plains, "auto")

# Merge runs (average over 10 replicates)
temp1_no_plains.mk2 <- mergeQ(slist_temp1_no_plains[11:20])
temp1_no_plains.mk3 <- mergeQ(slist_temp1_no_plains[21:30])
temp1_no_plains.mk4 <- mergeQ(slist_temp1_no_plains[31:40])
temp1_no_plains.mk5 <- mergeQ(slist_temp1_no_plains[41:50])
temp1_no_plains.mk6 <- mergeQ(slist_temp1_no_plains[51:60])
temp1_no_plains.mk7 <- mergeQ(slist_temp1_no_plains[61:70])
temp1_no_plains.mk8 <- mergeQ(slist_temp1_no_plains[71:80])
temp1_no_plains.mk9 <- mergeQ(slist_temp1_no_plains[81:90])
temp1_no_plains.mk10 <- mergeQ(slist_temp1_no_plains[91:100])
temp1_no_plains.mk11 <- mergeQ(slist_temp1_no_plains[101:110])



temp1.merged.list <- as.qlist(list(temp1_no_plains.mk2[[1]], temp1_no_plains.mk3[[1]], temp1_no_plains.mk4[[1]], 
                          temp1_no_plains.mk5[[1]], temp1_no_plains.mk6[[1]], temp1_no_plains.mk7[[1]], temp1_no_plains.mk8[[1]],
                          temp1_no_plains.mk9[[1]], temp1_no_plains.mk10[[1]], temp1_no_plains.mk11[[1]]))

ind_label_temp1_no_plains <- readQ(files = sfiles_temp1_no_plains[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp1_no_plains <- rownames_to_column(ind_label_temp1_no_plains, "ID")
locations_temp1_no_plains <- left_join(ind_label_temp1_no_plains, temp1_no_plains.data, by = "ID")

loc_labels_temp1_no_plains <- locations_temp1_no_plains$HERD %>% as.data.frame()

herd_order_temp1_no_plains <- c("EINPW", "HZ", "MBS", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")
struct_temp1_no_plains_cols <- c(
  "#ee964c", "#3b96cf", "#7d2a7c", "#e39bc1", "#e73f4e", 
  "#4b9b46", "#f5df93", "#858434", "#ba552b", 
  "#81c7c7", "#2d4a5e"
)

plotQ(temp1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp1_no_plains, 
            subsetgrp = herd_order_temp1_no_plains, 
            clustercol = master_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8", "K = 9", "K = 10", "K = 11"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "Temp1_no_plains",
            exportpath = "./plots/structure_plots/")
```

#### Temp 2 (Without Plains)
```{r}
temp2_no_plains.data <- cleaned.data %>% 
  select(-ID) %>%
  filter(HERD != "EINPP") %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID))
# Align runs
slist_temp2_no_plains <- alignK(slist_temp2_no_plains, "auto")

# Merge runs (average over 10 replicates)
temp2_no_plains.mk2 <- mergeQ(slist_temp2_no_plains[11:20])
temp2_no_plains.mk3 <- mergeQ(slist_temp2_no_plains[21:30])
temp2_no_plains.mk4 <- mergeQ(slist_temp2_no_plains[31:40])
temp2_no_plains.mk5 <- mergeQ(slist_temp2_no_plains[41:50])
temp2_no_plains.mk6 <- mergeQ(slist_temp2_no_plains[51:60])
temp2_no_plains.mk7 <- mergeQ(slist_temp2_no_plains[61:70])
temp2_no_plains.mk8 <- mergeQ(slist_temp2_no_plains[71:80])
temp2_no_plains.mk9 <- mergeQ(slist_temp2_no_plains[81:90])
temp2_no_plains.mk10 <- mergeQ(slist_temp2_no_plains[91:100])
temp2_no_plains.mk11 <- mergeQ(slist_temp2_no_plains[101:110])
temp2_no_plains.mk12 <- mergeQ(slist_temp2_no_plains[111:120])
temp2_no_plains.mk13 <- mergeQ(slist_temp2_no_plains[121:130])
temp2_no_plains.mk14 <- mergeQ(slist_temp2_no_plains[131:140])
temp2_no_plains.mk15 <- mergeQ(slist_temp2_no_plains[141:150])


temp2_no_plains.merged.list <- as.qlist(list(temp2_no_plains.mk2[[1]], temp2_no_plains.mk3[[1]], 
                          temp2_no_plains.mk4[[1]], 
                          temp2_no_plains.mk5[[1]], temp2_no_plains.mk6[[1]], 
                          temp2_no_plains.mk7[[1]], temp2_no_plains.mk8[[1]],
                          temp2_no_plains.mk9[[1]], temp2_no_plains.mk10[[1]], 
                          temp2_no_plains.mk11[[1]], temp2_no_plains.mk12[[1]],
                          temp2_no_plains.mk13[[1]], temp2_no_plains.mk14[[1]], 
                          temp2_no_plains.mk15[[1]]))

ind_label_temp2_no_plains <- readQ(files = sfiles_temp2_no_plains[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp2_no_plains <- rownames_to_column(ind_label_temp2_no_plains, "ID")
locations_temp2_no_plains <- left_join(ind_label_temp2_no_plains, temp2_no_plains.data, by = "ID")

loc_labels_temp2_no_plains <- locations_temp2_no_plains$HERD %>% as.data.frame()

herd_order_temp2_no_plains <- c("AISH", "CL", "EINPP", "EINPW", "ETH", "HZ", "NAH", "NORD", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

struct_temp2_no_plains_cols <- c("#3b96cf",
  "#ee964c",  "#7d2a7c","#e39bc1", "#e73f4e", 
  "#4b9b46",  "#f5df93", "#858434", "#ba552b", 
  "#81c7c7", "#2d4a5e", "#44c29a",
  "#542e44", "#ff7f32", "#607d8b"
)

plotQ(temp2_no_plains.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp2_no_plains, 
            subsetgrp = herd_order_temp2_no_plains, 
            clustercol = master_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", 
                      "K = 8", "K = 9", "K = 10", "K = 11", "K = 12", "K = 13",
                      "K = 14", "K = 15"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "Temp2_no_plains",
            exportpath = "./plots/structure_plots/")
```





# Combined Plots
```{r}
### PCA w/ plains bison outgroup

# PCA w/ plains T1
pca1_tight <- pca1_nl + 
  theme(
    text = element_text(size = 20)
    )


# STRUCTURE T1 w/ plains
plot1 <- plotQ(c(temp1.mk2, temp1.mk3, temp1.mk4, temp1.mk5, temp1.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4", "K=5*", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plottemp1.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
t1 <- plot1$plot[[1]]

# Combine Global w/ plains T1
row_1 <- grid.arrange(pca1_tight, t1,
                      ncol = 2,
                      widths = c(5, 7.5))



# PCA w/ plains T2
pca2_tight <- pca2_nl + 
  theme(
    text = element_text(size = 20)
    )


# STRUCTURE T1 w/ plains
plot2 <- plotQ(c(temp2.mk2, temp2.mk3, temp2.mk4, temp2.mk5, temp2.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_temp2, 
            subsetgrp = herd_order_temp2, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4", "K=5*", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plottemp2.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
t2 <- plot2$plot[[1]]


# Combine Global w/ plains T1
row_2 <- grid.arrange(pca2_tight, t2,
                      ncol = 2,
                      widths = c(10, 15))

################ Without Plains from Elk Island

# PCA w/ plains T1
pca1.noEINPP_tight <- pca1_nl.noEINPP + 
  theme(
    text = element_text(size = 20)
    )


# STRUCTURE T1 w/ plains
plot1_no_plains <- plotQ(c(temp1_no_plains.mk2, temp1_no_plains.mk3, temp1_no_plains.mk4, temp1_no_plains.mk5, temp1_no_plains.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_temp1_no_plains, 
            subsetgrp = herd_order_temp1_no_plains, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4*", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plottemp1_no_plains.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
t1_no_plains <- plot1_no_plains$plot[[1]]

# Combine Global w/ plains T1
row_3 <- grid.arrange(pca1.noEINPP_tight, t1_no_plains,
                      ncol = 2,
                      widths = c(10, 15))

# PCA w/o plains T2
pca2.noEINPP_tight <- pca2_nl.noEINPP + 
  theme(
    text = element_text(size = 20)
    )

# STRUCTURE T1 w/ plains
plot2_no_plains <- plotQ(c(temp2_no_plains.mk2, temp2_no_plains.mk3, temp2_no_plains.mk4, temp2_no_plains.mk5, temp2_no_plains.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_temp2_no_plains, 
            subsetgrp = herd_order_temp2_no_plains, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4*", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plottemp2_no_plains.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
t2_no_plains <- plot2_no_plains$plot[[1]]

# Combine Global w/ plains T1
row_4 <- grid.arrange(pca2.noEINPP_tight, t2_no_plains,
                      ncol = 2,
                      widths = c(10, 15))


#### Label Sections
# 3. Apply the 'A' Label
A_labeled <- arrangeGrob(row_1, 
                         top = textGrob("A (1986 - 2015)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

B_labeled <- arrangeGrob(row_2, 
                         top = textGrob("B (2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

top_half <- arrangeGrob(A_labeled, 
                        rectGrob(gp = gpar(col = NA)), # Spacer between rows
                        B_labeled, 
                        ncol = 1, 
                        heights = c(10, 1.5, 10),
                        top = textGrob("Including Plains Bison Outgroup", 
                                       gp = gpar(fontsize = 35, fontface = "bold")))

C_labeled <- arrangeGrob(row_3, 
                         top = textGrob("C (1986 - 2015)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

D_labeled <- arrangeGrob(row_4, 
                         top = textGrob("D (2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold"))) 

bottom_half <- arrangeGrob(C_labeled, 
                           rectGrob(gp = gpar(col = NA)), # Spacer between rows
                           D_labeled, 
                           ncol = 1, 
                           heights = c(10, 1.5, 10), 
                           top = textGrob("Excluding Plains Bison Outgroup", 
                                          gp = gpar(fontsize = 35, fontface = "bold")))

#### Combine PCAs and STRUCTURE barplots
comb <- arrangeGrob(top_half, 
                    rectGrob(gp = gpar(col = NA)), # Blank space
                    bottom_half,
                    ncol = 1, 
                    heights = c(20, 2, 20)) # Give the middle some space to account for the separating line


#### Pull legend from global PCA
pcag_horizontal <- pcag + 
  theme(legend.direction = "horizontal", 
        legend.position = "top",

        legend.justification = "left", 
        legend.box.just = "top",
        legend.text = element_text(size = 20), 
        legend.title = element_text(size = 25, face = "bold"),
        legend.margin = margin(l = 0)) + 
  guides(color = guide_legend(
    title = "Herd", 
    nrow = 3, 
    byrow = TRUE,
    override.aes = list(size = 7, alpha = 1) 
  ))

# Isolate legend
legend_bottom <- get_legend(pcag_horizontal)

# Legend move to left
legend_shifted <- arrangeGrob(legend_bottom, 
                              vp = viewport(x = 0.08, just = "left", width = 0.8)) 

# Combine Structuring and legend
comb_leg <- grid.arrange(
  comb, 
  legend_shifted,
  ncol = 1, 
  heights = c(46, 4), 
  padding = unit(0, "line")
)

comb_leg_sep <-grobTree(comb_leg, linesGrob(x = unit(c(0, 1), "npc"), 
                                       y = unit(0.543, "npc"), 
                                       gp = gpar(col = "black", lwd = 2)))
 

# Save plots
ggsave("./plots/combined_global_leg_sep.svg",
       plot = comb_leg_sep,
       height = 25,
       width = 20,
       bg = "white",
       units = "in")
```


# Plains Admixture
#### Assess Q-values (membership probabilities of assigning to clusters)
```{r}
temp1.order <- data.frame(plot1$data$grplab[[2]]) %>% 
  rownames_to_column("Ind")

temp1.k2 <- data.frame(temp1.mk2) %>% 
  rownames_to_column("Ind") %>% 
  left_join(temp1.order, by = "Ind") %>% 
  rename(Herd = ".",
         Wood = X2.Cluster1,
         Plains = X2.Cluster2) %>% 
  select(-Ind) %>% 
  relocate(Herd, .before = Wood)
  
temp1.Q.sum <- temp1.k2 %>% 
  group_by(Herd) %>% 
  summarise(
    Wood_mean = mean(Wood),
    Wood_SD = sd(Wood),
    Plains_mean = mean(Plains),
    Plains_SD = sd(Plains)
  )

# Summary per herd of Q assignments
temp1.Q.sum

temp2.order <- data.frame(plot2$data$grplab[[2]]) %>% 
  rownames_to_column("Ind")

temp2.k2 <- data.frame(temp2.mk2) %>% 
  rownames_to_column("Ind") %>% 
  left_join(temp2.order, by = "Ind") %>% 
  rename(Herd = ".",
         Wood = X2.Cluster2, # Swapped from previous
         Plains = X2.Cluster1) %>% 
  select(-Ind) %>% 
  relocate(Herd, .before = Wood)
  
temp2.Q.sum <- temp2.k2 %>% 
  group_by(Herd) %>% 
  summarise(
    Wood_mean = mean(Wood),
    Wood_SD = sd(Wood),
    Plains_mean = mean(Plains),
    Plains_SD = sd(Plains)
  )

# Summary per herd of Q assignments
temp2.Q.sum
```







