---
title: "Locus filtering"
author: "Alex Depew"
date: "2026-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is for data quality control, specificially locus quality control, to ensure accuracy of subsequent analyses.
Files used described below:

- `Cleaned_samples.24.12.2025`
   - Contains all bison microsatellite data from wood and plains bison with added metadata for year collected

## Load Required Packages
```{r warning = FALSE, message = FALSE}
rm(list = ls())
library(adegenet)
library(PopGenReport)
library(poppr)
library(pegas)
library(strataG)
library(tidyverse)
library(graph4lg)
library(svglite)
```

## Load Required Data
```{r}
bison.data <- read.csv("./datafiles/Cleaned_samples.16.02.2026.csv",
                       check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN
```

## Locus Filtering
Preliminary locus filtering relied on *a priori* information into X-linked loci, followed by the removal of sex markers (not used in analyses). Further tests include:

- Test for polymorphism (do the markers exhibit sufficient variability to be informative?)
- Test for HWE deviations
- Test for linkage disequilibrium
- Null alleles
- Allelic Dropout

## Determine percent of complete genotypes by loci
```{r}
locmiss_bison <-  propTyped(bison.genind, by = "loc")
locmiss_bison[which(locmiss_bison < 0.80)] # print loci with < 80% complete genotypes

barplot(locmiss_bison, ylim = c(0,1), ylab = "Complete genotypes (proportion)", xlab = "Locus", las = 2, cex.names = 0.7)

bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)

```
Locus BM4028 contained more >20% missing data, and therefore was removed from the global dataset.

## Check loci are polymorphic
```{r}
summary(bison.genind)
isPoly(bison.genind) %>% summary
```
All loci are polymorphic, as "TRUE" is returned by the *isPoly* function. The summary of the genind object indicates that the loci contain between 4 and 14 loci. No loci are monomorphic, and therefore none must be removed for being uninformative.

# Data Subsetting
Subsequent genotype quality control measures must be completed within temporal groups, within herds, as to not induce the Wahlund effect. By evaluating deviations from Hardy-Weinberg Equilibrium (HWE), Linkage Disequilibrium (LD), and the presence of null alleles at this granular level, it is possible to distinguish between localized biological variation and systematic technical errors that persist across the entire population.

Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=2015,]

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN >=2016, ]

# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1)) # Remove ETH (n = 6) and NORD (n = 5). We will keep WENT to compare with WBNP 
table(pop(temp2)) # Remove "WBNP" samples without subherd designations


# Remove herds from temporal group 2 
remove_herds_temp1 <- c("ETH", "NORD")
temp1 <- temp1[!temp1@pop %in% remove_herds_temp1]

remove_herds_temp2 <- "WBNP"
temp2 <- temp2[!temp2@pop %in% remove_herds_temp2]

# Create genind summaries for both temporal groupings
temp1.sum <- summary(temp1)
temp2.sum <- summary(temp2)
```

## Test for HWE deviations
Because of the variance in sample sizes, and low sample sizes seen in some herds, a Markov Chain Monte Carlo exact test can be used to evaluate deviations from HWE.
```{r}
# In order to use the genepop's MCMC method of deterimining HWE deviations, the strataG implementation can be used. the bison "genind" objects must be reformatted as "gtypes" objects
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)

# Further, HWE deviations and linkage disequilibrium are tested within each population. Deviations or LD which occur across populations (at the same locus/loci) may indicate technical errors, adaptation, or linkage
pop_list_temp1 <- strataSplit(temp1.g)
pop_list_temp2 <- strataSplit(temp2.g)
```

### Test for HWE
Test for hardy weinberg equilibrium intra-population
```{r}
# Apply HWE test from genepop. The strataG wrapper formats this back into R nicely for subsequent manipulation

# Temp1 ----
hwe_results_list_temp1 <- lapply(pop_list_temp1, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp1 <- as.data.frame(hwe_results_list_temp1) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 

# Temp2 ----
hwe_results_list_temp2 <- lapply(pop_list_temp2, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp2 <- as.data.frame(hwe_results_list_temp2) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 
```

### Summarize HWE Deviations
```{r}
# Temp1 ----
temp1.HWE.sum <- HWE.temp1 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp1.HWE.sum)

# Temp2 ----
temp2.HWE.sum <- HWE.temp2 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp2.HWE.sum)
```

#### HWE Sig table
```{r}
# Temp1
HWE.sig.temp1 <- HWE.temp1 %>% 
  filter(HWE_p_value <= 0.05)

wide_HWE.temp1 <- HWE.temp1 %>%
  select(Locus, Population, p.adj_herd) %>%
  pivot_wider(
    names_from = Population, 
    values_from = p.adj_herd
  )  %>% 
  mutate(across(-Locus, ~ {
    # 1. Force 4 decimal places (turns 1 into "1.0000")
    formatted <- sprintf("%.4f", .)
    
    # 2. Add asterisk if the ORIGINAL numeric value was <= 0.05
    if_else(. <= 0.05, paste0(formatted, "*"), formatted)
  }))

write.csv(wide_HWE.temp1, "./tables/temp1_v2.HWE.p_adjust.csv", row.names = F)


# Temp2 
wide_HWE.temp2 <- HWE.temp2 %>%
  select(Locus, Population, p.adj_herd) %>%
  pivot_wider(
    names_from = Population, 
    values_from = p.adj_herd
  ) %>% 
  mutate(across(-Locus, ~ {
    # 1. Force 4 decimal places (turns 1 into "1.0000")
    formatted <- sprintf("%.4f", .)
    
    # 2. Add asterisk if the ORIGINAL numeric value was <= 0.05
    if_else(. <= 0.05, paste0(formatted, "*"), formatted)
  }))

write.csv(wide_HWE.temp2, "./tables/temp2_v2.HWE.p_adjust.csv", row.names = F)

```

## Test for Linkage Disequilibrium
Pairwise test for LD using Genepop through *strataG*. If LD is present in loci across the majority of herds, it may indicate physical linkage as opposed to local variation.
```{r}
# Temp1 ----
LD_results_temp1 <- lapply(pop_list_temp1, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp1 <- bind_rows(LD_results_temp1, .id = "Population") # Put into dataframe format

LD.df.temp1 <- LD.df.temp1 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp1.sig <- LD.df.temp1 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp1.sig, "./tables/temp1LD_v2.csv", row.names = F)


# Temp2 ----
LD_results_temp2 <- lapply(pop_list_temp2, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp2 <- bind_rows(LD_results_temp2, .id = "Population") # Put into dataframe format

LD.df.temp2 <- LD.df.temp2 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp2.sig <- LD.df.temp2 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp2.sig, "./tables/temp2LD_v2.csv", row.names = F)
```
The dataframes containing only significant pairwise comparisons following the Benjamini-Hochberg correction return significant p-values in populations which are represented by relatively small sample sizes: Nahanni (n = 8). Instead of indicating true linkage disequilibrium, the lack of intra-population allele data likely inflates the variance in allele frequency estimates, increasing the probability of detecting spurious non-random locus associations due to sampling effects. No significant pairs of loci exist on the same chromose, negating the possibility of physical linkage. 


## Test for null alleles
To test for null alleles and allelic dropout, a combination of diagnostic and correction methods can be used. Microchecker is an external program which can test for the presence of null alleles using heterozygote deficiency methods as described by Chakraborty (1994) and Brookfield (1996). It also assess the presence of short allele dominance by comparing the proportion of short to long repeats. In order to use this program, datasets must be reformatted into genepop format using the package *graph4lg*. The estimation maximization method as outlined by Dempster et al. (1977) provides a more accurate estimation of null allele frequency and is used in *FreeNA*. Files must be in genepop* format.
```{r}
# Format for Micro-Checker
genind_to_genepop(temp1,
                  output = "./microchecker/temp1.microchecker.txt")

genind_to_genepop(temp2,
                  output = "microchecker/temp2.microchecker.txt")

# Format for FreeNA
genind_to_genepop(temp1,
                  output = "./FreeNA/temp1.genepop.txt")

genind_to_genepop(temp2,
                  output = "./FreeNA/temp2.genepop.txt")

# *Despite FreeNA taking "genepop" input files, the POP labels must be stacked as a column instead of as a row. This can be done by a) copying each of the files in the FreeNA file folder to a new "tempX.FreeNA.txt" file, b) editing the locus names from being in row format to stacked, with no comma separation.
```

### Obtain null allele frequency
Micro-checker can be used as a diagnostic tool for assessing the presence of null alleles in the data. The methods are based on heterozygosity deficiency methods.

#### FreeNA Results
```{r}
# Locus Labels
loc_lab <- c(
  "BMS1117", "TGLA53", "BMC4214", "BM1225", "BM47", 
  "BMS1315", "IL4", "BM2113", "BMS1716", "BM1706", 
  "BMS1675", "BMS1857", "BMS1001", "BMS601", "BM720", 
  "ILSTS102", "BMS1074", "BMS1172", "BMS2258", "BM757", 
  "SPS115", "BMS527", "BM2830", "INRA194", "BMS941", 
  "INRA037", "BM4107", "TGLA122", "HUJ246", "BMS528", 
  "BM4440", "TGLA44", "BMS510", "BMS1355", "BM1824", 
  "BMS812", "BMS2639", "BMS410", "AGLA232", "BM17132", 
  "BM711", "INRA133", "RM372", "BM4311", "BMS1747", 
  "BL1036", "BM1862", "BMS1862", "URB011"
)

# Match locus to numeric labels in null file
loc_match <- tibble(
  Locus_ID = 1:49, 
  Locus = loc_lab
)

# Temp1 ----
temp1_pops_lab <- c("EINPP", "EINPW", "HZ", "MBS", "RL", "WAB", "WBNPGR", 
                 "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

# Match population to numeric labels in null file
temp1_pop_match <- tibble(
  Pop = 1:12,
  Population = temp1_pops_lab
)

temp1.FreeNA.nulls <- read.table("./FreeNA/FreeNA_out/temp1.FreeNA.nullfreq.txt",
                          col.names = c("Locus_ID", "Pop", "FreeNA"))

temp1.FreeNA.nulls <- left_join(temp1.FreeNA.nulls, loc_match, by = "Locus_ID") %>% 
  select(-Locus_ID) %>% 
  left_join(temp1_pop_match, temp1.FreeNA.nulls, by = "Pop") %>% 
  select(-Pop) %>% 
  relocate(FreeNA, .after = Population) %>% 
  relocate(Locus, .after = Population)

# Temp2 ----
temp2_pops_lab <- c("AISH", "CL", "EINPP",
                    "EINPW", "ETH", "HZ",
                    "NAH", "NORD", "RL",
                    "WAB", "WBNPPGR", "WBNPHC",
                    "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

# Match population to numeric labels in null file
temp2_pop_match <- tibble(
  Pop = 1:16,
  Population = temp2_pops_lab
)

temp2.FreeNA.nulls <- read.table("./FreeNA/FreeNA_out/temp2.FreeNA.nullfreq.txt",
                          col.names = c("Locus_ID", "Pop", "FreeNA"))

temp2.FreeNA.nulls <- left_join(temp2.FreeNA.nulls, loc_match, by = "Locus_ID") %>% 
  select(-Locus_ID) %>% 
  left_join(temp2_pop_match, temp2.FreeNA.nulls, by = "Pop") %>% 
  select(-Pop) %>% 
  relocate(FreeNA, .after = Population) %>% 
  relocate(Locus, .after = Population)
```

#### Identify Significant Null Alleles
```{r}
# First, remove populations with sample sizes, as they are more likely to show heterozygote deficiencies and skew null allele calling
table(pop(temp1)) # WENT
table(pop(temp2)) # NAH, NORD, WAB, WENT

# Create significant dataframes of called nulls but remove the small population sizes
null.sig.1 <- temp1.FreeNA.nulls %>% 
  filter(#!Population == "WENT",
         FreeNA >= 0.10)

null.sig.2 <- temp2.FreeNA.nulls %>% 
  filter(#!Population %in% c("NAH", "NORD", "WAB", "WENT"), 
         FreeNA >= 0.10)

# Merge to see list across time 
n1 <- null.sig.1 %>% select(Population, Locus, T1 = FreeNA)
n2 <- null.sig.2 %>% select(Population, Locus, T2 = FreeNA)

compiled_patterns <- n1 %>%
  full_join(n2, by = c("Population", "Locus")) %>%
  # Replace any 0 with NA across all Timepoint columns
  mutate(across(starts_with("T"), ~na_if(., 0))) %>%
  # Calculate how many time points were NOT NA (actual occurrences)
  mutate(Prevalence = rowSums(!is.na(across(starts_with("T"))))) %>%
  arrange(desc(Prevalence), Population, Locus)

table(compiled_patterns$Locus)

write.csv(compiled_patterns, "./tables/Sig.Nulls_v2.csv", row.names = F)
```
#### Visualize Null Allele Frequency
Averaged at Loci
```{r}
temp1.null.avg <- temp1.FreeNA.nulls %>%
  group_by(Locus) %>%
  summarise(
    Mean = mean(FreeNA, na.rm = TRUE),
    SD = sd(FreeNA, na.rm = TRUE)
  ) %>%
  rename(Temp1_mean = Mean, Temp1_sd = SD)

temp2.null.avg <- temp2.FreeNA.nulls %>%
  group_by(Locus) %>%
  summarise(
    Mean = mean(FreeNA, na.rm = TRUE),
    SD = sd(FreeNA, na.rm = TRUE)
  ) %>%
  rename(Temp2_mean = Mean, Temp2_sd = SD)

avg.nulls <- temp1.null.avg %>%
  left_join(temp2.null.avg, by = "Locus") %>%
  pivot_longer(
    cols = -Locus, 
    names_to = c("Timepoint", ".value"), 
    names_sep = "_"
  )
  

ggplot(avg.nulls, aes(x = fct_rev(Locus), y = mean, fill = fct_rev(Timepoint))) + 
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = pmax(0, mean - sd), ymax = mean + sd),
    position = position_dodge(width = 0.8), 
    width = 0.4, 
    alpha = 0.7
  ) +
  coord_flip() +  
  scale_fill_manual(
    values = c("Temp1" = "steelblue", "Temp2" = "darkorange"),
    breaks = c("Temp1", "Temp2"), 
    labels = c("Temp1" = "T1", "Temp2" = "T2")
  ) +
  theme_minimal() +
  labs(
    x = "Locus", 
    y = "Mean Null Allele Rate (+SD)",
    fill = "Cohort" 
  ) +
  theme(
    # 1. Bold and Large Locus Names (Y-axis)
    axis.text.y = element_text(size = 9, face = "bold", color = "black"), 
    
    # 2. Bold and Large Rates (X-axis)
    axis.text.x = element_text(size = 9, face = "bold", color = "black"),
    
    # 3. Bold Axis Titles
    axis.title = element_text(size = 12, face = "bold"),
    
    # 4. Bold Legend Title and Text
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10, face = "bold"),
    
    panel.grid.major.y = element_blank(),
    legend.position = "right"
  )

ggsave(
  filename = "./plots/Locus_Nulls.svg", 
  plot = last_plot(),
  device = "svg",
  width = 10,      
  height = 14,    
  units = "in",
  dpi = 300,     
  bg = "white" 
)
```


## Test for allelic dropout
Microdrop requires structure input files, yet the structure input files do not work unless they are delimited with EXACTLY one space. Note: you must delete the "POP" and "IND" column headers in nano text editor prior to running the program, as the input requires N-2 column names (colnames = locnames = 49).
```{r}
# Microdrop requires STRUCTURE input files but they need to be space delimited
temp1.str <- temp1
pop(temp1.str) <- as.numeric(as.factor(pop(temp1.str)))

temp2.str <- temp2
pop(temp2.str) <- as.numeric(as.factor(pop(temp2.str)))

```


```{r}
# --- TEMP 1 ---
temp1_pop_list <- seppop(temp1.str)

lapply(names(temp1_pop_list), function(pop) {
    out_path <- paste0("./microdrop_2/temp/temp1_", pop, ".struct.txt")
    final_path <- paste0("./microdrop_2/input/temp1_", pop, ".microdrop.txt") # Define final_path
    
    genind_to_structure(temp1_pop_list[[pop]], output = out_path)
    temp_data <- read.table(out_path, skip = 1)
    
    # Create and write header
    locus_header <- c("Indiv", "Pop", locNames(temp1_pop_list[[pop]]))
    cat(paste(locus_header, collapse = " "), "\n", file = final_path)
    
    # Append data (Note: append = TRUE is critical here)
    write.table(temp_data, 
              file = final_path, 
              append = TRUE,
              sep = " ", 
              row.names = FALSE, 
              col.names = FALSE, 
              quote = FALSE)
    
    message("Finished processing population: ", pop)
})

# --- TEMP 2 ---
temp2_pop_list <- seppop(temp2.str)

lapply(names(temp2_pop_list), function(pop) {
    out_path <- paste0("./microdrop_2/temp/temp2_", pop, ".struct.txt")
    final_path <- paste0("./microdrop_2/input/temp2_", pop, ".microdrop.txt")
    
    genind_to_structure(temp2_pop_list[[pop]], output = out_path)
    temp_data <- read.table(out_path, skip = 1)
    
    locus_header <- c("Indiv", "Pop", locNames(temp2_pop_list[[pop]]))
    cat(paste(locus_header, collapse = " "), "\n", file = final_path)
    
    write.table(temp_data, 
              file = final_path, 
              append = TRUE,
              sep = " ", 
              row.names = FALSE, 
              col.names = FALSE, 
              quote = FALSE)
    
    message("Finished processing population: ", pop)
})
```

### All in One: Input files and Parameter files
```{r}
# --- Function to generate param files ---
# We put this in a list so we can loop through temp1, temp2, temp3
data_sets <- list("temp1" = temp1.str, "temp2" = temp2.str)

lapply(names(data_sets), function(prefix) {
  
  obj <- data_sets[[prefix]]
  pop_list <- seppop(obj)
  
  lapply(names(pop_list), function(pop) {
    
    # 1. Setup paths and data
    pop_obj <- pop_list[[pop]]
    file_name <- paste0(prefix, "_", pop, ".microdrop.txt")
    data_path <- paste0("./microdrop_2/input/", file_name)
    param_path <- paste0("./microdrop_2/params/param_", prefix, "_", pop, ".txt")
    temp_struct <- paste0("./microdrop_2/temp/", prefix, "_", pop, ".struct.txt")
    
    # 2. Generate the Data File (with header)
    genind_to_structure(pop_obj, output = temp_struct)
    temp_data <- read.table(temp_struct, skip = 1)
    
    locus_header <- c("Indiv", "Pop", locNames(pop_obj))
    cat(paste(locus_header, collapse = " "), "\n", file = data_path)
    write.table(temp_data, file = data_path, append = TRUE, sep = " ", 
                row.names = FALSE, col.names = FALSE, quote = FALSE)
    
    # 3. Calculate Parameters for this specific population
    num_inds <- nInd(pop_obj)
    num_loci <- nLoc(pop_obj)
    out_folder <- paste0(prefix, "_", pop, "_results")
    
# 4. Create the Param File content (Note the added ./input/ path)
    param_content <- c(
      "# This is a paramfile for MicroDrop generated automatically",
      "###----Main Parameters----###",
      paste0("DATA_FILE      ", "./input/", file_name), # Pointing MicroDrop to the input folder
      paste0("INDS           ", num_inds),
      paste0("LOCI           ", num_loci),
      "NON_DATA_ROWS  1",
      "NON_DATA_COLS  2",
      paste0("OUT_FOLDER     ", out_folder),
      "TOLERANCE      0.001",
      "MAX_ITER       10000",
      "MAX_ALLELES    100",
      "N_PERMU        100",
      "N_REPS         3",
      "N_IMPT         4",
      "",
      "###----Advanced Parameters----###",
      "ASMPT          0",
      "FCTR           0"
    )
    
    # 5. Write the Param File
    writeLines(param_content, con = param_path)
    
    message("Finished Data and Param files for: ", prefix, "_", pop)
  })
})
```


```{r}
# 1. Find all MLE_Rates_loc files
# Adjust the path to match your structure (looking inside each _results folder)
mle_files <- list.files(path = "./microdrop_2/results", 
                        pattern = "MLE_Rates_loc", 
                        recursive = TRUE, 
                        full.names = TRUE)

# 2. Function to read and label by population
read_mle <- function(file_path) {
  # Extract population name (e.g., "temp1_1") from the folder path
  pop_id <- gsub("./|/MLE_Rates_loc|_results", "", file_path)
  
  # Read the tab-separated file
  # Note: If it's space-separated, remove sep="\t"
  df <- read.table(file_path, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  
  # Add the population column
  df$Population <- pop_id
  return(df)
}

# 3. Combine everything
library(dplyr)
all_mle_rates <- lapply(mle_files, read_mle) %>% bind_rows()

# 4. Quick look at the top dropout loci across all pops
dropout <- all_mle_rates %>% 
  # Sort by highest dropout first
  arrange(desc(dropout_rate)) %>% 
  # Fix the names and extract the temporal group
  mutate(
    Pop = str_replace(Population, "microdropresult", ""),
    Temporal_Group = str_extract(Pop, "temp\\d")
  ) %>% 
  select(-Population)

# 

temp1.dropout <- dropout %>% 
  filter(str_detect(Temporal_Group, "^temp1")) %>% 
  mutate(herd = case_match(Pop,
    "microdrop_resulttemp1_1"  ~ "EINPP",
    "microdrop_resulttemp1_2"  ~ "EINPW",
    "microdrop_resulttemp1_3"  ~ "HZ",
    "microdrop_resulttemp1_4"  ~ "MBS",
    "microdrop_resulttemp1_5"  ~ "RL",
    "microdrop_resulttemp1_6"  ~ "WAB",
    "microdrop_resulttemp1_7"  ~ "WBNPGR",
    "microdrop_resulttemp1_8"  ~ "WBNPHC",
    "microdrop_resulttemp1_9"  ~ "WBNPLB",
    "microdrop_resulttemp1_10" ~ "WBNPNY",
    "microdrop_resulttemp1_11" ~ "WBNPPAD",
    "microdrop_resulttemp1_12" ~ "WENT",
    .default = Pop # Keeps the original name if no match found
  ))

temp1.loc.avg <- temp1.dropout %>%
  group_by(locusID) %>%
  summarize(
    mean_dropout = mean(dropout_rate, na.rm = TRUE),
    sd_dropout = sd(dropout_rate, na.rm = TRUE) # Useful for checking consistency
  ) %>%
  arrange(desc(mean_dropout))


temp2.dropout <- dropout %>% 
  filter(str_detect(Temporal_Group, "^temp2")) %>% 
  mutate(herd = case_when(
    Pop == "microdrop_resulttemp2_1"  ~ "AISH",
    Pop == "microdrop_resulttemp2_2"  ~ "CL",
    Pop == "microdrop_resulttemp2_3"  ~ "EINPP",
    Pop == "microdrop_resulttemp2_4"  ~ "EINPW",
    Pop == "microdrop_resulttemp2_5"  ~ "ETH",
    Pop == "microdrop_resulttemp2_6"  ~ "HZ",
    Pop == "microdrop_resulttemp2_7"  ~ "NAH",
    Pop == "microdrop_resulttemp2_8"  ~ "NORD",
    Pop == "microdrop_resulttemp2_9"  ~ "RL",
    Pop == "microdrop_resulttemp2_10" ~ "WAB",
    Pop == "microdrop_resulttemp2_11" ~ "WBNPPGR",
    Pop == "microdrop_resulttemp2_12" ~ "WBNPHC",
    Pop == "microdrop_resulttemp2_13" ~ "WBNPLB",
    Pop == "microdrop_resulttemp2_14" ~ "WBNPNY",
    Pop == "microdrop_resulttemp2_15" ~ "WBNPPAD",
    Pop == "microdrop_resulttemp2_16" ~ "WENT",
    TRUE ~ Pop
  ))

temp2.loc.avg <- temp2.dropout %>%
  group_by(locusID) %>%
  summarize(
    mean_dropout = mean(dropout_rate, na.rm = TRUE),
    sd_dropout = sd(dropout_rate, na.rm = TRUE) # Useful for checking consistency
  ) %>%
  arrange(desc(mean_dropout))

comb_dropout <- temp1.loc.avg %>% 
  left_join(temp2.loc.avg, by = "locusID") %>% 
  rename(mean_dropout_temp1 = mean_dropout.x,
         sd_temp1 = sd_dropout.x,
         mean_dropout_temp2 = mean_dropout.y,
         sd_temp2 = sd_dropout.y)

# Write out to csv
write.csv(comb_dropout, "./tables/LOC_DROPOUT_RATES.csv")

# Prepare and Truncate Data
locus_plot_data <- comb_dropout %>%
  pivot_longer(
    cols = c(mean_dropout_temp1, mean_dropout_temp2),
    names_to = "Timepoint",
    values_to = "Mean_Dropout"
  ) %>%
  mutate(
    SD = if_else(Timepoint == "mean_dropout_temp1", sd_temp1, sd_temp2),
    lower_error = pmax(0, Mean_Dropout - SD),
    upper_error = Mean_Dropout + SD,
    # SET LEVELS TO REVERSE ORDER: Temp 1 will now be on top
    Timepoint = factor(if_else(Timepoint == "mean_dropout_temp1", "T1", "T2"), 
                       levels = c("T2", "T1"))
  )

p_drop <- ggplot(locus_plot_data, 
       aes(x = fct_rev(locusID), 
           y = Mean_Dropout, 
           fill = Timepoint)) + # Removed fct_rev here
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = lower_error, ymax = upper_error), 
                position = position_dodge(width = 0.8), 
                width = 0.3, alpha = 0.6) +
  coord_flip() + 
  
  scale_fill_manual(
    values = c("T1" = "steelblue", "T2" = "darkorange"),
    # 'breaks' keeps the legend looking like T1 then T2
    breaks = c("T1", "T2") 
  ) +
  
  theme_minimal() +
  labs(x = "Locus",
       y = "Mean Dropout Rate (+SD)",
       fill = "Cohort") +
  theme(
    # 1. Bold and Large Locus Names (Y-axis)
    axis.text.y = element_text(size = 9, face = "bold", color = "black"), 
    
    # 2. Bold and Large Rates (X-axis)
    axis.text.x = element_text(size = 9, face = "bold", color = "black"),
    
    # 3. Bold Axis Titles
    axis.title = element_text(size = 12, face = "bold"),
    
    # 4. Bold Legend Title and Text
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10, face = "bold"),
    
    panel.grid.major.y = element_blank(),
    legend.position = "right"
  )


ggsave(
  filename = "./plots/Locus_Dropout.svg", 
  plot = last_plot(),
  device = "svg",
  width = 10,      
  height = 14,    
  units = "in",
  dpi = 300,     
  bg = "white" 
)
```


# Format final dataset for use in analyses
Removal of herds with insufficient sample sizes for analyses, remove loci which have demonstrated high levels of missing data, and which have a propensity for null alleles. 
```{r}
# Remove loci called for null alleles
analysis.gen <- bison.genind[, loc = !locNames(bison.genind) == "BMS601"] # Remove locus
analysis.gen # Ensure loci were removed (now 48)


# Remove populations with insufficient sample sizes (Temp2)
temp1_remove <- (pop(analysis.gen) %in% c("ETH", "NORD")) & (analysis.gen@other$YEAR.ASSIGN <=2015) # Isolate samples for removal
analysis.gen <- analysis.gen[!temp1_remove, ] # Remove

# Remove populations with insufficient sample sizes (Temp3)
temp2_remove <- (pop(analysis.gen) == "WBNP") & (analysis.gen@other$YEAR.ASSIGN >= 2016)

analysis.gen <- analysis.gen[!temp2_remove, ] # Remove

# Print out final dataset
analysis.gen

# Create filtered dataset for working with in subsequent analyses
analysis.data <- bison.data %>% 
  select(-BMS601, -BM4028) %>% # Remove locus with nulls (BMS601) and missing data (BM4028)
    filter(!(HERD %in% c("ETH", "NORD") & # Insufficient sample sizes
           YEAR.ASSIGN <=2015)) %>% 
    filter(!(HERD == "WBNP" & YEAR.ASSIGN >= 2016)) %>% # No subherd designation
    filter(!(HERD %in% c("WBNPGR", "WBNPHC", "WBNPNY", "WBNPLB", "WBNP") & YEAR.ASSIGN %in% 2000:2015))

  
# Write out to csv
write.csv(analysis.data, "./datafiles/Revised_cleandata_11.03.2026.csv", row.names = F)
```




