---
title: "Locus filtering"
author: "Alex Depew"
date: "2026-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is for data quality control, specificially locus quality control, to ensure accuracy of subsequent analyses.
Files used described below:

- `Cleaned_samples.24.12.2025`
   - Contains all bison microsatellite data from wood and plains bison with added metadata for year collected

## Load Required Packages
```{r warning = FALSE, message = FALSE}
rm(list = ls())
library(adegenet)
library(PopGenReport)
library(poppr)
library(pegas)
library(strataG)
library(tidyverse)
library(graph4lg)
```

## Load Required Data
```{r}
bison.data <- read.csv("./Cleaned_samples.16.02.2026.csv",
                       check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN
```

## Locus Filtering
Preliminary locus filtering relied on *prima facie* information into X-linked loci, followed by the removal of sex markers (not used in analyses). Further tests include:

- Test for polymorphism (do the markers exhibit sufficient variability to be informative?)
- Test for HWE deviations
- Test for linkage disequilibrium
- Null alleles
- Allelic Dropout

## Determine percent of complete genotypes by loci
```{r}
locmiss_bison <-  propTyped(bison.genind, by = "loc")
locmiss_bison[which(locmiss_bison < 0.80)] # print loci with < 80% complete genotypes

barplot(locmiss_bison, ylim = c(0,1), ylab = "Complete genotypes (proportion)", xlab = "Locus", las = 2, cex.names = 0.7)

bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)

```
Locus BM4028 contained more >20% missing data, and therefore was removed from the global dataset.

## Check loci are polymorphic
```{r}
summary(bison.genind)
isPoly(bison.genind) %>% summary
```
All loci are polymorphic, as "TRUE" is returned by the *isPoly* function. The summary of the genind object indicates that the loci contain between 4 and 14 loci. No loci are monomorphic, and therefore none must be removed for being uninformative.

# Data Subsetting
Subsequent genotype quality control measures must be completed within temporal groups, within herds, as to not induce the Wahlund effect. By evaluating deviations from Hardy-Weinberg Equilibrium (HWE), Linkage Disequilibrium (LD), and the presence of null alleles at this granular level, it is possible to distinguish between localized biological variation and systematic technical errors that persist across the entire population.

Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

# Remove herds from temporal group 2 
remove_herds2 <- c("WBNP", "ETH", "NORD", "WENT", "WBNPNY", "WBNPPAD")
temp2 <- temp2[!temp2@pop %in% remove_herds2]
temp2.sum <- summary(temp2) # Recalculate summary
temp2.sum
table(pop(temp2))

# Remove Wentzel from temporal group 3
remove_herds3 <- c("WENT", "WBNP")
temp3 <- temp3[!temp3@pop %in% remove_herds3]
table(pop(temp3))
```

## Test for HWE deviations
Because of the variance in, and small sample sizes seen in some herds, a Markov Chain Monte Carlo exact test can be used to evaluate deviations from HWE.
```{r}
# In order to use the genepop's MCMC method of deterimining HWE deviations, the strataG implementation can be used. the bison "genind" objects must be reformatted as "gtypes" objects
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)
temp3.g <- genind2gtypes(temp3)

# Further, HWE deviations and linkage disequilibrium are tested within each population. Deviations or LD which occur across populations (at the same locus/loci) may indicate technical errors, adaptation, or linkage
pop_list_temp1 <- strataSplit(temp1.g)
pop_list_temp2 <- strataSplit(temp2.g)
pop_list_temp3 <- strataSplit(temp3.g)
```

### Test for HWE
Test for hardy weinberg equilibrium intra-population
```{r}
# Apply HWE test from genepop. The strataG wrapper formats this back into R nicely for subsequent manipulation

# Temp1 ----
hwe_results_list_temp1 <- lapply(pop_list_temp1, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp1 <- as.data.frame(hwe_results_list_temp1) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 

# Temp2 ----
hwe_results_list_temp2 <- lapply(pop_list_temp2, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp2 <- as.data.frame(hwe_results_list_temp2) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 


# Temp3 ----
hwe_results_list_temp3 <- lapply(pop_list_temp3, function(pop) {
  hweTest(
    pop, 
    use.genepop = T,     
    which = "Proba",         
    dememorization = 10000,  
    batches = 500,           
    num.rep = 5000,
    delete.files = TRUE
  )
})

HWE.temp3 <- as.data.frame(hwe_results_list_temp3) %>% # Make results as df
  rownames_to_column("Locus") %>%  
  pivot_longer(
    cols = -c(Locus), 
    names_to = "Population",       
    values_to = "HWE_p_value"      
  ) %>% 
  # p-value correction by Population
  group_by(Population) %>% 
  mutate(p.adj_herd = p.adjust(HWE_p_value, method = "BH")) %>% 
  ungroup() 
```

### Summarize HWE Deviations
```{r}
# Temp1 ----
temp1.HWE.sum <- HWE.temp1 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp1.HWE.sum)

# Temp2 ----
temp2.HWE.sum <- HWE.temp2 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp2.HWE.sum)

# Temp3 ----
temp3.HWE.sum <- HWE.temp3 %>%
  group_by(Population) %>%
  summarise(
    HWE_Deviations = sum(p.adj_herd <= 0.05, na.rm = TRUE),
  ) %>%
  arrange(desc(HWE_Deviations)) # Sort with most deviations at the top

# View the summary table
View(temp3.HWE.sum)

```


## Test for Linkage Disequilibrium
Pairwise test for LD using Genepop through *strataG*. If LD is present in loci across the majority of herds, it may indicate physical linkage as opposed to local variation.
```{r}
# Temp1 ----
LD_results_temp1 <- lapply(pop_list_temp1, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp1 <- bind_rows(LD_results_temp1, .id = "Population") # Put into dataframe format

LD.df.temp1 <- LD.df.temp1 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp1.sig <- LD.df.temp1 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp1.sig, "./tables/temp1LD.csv", row.names = F)


# Temp2 ----
LD_results_temp2 <- lapply(pop_list_temp2, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp2 <- bind_rows(LD_results_temp2, .id = "Population") # Put into dataframe format

LD.df.temp2 <- LD.df.temp2 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup()

# Dataframe of significant pairwise values
LD.temp2.sig <- LD.df.temp2 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp2.sig, "./tables/temp2LD.csv", row.names = F)


# Temp3 ----
LD_results_temp3 <- lapply(pop_list_temp3, function(pop) {
  LDgenepop(
  pop,
  dememorization = 10000,
  batches = 100,
  iterations = 5000,
  delete.files = TRUE
  )
})

LD.df.temp3 <- bind_rows(LD_results_temp3, .id = "Population") # Put into dataframe format

LD.df.temp3 <- LD.df.temp3 %>%
  group_by(Population) %>%  # Separate by populations
  mutate(p.adj_herd = p.adjust(p.value, method = "BH")) %>% # Perform BH correction within each population
  ungroup() 

# Dataframe of significant pairwise values
LD.temp3.sig <- LD.df.temp3 %>% 
  filter(p.adj_herd <= 0.05 & 
           !is.na(Locus.2))

write.csv(LD.temp3.sig, "./tables/temp3LD.csv", row.names = F)
```
The dataframes containing only significant pairwise comparisons following the Benjamini-Hochberg correction return significant p-values in populations which are represented by relatively small sample sizes (n = 8). Instead of indicating true linkage disequilibrium, the lack of intra-population allele data likely inflates the variance in allele frequency estimates, increasing the probability of detecting spurious non-random locus associations due to sampling effects. Significance for non-random association was detected in EINPP and/or EINPW in all temporal groups, yet the loci purported to have significant association were isolated to either EINPP or EINPW (potential local variation/fixation), and were not concordant across temporal groups. 


## Test for null alleles and allelic dropout
To test for null alleles and allelic dropout, a combination of diagnostic and correction methods can be used. Microchecker is an external program which can test for the presence of null alleles using heterozygote deficiency methods as described by Chakraborty (1994) and Brookfield (1996). It also assess the presence of short allele dominance by comparing the proportion of short to long repeats. In order to use this program, datasets must be reformatted into genepop format using the package *graph4lg*.
```{r}
genind_to_genepop(temp1,
                  output = "genepop_files/temp1.genepop.txt")

genind_to_genepop(temp2,
                  output = "genepop_files/temp2.genepop.txt")

genind_to_genepop(temp3,
                  output = "genepop_files/temp3.genepop.txt")
```

### Obtain null allele frequency
As the percentage of missing data is small, the Brookfield method may be employed to estimate null allele frequency.
```{r}
# Locus Labels
loc_lab <- c(
  "BMS1117", "TGLA53", "BMC4214", "BM1225", "BM47", 
  "BMS1315", "IL4", "BM2113", "BMS1716", "BM1706", 
  "BMS1675", "BMS1857", "BMS1001", "BMS601", "BM720", 
  "ILSTS102", "BMS1074", "BMS1172", "BMS2258", "BM757", 
  "SPS115", "BMS527", "BM2830", "INRA194", "BMS941", 
  "INRA037", "BM4107", "TGLA122", "HUJ246", "BMS528", 
  "BM4440", "TGLA44", "BMS510", "BMS1355", "BM1824", 
  "BMS812", "BMS2639", "BMS410", "AGLA232", "BM17132", 
  "BM711", "INRA133", "RM372", "BM4311", "BMS1747", 
  "BL1036", "BM1862", "BMS1862", "URB011"
)

# Match locus to numeric labels in null file
loc_match <- tibble(
  Locus = 1:49, 
  Locus_Name = loc_lab
)

Pop_lab <- c("EINPP", "EINPW", "MBS", "WBNPGR", "WBNPHC", 
             "WBNPLB", "WBNPNY", "WBNPPAD")
# Match population to numeric labels in null file
pop_match <- tibble(
  Pop = 1:8,
  Pop_Name = Pop_lab
)

temp1.nulls <- read.table("./genepop_files/FreeNA_out/temp1.Null_freq.txt",
                          col.names = c("Locus", "Pop", "Null_Freq"))

temp1.nulls <- left_join(temp1.nulls, loc_match, by = "Locus") %>% 
  select(-Locus) %>% 
  left_join(pop_match, temp1.nulls, by = "Pop") %>% 
  select(-Pop) %>% 
  relocate(Null_Freq, .after = Pop_Name) %>% 
  relocate(Locus_Name, .after = Pop_Name)
```

## Test for allelic dropout
```{r}
# Microdrop requires STRUCTURE input files
genind_to_structure(temp1,
                    output = "./structure/temp1.structure.txt")

```


