---
title: "Effective Population Size"
author: "Alex Depew"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to calculate effective population size (Ne) in all study herds. 

Files used described below:

- `Cleaned_locus_data_20.02.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Cleaned_locus_data_20.02.2026.csv",
                         check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset Data
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
```

## Input File Creation
NeEstimator input files are in *genepop* format. 
```{r}
genind_to_genepop(temp1,
                  output = "./NeEstimator/temp1.ne.input.txt")

genind_to_genepop(temp2,
                  output = "./NeEstimator/temp2.ne.input.txt")

genind_to_genepop(temp3,
                  output = "./NeEstimator/temp3.ne.input.txt")
```

## Read Files
```{r}
library(tidyverse)
# Header
NE_header <- c("Population", "n", "Weighted.Mean", "Ind.Alleles","r^2", "Exp(r^2)", "Ne", "par_Low", "par_High", "jack_Low", "jack_High", "Eff.df")

# First Temporal Group
temp1_Ne_0.05 <- read.table("./NeEstimator/temp1_NE_0.05",
                            sep = "\t",
                            col.names = NE_header) %>% 
                 mutate(jack_Low = as.character(jack_Low)) %>%
                 mutate(Population = str_remove_all(Population, "^\\d+:|_\\d+\\s*$")) %>% # Removes clutter from the Population column names
                 select(-c(par_Low, par_High)) # Removes parametric CIs

# Second Temporal Group
temp2_Ne_0.05 <- read.table("./NeEstimator/temp2_NE_0.05",
                            sep = "\t",
                            col.names = NE_header) %>% 
                 mutate(jack_Low = as.character(jack_Low)) %>%
                 mutate(Population = str_remove_all(Population, "^\\d+:|_\\d+\\s*$")) %>% 
                 select(-c(par_Low, par_High))

# Third Temporal Group
temp3_Ne_0.05 <- read.table("./NeEstimator/temp3_NE_0.05",
                            sep = "\t",
                            col.names = NE_header) %>% 
                 mutate(jack_Low = as.character(jack_Low)) %>%
                 mutate(Population = str_remove_all(Population, "^\\d+:|_\\d+\\s*$")) %>% 
                 select(-c(par_Low, par_High))


# Combine into one Ne dataframe
Ne_table <- temp1_Ne_0.05 %>% 
  bind_rows(temp2_Ne_0.05, temp3_Ne_0.05)

# Write to csv
write.csv(Ne_table, "./tables/Ne_estimation.csv", row.names = F)
```


