---
title: "PCA_STRUCT"
author: "Alex Depew"
date: "2026-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(dplyr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.16.02.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

# Remove herds from temporal group 2 and 3
remove_herds <- c("WBNPPAD", "WBNPNY", "ETH", "NORD", "WENT")
temp2 <- temp2[!pop(temp2) %in% remove_herds, ]
temp2.sum <- summary(temp2) # Recalculate summary

temp3 <- temp3[!pop(temp3) %in% "WBNP"]
temp3.sum <- summary(temp3) # Recalculate summary
```

Remove Elk Island Plains for Structure Analyses
```{r}
temp1 <- temp1[temp1@pop != "EINPP"]
temp1.sum <- summary(temp1)

temp2 <- temp2[temp2@pop != "EINPP"]
temp2.sum <- summary(temp2)

temp3 <- temp3[temp3@pop != "EINPP"]
temp3.sum <- summary(temp3)
```


# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 
```{r}
custom_palette <- c(
  "#E31A1C", "#1F78B4", "#FFFF99", "#33A02C", "#FF7F00", 
  "#6A3D9A", "#B15928", "#A6CEE3", "#B2DF8A", "#FB9A99", 
  "#FDBF6F", "#CAB2D6", "#000000", "#8DD3C7", "#BEBADA",
  "#2F4F4F", "#FF1493")

all_herds <- c(
  "AISH", "CL", "EINPP", "EINPW", "ETH", "HZ", "MBS", "NAH", "NORD", 
  "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

herd_col_set <- setNames(custom_palette, all_herds)

```


### Temporal Group 1
```{r}
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp1 <- 100*pca.temp1$eig/sum(pca.temp1$eig)

# Plot
pdf(file = "./plots/pca_temp1.pdf", width = 10, height = 10)
cols_temp1 <- herd_col_set[levels(pop(temp1))]

s.class(pca.temp1$li, fac = pop(temp1), col = transp(cols_temp1, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp1[1], 1), "% | PC2: ", round(eig.perc.temp1[2], 1), "%"),)

add.scatter.eig(pca.temp1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.temp1$c1^2)
```



### Temporal Group 2
```{r}
x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp2 <- 100*pca.temp2$eig/sum(pca.temp2$eig)

# Plot
pdf(file = "./plots/pca_temp2.pdf", width = 10, height = 10)
cols_temp2 <- herd_col_set[levels(pop(temp2))] # Map colours to herd

s.class(pca.temp2$li, fac = pop(temp2), col = transp(cols_temp2, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp2[1], 1), "% | PC2: ", round(eig.perc.temp2[2], 1), "%"),)

add.scatter.eig(pca.temp2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()


# Allele contributions to variance
loadingplot(pca.temp2$c1^2)

```

### Temporal Group 3
```{r}
x.temp3 <- tab(temp3, freq = TRUE, NA.method = "mean")
pca.temp3 <- dudi.pca(x.temp3, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp3 <- 100*pca.temp3$eig/sum(pca.temp3$eig)

# Plot
pdf(file = "./plots/pca_temp3.pdf", width = 10, height = 10)
cols_temp3 <- herd_col_set[levels(pop(temp3))]


s.class(pca.temp3$li, fac = pop(temp3), col = transp(cols_temp3, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp3[1], 1), "% | PC2: ", round(eig.perc.temp3[2], 1), "%"),)

add.scatter.eig(pca.temp3$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.temp3$c1^2)
```


# STRUCTURE Analyses
```{r}
# Populations need to be in unique integer format for input into STRUCTURE:
temp1.struct.gen <- temp1
pop(temp1.struct.gen) <- as.numeric(as.factor(pop(temp1.struct.gen)))
summary(temp1.struct.gen)


temp2.struct.gen <- temp2
pop(temp2.struct.gen) <- as.numeric(as.factor(pop(temp2.struct.gen)))
summary(temp2.struct.gen)


temp3.struct.gen <- temp3
pop(temp3.struct.gen) <- as.numeric(as.factor(pop(temp3.struct.gen)))
summary(temp3.struct.gen)

# Format in STRUCTURE format
library(graph4lg)
genind_to_structure(temp1.struct.gen, output = "structure/temp1.structure.txt")
genind_to_structure(temp2.struct.gen, output = "structure/temp2.structure.txt")
genind_to_structure(temp3.struct.gen, output = "structure/temp3.structure.txt")
```


# STRUCTURE VISUALIZATION

## Subset csv
The data csv file can be subset to match the samples included in each STRUCTURE run. This will be used to match the sample ID's for plotting on the STRUCTURE plot.
```{r}
temp1.data <- bison.data %>%
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>%
  rename(ID = Spreadsheet_ID) %>%     
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

temp2.data <- bison.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2000 &
           YEAR.ASSIGN <=2015) %>%
  filter(!HERD %in% remove_herds) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

  
temp3.data <- temp2.data <- bison.data %>% 
  select(-ID) %>%
  filter(YEAR.ASSIGN >=2016) %>%
  filter(HERD != "WBNP") %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

```

#### Select Colour Palette for STRUCTURE Plotting
```{r}
clist <- list(
"shiny"=c("#1D72F5","#DF0101","#77CE61", "#FF9326","#A945FF","#0089B2","#FDF060","#FFA6B2","#BFF217","#60D5FD","#CC1577","#F2B950","#7FB21D","#EC496F","#326397","#B26314","#027368","#A4A4A4","#610B5E"),
"strong"=c("#11A4C8","#63C2C5","#1D4F9F","#0C516D","#2A2771","#396D35","#80C342","#725DA8","#B62025","#ED2224","#ED1943","#ED3995","#7E277C","#F7EC16","#F8941E","#8C2A1C","#808080"),
"oceanfive"=c("#00A0B0", "#6A4A3C", "#CC333F", "#EB6841", "#EDC951"),
"keeled"=c("#48B098", "#91CB62", "#FFEE3B", "#FB9013", "#FF3C28"),
"vintage"=c("#400F13", "#027368", "#A3BF3F", "#F2B950", "#D93A2B"),
"muted"=c("#46BDDD","#82DDCE","#F5F06A","#F5CC6A","#F57E6A"),
"teal"=c("#CFF09E","#A8DBA8","#79BD9A","#3B8686","#0B486B"),
"merry"=c("#5BC0EB","#FDE74C","#9BC53D","#E55934","#FA7921"),
"funky"=c("#A6CEE3", "#3F8EAA", "#79C360", "#E52829", "#FDB762","#ED8F47","#9471B4"),
"retro"=c("#01948E","#A9C4E2","#E23560","#01A7B3","#FDA963","#323665","#EC687D"),
"cb_paired"=c("#A6CEE3","#1F78B4","#B2DF8A","#33A02C","#FB9A99","#E31A1C","#FDBF6F","#FF7F00","#CAB2D6","#6A3D9A","#FFFF99","#B15928"),
"cb_set3"=c("#8DD3C7","#FFFFB3","#BEBADA","#FB8072","#80B1D3","#FDB462","#B3DE69","#FCCDE5","#D9D9D9","#BC80BD","#CCEBC5","#FFED6F"),
"morris"=c("#4D94CC","#34648A","#8B658A","#9ACD32","#CC95CC","#9ACD32","#8B3A39","#CD6601","#CC5C5B","#8A4500"),
"wong"=c("#000000","#E69F00","#56B4E9","#009E73","#F0E442","#006699","#D55E00","#CC79A7"),
"krzywinski"=c("#006E82","#8214A0","#005AC8","#00A0FA","#FA78FA","#14D2DC","#AA0A3C","#FA7850","#0AB45A","#F0F032","#A0FA82","#FAE6BE"))
```


```{r}
library(tidyverse)
library(pophelper)
library(gridExtra)
```

### Temporal Group 1
Load Data
```{r}
sfiles_temp1 <- list.files("./structure/STRUCTURE_RESULTS/temp1/", full.names = TRUE)
slist_temp1 <- readQ(files = sfiles_temp1, filetype = "structure")
```

Optimal K clusters
```{r}
struct_sum_temp1 <- summariseQ(tabulateQ(slist_temp1))

p1 <- evannoMethodStructure(data = struct_sum_temp1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "evannoMethodStructure_Temp1",
                           exportpath = "./plots/structure_plots/") 

grid.arrange(p1)
```
The methods described by Pritchard (2000) and Evanno (2006) both indicate the most likely number of populations is 3.

### Plotting
```{r, results='hide'}
ind_label_temp1 <- readQ(files = sfiles_temp1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp1 <- rownames_to_column(ind_label_temp1, "ID")

locations_temp1 <- left_join(ind_label_temp1, temp1.data, by = "ID")

## 
slist <- slist_temp1

temp1_ind_order <- locations_temp1
rownames(slist[[1]]) <- temp1_ind_order$ID
if(length(unique(sapply(slist,nrow)))==1) slist <- lapply(slist,"rownames<-",temp1_ind_order$ID)
lapply(slist, rownames)[1:70]

slist_temp1 <- slist 
slist_align1 <- alignK(slist_temp1, "within") # Align replicate runs
slist_align1 <- slist_align1[11:70]
class(slist_align1) <- "qlist"
merged_temp1 <- mergeQ(slist_align1)

loc_labels_temp1 <- locations_temp1$HERD %>% as.data.frame()
colnames(loc_labels_temp1) <- "Region"

herd_order_temp1 <- c("EINPW", "MBS", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")


plotQ(merged_temp1, 
      imgoutput = "join",
            basesize = 10, 
            clustercol = colours, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "Temp1_K3",
            exportpath = "./plots/structure_plots/")

```




```{r}
# Print summary of runs (Evanno method plot is based on these values)
summariseQ(tabulateQ(slist_temp1))

# Align and merge across K
m_k3 <- mergeQ(alignK(slist_temp1[21:30])) # This aggregates all 10 replicates of K = 3

# Structure Plot
plotQ(,
      imgoutput = "join",
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "Temp1",
            exportpath = "./plots/structure_plots/")

```


### Temporal Group 2
Load Data
```{r}
sfiles_temp2 <- list.files("./structure/STRUCTURE_RESULTS/temp2", full.names = TRUE)
slist_temp2 <- readQ(files = sfiles_temp2, filetype = "structure")
```

Optimal K clusters
```{r}
struct_sum_temp2 <- summariseQ(tabulateQ(slist_temp2))

p2 <- evannoMethodStructure(data = struct_sum_temp2, exportplot = F,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7)
grid.arrange(p2)
```

Plotting
```{r}
# Organize into individual and population order
ind_label_temp2 <- readQ(files = sfiles_temp2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp2 <- rownames_to_column(ind_label_temp2, "ID")

locations_temp2 <- left_join(ind_label_temp2, temp2.data, by = "ID")

slist <- slist_temp2

temp2_ind_order <- locations_temp2
rownames(slist[[1]]) <- temp2_ind_order$ID
if(length(unique(sapply(slist,nrow)))==1) slist <- lapply(slist,"rownames<-",temp2_ind_order$ID)
lapply(slist, rownames)[0:50]

slist_temp2 <- slist
slist_align2 <- alignK(slist_temp2, "within")
merged_temp2 <- mergeQ(slist_align2[41:50]) # Merge K = 2 runs


loc_labels_temp2 <- locations_temp2$HERD %>% as.data.frame()
colnames(loc_labels_temp2) <- "Region"

herd_order_temp2 <- c("EINPW", "HZ", "MBS", "RL", "WAB")

# Structure Plot
plotQ(slist_align2[45], 
            basesize = 10,
            clustercol = colours, 
            grplab = loc_labels_temp2, 
            subsetgrp = herd_order_temp2, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 3, 
            grplabangle = 0,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "Temp2_K5",
            exportpath = "./plots/structure_plots/")
```

### Temporal Group 3
Load Data
```{r}
sfiles_temp3 <- list.files("./structure/results_temp3_f/", full.names = TRUE)
slist_temp3 <- readQ(files = sfiles_temp3, filetype = "structure")
```


Optimal K clusters
```{r}
struct_sum_temp3 <- summariseQ(tabulateQ(slist_temp3))

p3 <- evannoMethodStructure(data = struct_sum_temp3, 
                            exportplot = F,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                          )
grid.arrange(p3)
```


```{r}
# Organize into individual and population order
ind_label_temp3 <- readQ(files = sfiles_temp3[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp3 <- rownames_to_column(ind_label_temp3, "ID")

locations_temp3 <- left_join(ind_label_temp3, temp3.data, by = "ID")

slist <- slist_temp3

temp3_ind_order <- locations_temp3
rownames(slist[[1]]) <- temp3_ind_order$ID
if(length(unique(sapply(slist,nrow)))==1) slist <- lapply(slist,"rownames<-",temp3_ind_order$ID)
lapply(slist, rownames)[0:150]

slist_temp3 <- slist
slist_align3 <- alignK(slist_temp3, "within")
merged_temp3 <- mergeQ(slist_align3)

loc_labels_temp3 <- locations_temp3$HERD %>% as.data.frame()
colnames(loc_labels_temp3) <- "Region"

herd_order_temp3 <- c("AISH", "CL", "EINPW", "ETH", "HZ", "NAH", 
                      "NORD", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", 
                      "WBNPNY", "WBNPPAD", "WENT")


# Structure Plot
plotQ(slist_align3[75], 
            basesize = 10, 
            clustercol = colours, 
            grplab = loc_labels_temp3, 
            subsetgrp = herd_order_temp3, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 3, 
            grplabangle = 0,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            exportpath = "./plots/structure_plots/")
```