---
title: "PCA_STRUCT"
author: "Alex Depew"
date: "2026-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("Cleaned_locus_data_20.02.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)
```

Remove Elk Island Plains for Structure Analyses
```{r}
temp1 <- temp1[temp1@pop != "EINPP"]
temp1.sum <- summary(temp1)

temp2 <- temp2[temp2@pop != "EINPP"]
temp2.sum <- summary(temp2)

temp3 <- temp3[temp3@pop != "EINPP"]
temp3.sum <- summary(temp3)
```


# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 
```{r}
custom_palette <- c(
  "#E31A1C", # AISH - Bold Red
  "#6A3D9A", # CL - Deep Purple
  "#FF7F00", # EINPP - Bright Orange
  "#A6CEE3", # EINPW - Sky Blue
  "#B15928", # ETH - Burnt Sienna
  "#FB9A99", # HZ - Soft Pink
  "#FDBF6F", # MBS - Light Orange
  "#CAB2D6", # NAH - Lavender
  "#FFFF99", # NORD - Pale Yellow
  "#1F78B4", # RL - Steel Blue
  "#E377C2", # WAB - Magenta
  # --- WBNP SUB-HERD BOLD SET (Replaced Lime) ---
  "#006400", # WBNPGR (Garden River) - Dark Forest Green
  "#FFD700", # WBNPHC (Hay Camp) - Bold Gold (High Visibility)
  "#00CED1", # WBNPLB (Little Buffalo) - Dark Turquoise
  "#FF4500", # WBNPNY (Nyarrowu) - Vermillion (Red-Orange)
  "#4169E1", # WBNPPAD (Delta) - Royal Blue
  # ----------------------------------------------
  "#333333"  # WENT - Graphite
)

all_herds <- c(
  "AISH", "CL", "EINPP", "EINPW", "ETH", "HZ", "MBS", "NAH", "NORD", 
  "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

herd_col_set <- setNames(custom_palette, all_herds)

```


### Temporal Group 1
```{r}
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp1 <- 100*pca.temp1$eig/sum(pca.temp1$eig)

# Plot
pdf(file = "./plots/pca_temp1.pdf", width = 10, height = 10)
cols_temp1 <- herd_col_set[levels(pop(temp1))]

s.class(pca.temp1$li, fac = pop(temp1), col = transp(cols_temp1, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp1[1], 1), "% | PC2: ", round(eig.perc.temp1[2], 1), "%"),)

add.scatter.eig(pca.temp1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.temp1$c1^2)
```



### Temporal Group 2
```{r}
x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp2 <- 100*pca.temp2$eig/sum(pca.temp2$eig)

# Plot
pdf(file = "./plots/pca_temp2.pdf", width = 10, height = 10)
cols_temp2 <- herd_col_set[levels(pop(temp2))] # Map colours to herd

s.class(pca.temp2$li, fac = pop(temp2), col = transp(cols_temp2, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp2[1], 1), "% | PC2: ", round(eig.perc.temp2[2], 1), "%"),)

add.scatter.eig(pca.temp2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()


# Allele contributions to variance
loadingplot(pca.temp2$c1^2)

```

### Temporal Group 3
```{r}
x.temp3 <- tab(temp3, freq = TRUE, NA.method = "mean")
pca.temp3 <- dudi.pca(x.temp3, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp3 <- 100*pca.temp3$eig/sum(pca.temp3$eig)

# Plot
pdf(file = "./plots/pca_temp3.pdf", width = 10, height = 10)
cols_temp3 <- herd_col_set[levels(pop(temp3))]


s.class(pca.temp3$li, fac = pop(temp3), col = transp(cols_temp3, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.temp3[1], 1), "% | PC2: ", round(eig.perc.temp3[2], 1), "%"),)

add.scatter.eig(pca.temp3$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.temp3$c1^2)
```
### Stack Plots
```{r}
pdf(file = "./plots/wb_pca_stacked.pdf", width = 30, height = 10) # Increase height for 3 plots

# Set up  grid 3 rows, 1 column
par(mfrow = c(1, 3)) 

# Plot 1
s.class(pca.temp1$li, fac = pop(temp1), col = transp(cols_temp1, 0.75),
        axesell = F, cstar = 0, cpoint = 3, grid = F, clabel = 2, csub = 4,
        sub = paste0("PC1: ", round(eig.perc.temp1[1], 1), "% | PC2: ", round(eig.perc.temp1[2], 1), "%")
)
mtext("a)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)


# Plot 2
s.class(pca.temp2$li, fac = pop(temp2), col = transp(cols_temp2, 0.75),
        axesell = F, cstar = 0, cpoint = 3, grid = F, clabel = 2, csub = 4,
        sub = paste0("PC1: ", round(eig.perc.temp2[1], 1), "% | PC2: ", round(eig.perc.temp2[2], 1), "%")
)
mtext("b)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)


# Plot 3
s.class(pca.temp3$li, fac = pop(temp3), col = transp(cols_temp3, 0.75),
        axesell = F, cstar = 0, cpoint = 3, grid = F, clabel = 2, csub = 4,
        sub = paste0("PC1: ", round(eig.perc.temp3[1], 1), "% | PC2: ", round(eig.perc.temp3[2], 1), "%")
)
mtext("c)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)

dev.off()
```
# Subset GENIND
```{r}
# WBNP-ONLY samples
wbnp1 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") & 
                        bison.genind@other$YEAR.ASSIGN <= 1999]


wbnp2 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") & 
                        bison.genind@other$YEAR.ASSIGN >= 2016]


admix1 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", 
                                               "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") & 
                      bison.genind@other$YEAR.ASSIGN <= 1999]


admix2 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", 
                                               "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") & 
                      bison.genind@other$YEAR.ASSIGN >= 2016]

```


## WBNP PCAs
```{r}
## WBNP 1 ----
x.wbnp1 <- tab(wbnp1, freq = TRUE, NA.method = "mean")
pca.wbnp1 <- dudi.pca(x.wbnp1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.wbnp1 <- 100*pca.wbnp1$eig/sum(pca.wbnp1$eig)

# Plot
pdf(file = "./plots/pca_wbnp1.pdf", width = 10, height = 10)
cols_wbnp1 <- herd_col_set[levels(pop(wbnp1))] # Map colours to herd

s.class(pca.wbnp1$li, fac = pop(wbnp1), col = transp(cols_wbnp1, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.wbnp1[1], 1), "% | PC2: ", round(eig.perc.wbnp1[2], 1), "%"),)

add.scatter.eig(pca.wbnp1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()


# Allele contributions to variance
loadingplot(pca.wbnp1$c1^2)

## WBNP 2 ----

x.wbnp2 <- tab(wbnp2, freq = TRUE, NA.method = "mean")
pca.wbnp2 <- dudi.pca(x.wbnp2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.wbnp2 <- 100*pca.wbnp2$eig/sum(pca.wbnp2$eig)

# Plot
pdf(file = "./plots/pca_wbnp2.pdf", width = 10, height = 10)
cols_wbnp2 <- herd_col_set[levels(pop(wbnp2))] # Map colours to herd

s.class(pca.wbnp2$li, fac = pop(wbnp2), col = transp(cols_wbnp2, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.wbnp2[1], 1), "% | PC2: ", round(eig.perc.wbnp2[2], 1), "%"),)

add.scatter.eig(pca.wbnp2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()


# Allele contributions to variance
loadingplot(pca.wbnp2$c1^2)


## Admix 1 ----

x.admix1 <- tab(admix1, freq = TRUE, NA.method = "mean")
pca.admix1 <- dudi.pca(x.admix1, center = T, scale = F, scannf = F, nf = 2)

pca.admix1$li[,1] <- pca.admix1$li[,1] * -1  # Flip PC1 (Left-Right)


# Axes % Variance
eig.perc.admix1 <- 100*pca.admix1$eig/sum(pca.admix1$eig)

# Plot
pdf(file = "./plots/pca_admix1.pdf", width = 10, height = 10)
cols_admix1 <- herd_col_set[levels(pop(admix1))] # Map colours to herd

s.class(pca.admix1$li, fac = pop(admix1), col = transp(cols_admix1, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.admix1[1], 1), "% | PC2: ", round(eig.perc.admix1[2], 1), "%"),)

add.scatter.eig(pca.admix1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()


# Allele contributions to variance
loadingplot(pca.admix1$c1^2)

## Admix 2 ----

x.admix2 <- tab(admix2, freq = TRUE, NA.method = "mean")
pca.admix2 <- dudi.pca(x.admix2, center = T, scale = F, scannf = F, nf = 2)



# Axes % Variance
eig.perc.admix2 <- 100*pca.admix2$eig/sum(pca.admix2$eig)

# Plot
pdf(file = "./plots/pca_admix2.pdf", width = 10, height = 10)
cols_admix2 <- herd_col_set[levels(pop(admix2))] # Map colours to herd

s.class(pca.admix2$li, fac = pop(admix2), col = transp(cols_admix2, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.admix2[1], 1), "% | PC2: ", round(eig.perc.admix2[2], 1), "%"),)

add.scatter.eig(pca.admix2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()

# Allele contributions to variance
loadingplot(pca.admix2$c1^2)
```


#### Stack WBNP PCA Plots
```{r}
pdf(file = "./plots/wbnp_pca_stacked.pdf", width = 15, height = 15) 

# Set up  grid 3 rows, 1 column
par(mfrow = c(2, 2)) 

# Plot 1
s.class(pca.wbnp1$li, fac = pop(wbnp1), col = transp(cols_wbnp1, 0.75),
        axesell = F, cstar = 0, cpoint = 2, grid = F, csub = 2,
        sub = paste0("PC1: ", round(eig.perc.wbnp1[1], 1), "% | PC2: ", round(eig.perc.wbnp1[2], 1), "%"))
mtext("a)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)

# Plot 2
s.class(pca.wbnp2$li, fac = pop(wbnp2), col = transp(cols_wbnp2, 0.75),
        axesell = F, cstar = 0, cpoint = 2, grid = F, csub = 2,
        sub = paste0("PC1: ", round(eig.perc.wbnp2[1], 1), "% | PC2: ", round(eig.perc.wbnp2[2], 1), "%"))
mtext("b)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)

# Plot 3
s.class(pca.admix1$li, fac = pop(admix1), col = transp(cols_admix1, 0.75),
        axesell = F, cstar = 0, cpoint = 2, grid = F, csub = 2,
        sub = paste0("PC1: ", round(eig.perc.admix1[1], 1), "% | PC2: ", round(eig.perc.admix1[2], 1), "%"))
mtext("c)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)

# Plot 4
s.class(pca.admix2$li, fac = pop(admix2), col = transp(cols_admix2, 0.75),
        axesell = F, cstar = 0, cpoint = 2, grid = F, csub = 2,
        sub = paste0("PC1: ", round(eig.perc.admix2[1], 1), "% | PC2: ", round(eig.perc.admix2[2], 1), "%"))
mtext("d)", side = 3, adj = 0.01, line = 1, cex = 2, font = 2)

dev.off()
```


# STRUCTURE Analyses
```{r}
# Populations need to be in unique integer format for input into STRUCTURE:
temp1.struct.gen <- temp1
pop(temp1.struct.gen) <- as.numeric(as.factor(pop(temp1.struct.gen)))
summary(temp1.struct.gen)


temp2.struct.gen <- temp2
pop(temp2.struct.gen) <- as.numeric(as.factor(pop(temp2.struct.gen)))
summary(temp2.struct.gen)


temp3.struct.gen <- temp3
pop(temp3.struct.gen) <- as.numeric(as.factor(pop(temp3.struct.gen)))
summary(temp3.struct.gen)

# Format in STRUCTURE format
library(graph4lg)
genind_to_structure(temp1.struct.gen, output = "structure/temp1.structure.txt")
genind_to_structure(temp2.struct.gen, output = "structure/temp2.structure.txt")
genind_to_structure(temp3.struct.gen, output = "structure/temp3.structure.txt")

# WBNP-ONLY samples
wbnp1.struct.gen <- wbnp1
pop(wbnp1.struct.gen) <- as.numeric(as.factor(pop(wbnp1.struct.gen)))
summary(wbnp1.struct.gen)

wbnp2.struct.gen <- wbnp2
pop(wbnp2.struct.gen) <- as.numeric(as.factor(pop(wbnp2.struct.gen)))
summary(wbnp2.struct.gen)

genind_to_structure(wbnp1.struct.gen, output = "structure/wbnp1.structure.txt")
genind_to_structure(wbnp2.struct.gen, output = "structure/wbnp2.structure.txt")

# WBNP with EINPW and EINPP
admix1.struct.gen <- admix1
pop(admix1.struct.gen) <- as.numeric(as.factor(pop(admix1.struct.gen)))
summary(admix1.struct.gen)

admix2.struct.gen <- admix2
pop(admix2.struct.gen) <- as.numeric(as.factor(pop(admix2.struct.gen)))
summary(admix2.struct.gen)

genind_to_structure(admix1.struct.gen, output = "structure/admix1.structure.txt")
genind_to_structure(admix2.struct.gen, output = "structure/admix2.structure.txt")
```


# STRUCTURE VISUALIZATION

## Subset csv
The data csv file can be subset to match the samples included in each STRUCTURE run. This will be used to match the sample ID's for plotting on the STRUCTURE plot.
```{r}
temp1.data <- cleaned.data %>%
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>%
  rename(ID = Spreadsheet_ID) %>%     
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

temp2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2000 &
           YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

  
temp3.data <- cleaned.data %>% 
  select(-ID) %>%
  filter(YEAR.ASSIGN >=2016) %>%
  filter(HERD != "WBNP") %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD != "EINPP")

wbnp1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")) %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) 

wbnp2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >= 2016) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")) %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) 

admix1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>% 
  filter(HERD %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")) %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) 

admix2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>% 
  filter(HERD %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")) %>% 
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) 
```

### Temporal Group 1

#### Load Data
```{r}
# List files
# Full datasets
sfiles_temp1 <- list.files("./structure/STRUCTURE_RESULTS/temp1/", full.names = TRUE)
sfiles_temp2 <- list.files("./structure/STRUCTURE_RESULTS/temp2/", full.names = TRUE)
sfiles_temp3 <- list.files("./structure/STRUCTURE_RESULTS/temp3/", full.names = TRUE)

# WBNP Subsets
sfiles_wbnp1 <- list.files("./structure/STRUCTURE_RESULTS/wbnp1/", full.names = TRUE)
sfiles_wbnp2 <- list.files("./structure/STRUCTURE_RESULTS/wbnp2/", full.names = TRUE)

sfiles_admix1 <- list.files("./structure/STRUCTURE_RESULTS/admix1/", full.names = TRUE)
sfiles_admix2 <- list.files("./structure/STRUCTURE_RESULTS/admix2/", full.names = TRUE)

# Qlist (pophelper)
slist_temp1 <- readQ(files = sfiles_temp1, filetype = "structure")
slist_temp2 <- readQ(files = sfiles_temp2, filetype = "structure")
slist_temp3 <- readQ(files = sfiles_temp3, filetype = "structure")

slist_wbnp1 <- readQ(files = sfiles_wbnp1, filetype = "structure")
slist_wbnp2 <- readQ(files = sfiles_wbnp2, filetype = "structure")

slist_admix1 <- readQ(files = sfiles_admix1, filetype = "structure")
slist_admix2 <- readQ(files = sfiles_admix2, filetype = "structure")
```

#### Optimal K clusters
```{r}
# Summarise K runs
struct_sum_temp1 <- summariseQ(tabulateQ(slist_temp1))
struct_sum_temp2 <- summariseQ(tabulateQ(slist_temp2))
struct_sum_temp3 <- summariseQ(tabulateQ(slist_temp3))

struct_sum_wbnp1 <- summariseQ(tabulateQ(slist_wbnp1))
struct_sum_wbnp2 <- summariseQ(tabulateQ(slist_wbnp2))

struct_sum_admix1 <- summariseQ(tabulateQ(slist_admix1))
struct_sum_admix2 <- summariseQ(tabulateQ(slist_admix2))

### Plot
# Full dataset
p1 <- evannoMethodStructure(data = struct_sum_temp1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp1_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf") 

grid.arrange(p1)

p2 <- evannoMethodStructure(data = struct_sum_temp2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp2_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf")

grid.arrange(p2)

p3 <- evannoMethodStructure(data = struct_sum_temp3, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp3_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf") 

grid.arrange(p3)

# WBNP subsets
w1 <- evannoMethodStructure(data = struct_sum_wbnp1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "WBNP1_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf")

grid.arrange(w1)

w2 <- evannoMethodStructure(data = struct_sum_wbnp2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "WBNP2_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf")

grid.arrange(w2)

a1 <- evannoMethodStructure(data = struct_sum_admix1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "admix1_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf") 

grid.arrange(a1)
 
a2 <- evannoMethodStructure(data = struct_sum_admix2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "admix2_evanno_K",
                           exportpath = "./plots/structure_plots/",
                           imgtype = "pdf") 

grid.arrange(a2)

```
## Evanno Data
```{r}
EMt1 <- data.frame(evannoMethodStructure(data = struct_sum_temp1, exportplot = F,
                           returnplot = F, returndata = T))


EMt2 <- data.frame(evannoMethodStructure(data = struct_sum_temp2, exportplot = F,
                           returnplot = F, returndata = T))


EMt3 <- data.frame(evannoMethodStructure(data = struct_sum_temp3, exportplot = F,
                           returnplot = F, returndata = T))

# WBNP subsets
EMwb1 <- data.frame(evannoMethodStructure(data = struct_sum_wbnp1, exportplot = F,
                           returnplot = F, returndata = T))



EMwb2 <- data.frame(evannoMethodStructure(data = struct_sum_wbnp2, exportplot = F,
                           returnplot = F, returndata = T))



EMad1 <- data.frame(evannoMethodStructure(data = struct_sum_admix1, exportplot = F,
                           returnplot = F, returndata = T))


 
EMad2 <- data.frame(evannoMethodStructure(data = struct_sum_admix2, exportplot = F,
                           returnplot = F, returndata = T))

```


### Barplotting
#### Temporal Group 1 (all samples - EINPP)
```{r}
# Align runs
slist_temp1 <- alignK(slist_temp1, "auto")

# Merge runs (average over 10 replicates)
temp1.mk2 <- mergeQ(slist_temp1[11:20])
temp1.mk3 <- mergeQ(slist_temp1[21:30])
temp1.mk4 <- mergeQ(slist_temp1[31:40])
temp1.mk5 <- mergeQ(slist_temp1[41:50])
temp1.mk6 <- mergeQ(slist_temp1[51:60])
temp1.mk7 <- mergeQ(slist_temp1[61:70])


temp1.merged.list <- as.qlist(list(temp1.mk2[[1]], temp1.mk3[[1]], temp1.mk4[[1]], 
                          temp1.mk5[[1]], temp1.mk6[[1]], temp1.mk7[[1]]))

ind_label_temp1 <- readQ(files = sfiles_temp1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp1 <- rownames_to_column(ind_label_temp1, "ID")
locations_temp1 <- left_join(ind_label_temp1, temp1.data, by = "ID")

loc_labels_temp1 <- locations_temp1$HERD %>% as.data.frame()

herd_order_temp1 <- c("EINPW", "MBS", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")

str_cols <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060")

plotQ(temp1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp1",
            exportpath = "./plots/structure_plots/")
```


#### Temporal Group 2 (all samples - EINPP)
```{r}
# Align runs
slist_temp2 <- alignK(slist_temp2, "auto")

# Merge runs (average over 10 replicates)
temp2.mk2 <- mergeQ(slist_temp2[11:20])
temp2.mk3 <- mergeQ(slist_temp2[21:30])
temp2.mk4 <- mergeQ(slist_temp2[31:40])
temp2.mk5 <- mergeQ(slist_temp2[41:50])



temp2.merged.list <- as.qlist(list(temp2.mk2[[1]], temp2.mk3[[1]], temp2.mk4[[1]], 
                          temp2.mk5[[1]]))

ind_label_temp2 <- readQ(files = sfiles_temp2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp2 <- rownames_to_column(ind_label_temp2, "ID")
locations_temp2 <- left_join(ind_label_temp2, temp2.data, by = "ID")

loc_labels_temp2 <- locations_temp2$HERD %>% as.data.frame()

herd_order_temp2 <- c("HZ", "MBS","RL", "WAB", "WENT")

str_cols <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060")

plotQ(temp2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp2, 
            subsetgrp = herd_order_temp2, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1.5,1),
      
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp2",
            exportpath = "./plots/structure_plots/")
```

#### Temporal Group 3 (all samples - EINPP)
```{r}
# Align runs
slist_temp3 <- alignK(slist_temp3, "auto")

# Merge runs (average over 10 replicates)
temp3.mk2 <- mergeQ(slist_temp3[11:20])
temp3.mk3 <- mergeQ(slist_temp3[21:30])
temp3.mk4 <- mergeQ(slist_temp3[31:40])
temp3.mk5 <- mergeQ(slist_temp3[41:50])
temp3.mk6 <- mergeQ(slist_temp3[51:60])
temp3.mk7 <- mergeQ(slist_temp3[61:70])
temp3.mk8 <- mergeQ(slist_temp3[71:80])
temp3.mk9 <- mergeQ(slist_temp3[81:90])
temp3.mk10 <- mergeQ(slist_temp3[91:100])
temp3.mk11 <- mergeQ(slist_temp3[101:110])
temp3.mk12 <- mergeQ(slist_temp3[111:120])
temp3.mk13 <- mergeQ(slist_temp3[121:130])
temp3.mk14 <- mergeQ(slist_temp3[131:140])
temp3.mk15 <- mergeQ(slist_temp3[141:150])


temp3.merged.list <- as.qlist(list(temp3.mk2[[1]], temp3.mk3[[1]], temp3.mk4[[1]], 
                          temp3.mk5[[1]], temp3.mk6[[1]], temp3.mk7[[1]],
                          temp3.mk8[[1]], temp3.mk9[[1]], temp3.mk10[[1]],
                          temp3.mk11[[1]], temp3.mk12[[1]], temp3.mk13[[1]], temp3.mk14[[1]], temp3.mk15[[1]]))

ind_label_temp3 <- readQ(files = sfiles_temp3[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp3 <- rownames_to_column(ind_label_temp3, "ID")
locations_temp3 <- left_join(ind_label_temp3, temp3.data, by = "ID")

loc_labels_temp3 <- locations_temp3$HERD %>% as.data.frame()

herd_order_temp3 <- c("AISH", "CL", "EINPW", "ETH", "HZ", "NAH", 
                      "NORD", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", 
                      "WBNPNY", "WBNPPAD", "WENT")

str_cols_1 <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", 
                "#1D72F5", "#FDF060", "#E619AD", "#00C5CD", "#8B4513", 
                "#228B22", "maroon", "#808080", "#F08080", "#4B0082")
plotQ(temp3.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp3, 
            subsetgrp = herd_order_temp3, 
            clustercol = str_cols_1,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7",
                      "K = 8", "K = 9", "K = 10", "K = 11", "K = 12", "K = 13", "K = 14", "K = 15"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 75,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp3",
            exportpath = "./plots/structure_plots/")
```
#### WBNP 1986 - 1999 (all samples - EINP)
```{r}
# Align runs
slist_wbnp1 <- alignK(slist_wbnp1, "auto")

# Merge runs (average over 10 replicates)
wbnp1.mk2 <- mergeQ(slist_wbnp1[11:20])
wbnp1.mk3 <- mergeQ(slist_wbnp1[21:30])
wbnp1.mk4 <- mergeQ(slist_wbnp1[31:40])
wbnp1.mk5 <- mergeQ(slist_wbnp1[41:50])

wbnp1.merged.list <- as.qlist(list(wbnp1.mk2[[1]], wbnp1.mk3[[1]], wbnp1.mk4[[1]], 
                          wbnp1.mk5[[1]]))

ind_label_wbnp1 <- readQ(files = sfiles_wbnp1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_wbnp1 <- rownames_to_column(ind_label_wbnp1, "ID")
locations_wbnp1 <- left_join(ind_label_wbnp1, wbnp1.data, by = "ID")

loc_labels_wbnp1 <- locations_wbnp1$HERD %>% as.data.frame()

herd_order_wbnp1 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD")

str_cols <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060")

plotQ(wbnp1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_wbnp1, 
            subsetgrp = herd_order_wbnp1, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 40,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Wbnp1",
            exportpath = "./plots/structure_plots/")
```

#### WBNP 2016 - 2025 (all samples - EINP)
```{r}
# Align runs
slist_wbnp2 <- alignK(slist_wbnp2, "auto")

# Merge runs (average over 10 replicates)
wbnp2.mk2 <- mergeQ(slist_wbnp2[11:20])
wbnp2.mk3 <- mergeQ(slist_wbnp2[21:30])
wbnp2.mk4 <- mergeQ(slist_wbnp2[31:40])
wbnp2.mk5 <- mergeQ(slist_wbnp2[41:50])
wbnp2.mk6 <- mergeQ(slist_wbnp2[51:60])

wbnp2_mergek2 <- mergeQ(slist_wbnp2[11:20])

wbnp2.merged.list <- as.qlist(list(wbnp2.mk2[[1]], wbnp2.mk3[[1]], wbnp2.mk4[[1]], 
                          wbnp2.mk5[[1]], wbnp2.mk6[[1]]))

ind_label_wbnp2 <- readQ(files = sfiles_wbnp2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_wbnp2 <- rownames_to_column(ind_label_wbnp2, "ID")
locations_wbnp2 <- left_join(ind_label_wbnp2, wbnp2.data, by = "ID")

loc_labels_wbnp2 <- locations_wbnp2$HERD %>% as.data.frame()

herd_order_wbnp2 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD", "WENT")

str_cols <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060")

plotQ(wbnp2.merged.list, 
            imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_wbnp2, 
            subsetgrp = herd_order_wbnp2, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 40,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Wbnp2",
            exportpath = "./plots/structure_plots/")
```
#### Admixture analysis 1986 - 1999 (all samples)
```{r}
# Align runs
slist_admix1 <- alignK(slist_admix1, "auto")

# Merge runs (average over 10 replicates)
admix1.mk2 <- mergeQ(slist_admix1[11:20])
admix1.mk3 <- mergeQ(slist_admix1[21:30])
admix1.mk4 <- mergeQ(slist_admix1[31:40])
admix1.mk5 <- mergeQ(slist_admix1[41:50])
admix1.mk6 <- mergeQ(slist_admix1[51:60])
admix1.mk7 <- mergeQ(slist_admix1[61:70])

admix1.merged.list <- as.qlist(list(admix1.mk2[[1]], admix1.mk3[[1]], admix1.mk4[[1]], 
                          admix1.mk5[[1]], admix1.mk6[[1]], admix1.mk7[[1]]))

ind_label_admix1 <- readQ(files = sfiles_admix1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_admix1 <- rownames_to_column(ind_label_admix1, "ID")
locations_admix1 <- left_join(ind_label_admix1, admix1.data, by = "ID")

loc_labels_admix1 <- locations_admix1$HERD %>% as.data.frame()

herd_order_admix1 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD")

str_cols_2 <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060", "#E619AD")

plotQ(admix1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_admix1, 
            subsetgrp = herd_order_admix1, 
            clustercol = str_cols_2,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 40,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "admix1",
            exportpath = "./plots/structure_plots/")
```
#### Admixture analysis 2016 - 2025 (all samples)
```{r}
# Align runs
slist_admix2 <- alignK(slist_admix2, "auto")

# Merge runs (average over 10 replicates)
admix2.mk2 <- mergeQ(slist_admix2[11:20])
admix2.mk3 <- mergeQ(slist_admix2[21:30])
admix2.mk4 <- mergeQ(slist_admix2[31:40])
admix2.mk5 <- mergeQ(slist_admix2[41:50])
admix2.mk6 <- mergeQ(slist_admix2[51:60])
admix2.mk7 <- mergeQ(slist_admix2[61:70])
admix2.mk8 <- mergeQ(slist_admix2[71:80])

admix2.merged.list <- as.qlist(list(admix2.mk2[[1]], admix2.mk3[[1]], admix2.mk4[[1]], 
                          admix2.mk5[[1]], admix2.mk6[[1]], admix2.mk7[[1]], admix2.mk8[[1]]))

ind_label_admix2 <- readQ(files = sfiles_admix2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_admix2 <- rownames_to_column(ind_label_admix2, "ID")
locations_admix2 <- left_join(ind_label_admix2, admix2.data, by = "ID")

loc_labels_admix2 <- locations_admix2$HERD %>% as.data.frame()

herd_order_admix2 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD", "WENT")

str_cols_2 <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060", "#E619AD")

plotQ(admix2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_admix2, 
            subsetgrp = herd_order_admix2, 
            clustercol = str_cols_2,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 40,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "admix2",
            exportpath = "./plots/structure_plots/")
```

## DAPC
#### WBNP samples 1986 - 1999
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp1 <- find.clusters(wbnp1, max.n.clust = 7, n.pca = (nInd(wbnp1) -1))

# Print BIC plot
pdf(file = "./plots/WBNP1_BIC.pdf", width = 7, height = 5)
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 1999" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(wbnp1), grp1$grp), col.lab=paste("inf", 1:5),
row.lab=paste("ori", 1:5))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx <- xvalDapc(tab(wbnp1, NA.method = "mean"), pop(wbnp1))

# Determine optimal number of PCs
numpc1 <- as.numeric(pramx$`Number of PCs Achieving Highest Mean Success`)

# DAPC with known parameters
dapc1 <- dapc(wbnp1, pop(wbnp1), n.pca = numpc1, n.da = (length(pop(wbnp1))-1))
dapc1


```
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp2 <- find.clusters(wbnp2, max.n.clust = 8, n.pca = (nInd(wbnp2) -1))

# Print BIC plot
pdf(file = "./plots/WBNP2_BIC.pdf", width = 7, height = 5)
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(wbnp2), grp2$grp), col.lab=paste("inf", 1:6),
row.lab=paste("ori", 1:6))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx2 <- xvalDapc(tab(wbnp2, NA.method = "mean"), pop(wbnp2),
                   xval.plot = TRUE)

# Determine optimal number of PCs
numpc2 <- as.numeric(pramx2$`Number of PCs Achieving Lowest MSE`)

# DAPC with known parameters
dapc2 <- dapc(wbnp2, pop(wbnp2), n.pca = numpc2, n.da = (length(pop(wbnp2))-1))
dapc2

```

```{r}
# Plot
pdf(file = "./plots/wbnp_DAPC_stacked.pdf", width = 20, height = 10) 

# Set up  grid 1 row, 2 columns
par(mfrow = c(1, 2)) 

# Plot 1
scatter(dapc1,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, col=str_cols,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.2, csub = 2,
sub = paste0("DA1: ", round(dapc1$eig[1], 1), "% | DA2: ", round(dapc1$eig[2], 1), "%"))
mtext("a)", side = 3, adj = -4.5, line = 2, cex = 2, font = 2)

# Plot 2
scatter(dapc2,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, col=str_cols,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.2, csub = 2,
sub = paste0("DA1: ", round(dapc2$eig[1], 1), "% | DA2: ", round(dapc2$eig[2], 1), "%"))
mtext("b)", side = 3, adj = -4.5, line = 2, cex = 2, font = 2)

dev.off()
```
#### Joint BIC Plot
```{r}
# Print BIC plot
pdf(file = "./plots/WBNP_BIC_Joined.pdf", width = 15, height = 5)

par(mfrow = c(1,2))
# Temporal group 1
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 1999" ,
     col = "blue", pch = 1)
mtext("a)", side = 3, adj = 0, line = 2, cex = 2, font = 2)

# Temporal group 3
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)
mtext("b)", side = 3, adj = 0, line = 2, cex = 2, font = 2)

dev.off()

```
# STRUCTURE Plots of interest
```{r}
# Temp1
plotQ(temp1.mk3, 
            basesize = 10, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 3",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp1_k3",
            exportpath = "./plots/structure_plots/")

# Temp2
plotQ(temp2.mk4, 
            basesize = 10, 
            grplab = loc_labels_temp2, 
            subsetgrp = herd_order_temp2, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 4",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp2_k4",
            exportpath = "./plots/structure_plots/")

# Temp3
plotQ(temp3.mk7, 
            basesize = 10, 
            grplab = loc_labels_temp3, 
            subsetgrp = herd_order_temp3, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 7",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp3_k7",
            exportpath = "./plots/structure_plots/")

# WBNP1
plotQ(wbnp1.mk2, 
            basesize = 10, 
            grplab = loc_labels_wbnp1, 
            subsetgrp = herd_order_wbnp1, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 2",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "WBNP1_k2",
            exportpath = "./plots/structure_plots/")

# WBNP2
plotQ(wbnp2.mk2, 
            basesize = 10, 
            grplab = loc_labels_wbnp2, 
            subsetgrp = herd_order_wbnp2, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 2",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "WBNP2_k2",
            exportpath = "./plots/structure_plots/")

# Admix 1
plotQ(admix1.mk3, 
            basesize = 10, 
            grplab = loc_labels_admix1, 
            subsetgrp = herd_order_admix1, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 3",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "admix1_k3",
            exportpath = "./plots/structure_plots/")

# Admix 2
plotQ(admix2.mk4, 
            basesize = 10, 
            grplab = loc_labels_admix2, 
            subsetgrp = herd_order_admix2, 
            clustercol = str_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 4",
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 15,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "pdf",
            outputfilename = "admix2_k4",
            exportpath = "./plots/structure_plots/")
```

