---
title: "PCA_STRUCT_2"
author: "Alex Depew"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Cleaned_locus_data_20.02.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Cleaned_locus_data_20.02.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=2015,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp2.sum <- summary(temp2)
```

Remove Elk Island Plains for Structure Analyses
```{r}
#temp1 <- temp1[temp1@pop != "EINPP"]
#temp1.sum <- summary(temp1)

#temp2 <- temp2[temp2@pop != "EINPP"]
#temp2.sum <- summary(temp2)
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 
```{r}
cols <- c(
  
  # --- EINPW Derivatives ---
  "#E31A1C", # AISH
  "#1F78B4", # CL
  "#33a02c", # EINPW
  "#CAB2D6", # ETH
  "#FB9A99", # HZ
  "#6A3D9A", # NAH
  "#800000", # NORD

  # --- WBNP Metapopulation ---
  "#1B9E77", # WBNPGR
  "#D95F02", # WBNPHC
  "#7570B3", # WBNPLB
  "#E7298A", # WBNPNY
  "#66A61E", # WBNPPAD
  "#E6AB02", # WENT
  "#A6761D", # RL
  "#666666", # WAB
  
  # --- WBNP Derivatives ---
  "darkred", # MBS
  
  # --- Plains Bison ---
  "#FF7F00" # EINPP


)
all_herds <- c(
  "AISH", "CL", "EINPP", "EINPW", "ETH", "HZ", "MBS", "NAH", "NORD", 
  "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

herd_cols <- setNames(cols, all_herds)


```


# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)

x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)
```


```{r}
levels_temp1 <- levels(pop(temp1))
temp1_cols <- herd_cols[levels_temp1]
png(filename="plots/pca_temp1v2.png",
    width = 1200, height = 840, units = "px", pointsize = 32)

# Now use the custom color palette in s.class
s.class(pca.temp1$li, fac = pop(temp1), xax = 1, yax = 2, col = transp(temp1_cols, .7), axesel = FALSE, cstar = 0, cpoint = 1, clabel = 0) 

# Add scatter points for eigenvalues as before
add.scatter.eig(pca.temp1$eig[1:20], 3, 1, 2, ratio = .15, "bottomright")
dev.off()
```
```{r}
# Colour selection
levels_temp2 <- levels(pop(temp2))
temp2_cols <- herd_cols[levels_temp2]

png(filename="plots/pca_temp2v2.png",
    width = 1200, height = 840, units = "px", pointsize = 32)

# Now use the custom color palette in s.class
s.class(pca.temp2$li, fac = pop(temp2), xax = 1, yax = 2, col = transp(temp2_cols, .7), axesel = FALSE, cstar = 0, cpoint = 1, clabel = 0) 

# Add scatter points for eigenvalues as before
add.scatter.eig(pca.temp2$eig[1:20], 3, 1, 2, ratio = .15, "bottomright")
dev.off()
```

