---
title: "PCA_STRUCT"
author: "Alex Depew"
date: "2026-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   


## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=1999,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >= 2000 & bison.genind@other$YEAR.ASSIGN <= 2015, ]
temp2.sum <- summary(temp2)

temp3 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp3.sum <- summary(temp3)


# Remove populations which do not contain significant sample sizes in temporal groups
table(pop(temp1))
table(pop(temp2)) # WBNP samples, ETH, NORD samples can be removed
table(pop(temp3))

# Remove herds from temporal group 2 and 3
remove_herds <- c("WBNPPAD", "WBNPNY", "ETH", "NORD", "WENT")
temp2 <- temp2[!pop(temp2) %in% remove_herds, ]
temp2.sum <- summary(temp2) # Recalculate summary

temp3 <- temp3[!pop(temp3) %in% "WBNP"]
temp3.sum <- summary(temp3) # Recalculate summary
```



# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 

### Temporal Group 1
```{r}
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)


s.class(pca.temp1$li, fac = pop(temp1), col = transp(funky(15)),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F)

# How much % variance do the axes represent?
eig.perc.temp1 <- 100*pca.temp1$eig/sum(pca.temp1$eig)
head(eig.perc.temp1)

# Allele contributions to variance
loadingplot(pca.temp1$c1^2)
```
### Temporal Group 2
```{r}
x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)


s.class(pca.temp2$li, fac = pop(temp2), col = transp(funky(15)),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F)

# How much % variance do the axes represent?
eig.perc.temp2 <- 100*pca.temp2$eig/sum(pca.temp2$eig)
head(eig.perc.temp2)

# Allele contributions to variance
loadingplot(pca.temp2$c1^2)
```
### Temporal Group 3
```{r}
x.temp3 <- tab(temp3, freq = TRUE, NA.method = "mean")
pca.temp3 <- dudi.pca(x.temp3, center = T, scale = F, scannf = F, nf = 2)


s.class(pca.temp3$li, fac = pop(temp3), col = transp(funky(15)),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F)

# How much % variance do the axes represent?
eig.perc.temp3 <- 100*pca.temp3$eig/sum(pca.temp3$eig)
head(eig.perc.temp3)

# Allele contributions to variance
loadingplot(pca.temp3$c1^2)
```
# DACP

# STRUCTURE Analyses
```{r}
# Populations need to be in unique integer format for input into STRUCTURE:
temp1.struct.gen <- temp1
pop(temp1.struct.gen) <- as.numeric(as.factor(pop(temp1.struct.gen)))


# Format in STRUCTURE format
library(graph4lg)
genind_to_structure(temp1.struct.gen, output = "structure/temp1.structure.txt")
```

```{r}
library(pophelper)
library(gridExtra)
sfiles <- list.files(path = "C:/Users/adepe/OneDrive/Documents/R Working Directory/Bison_github/structure/results_f/", full.names = TRUE)

slist <- readQ(files = sfiles)
readQ(files = sfiles, filetype = "structure")

sr1 <- summariseQ(tabulateQ(slist))

p <- evannoMethodStructure(data = sr1, exportplot = F,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7)
grid.arrange(p)
```

