---
title: "PCA_STRUCT_2"
author: "Alex Depew"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# File information
This script is used to evaluate temporal patterns in population structure in all wood bison herds in Canada. 

Files used described below:

- `Cleaned_locus_data_20.02.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
library(ggplot2)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Cleaned_locus_data_20.02.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into three temporal groups. Temporal groups are: 1: pre-1999, 2: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@other$YEAR.ASSIGN <=2015,]
temp1.sum <- summary(temp1) # Create summary of genind object

temp2 <- bison.genind[bison.genind@other$YEAR.ASSIGN >=2016,]
temp2.sum <- summary(temp2)
```

Remove Elk Island Plains for Structure Analyses
```{r}
#temp1 <- temp1[temp1@pop != "EINPP"]
#temp1.sum <- summary(temp1)

#temp2 <- temp2[temp2@pop != "EINPP"]
#temp2.sum <- summary(temp2)
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data. 

```{r}
library(Polychrome)

# 1. Get a unique list of all possible herds across the entire project
all_herds <- sort(unique(cleaned.data$HERD))

# 2. Create a palette with a seed for reproducibility
set.seed(675588585) # Using a fixed seed ensures the same colors are generated every time
herd_palette <- createPalette(length(all_herds), 
                              seedcolors = c("#5F4690", "#1D6996", "#38A6A5", "#CC503E"),
                              range = c(30, 100)) # Controls brightness/saturation

herd_palette <- c("#800000", 
                  "#469990", 
                  "#ffe119", 
                  "#3cb44b", 
                  "#911eb4",
                  "#f58231",
                  "#9A6324",
                  "#808000",
                  "#000000",
                  "#000075",
                  "#e6194B",
                  "#bfef45",
                  "#ffd8b1",
                  "#4363d8",
                  "#f032e6",
                  "#a9a9a9",
                  "#fabed4"

)

# 3. Assign names to the colors so they are "locked" to the herd ID
herd_cols <- setNames(as.character(herd_palette), all_herds)

# Preview the colors
swatch(herd_cols)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.temp1 <- tab(temp1, freq = TRUE, NA.method = "mean")
pca.temp1 <- dudi.pca(x.temp1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp1 <- 100*pca.temp1$eig/sum(pca.temp1$eig)

x.temp2 <- tab(temp2, freq = TRUE, NA.method = "mean")
pca.temp2 <- dudi.pca(x.temp2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.temp2 <- 100*pca.temp2$eig/sum(pca.temp2$eig)

##### -------
# Strictly for plotting (legend)
x.global <- tab(bison.genind, freq = TRUE, NA.method = "mean")
pca.global <- dudi.pca(x.global, center = T, scale = F, scannf = F, nf =2)
# Axes % Variance
eig.perc.global <- 100*pca.global$eig/sum(pca.global$eig)

#### axis limits
xlims <- c(-2, 3.5)
ylims <- c(-2, 2.1)
```


## PCA plot with ggplot
```{r}
temp1@pop <- factor(temp1@pop, levels = sort(unique(as.character(temp1@pop))))
levels_temp1 <- levels(temp1@pop)
sorted_cols <- herd_cols[sorted_levels]

temp1_cols <- herd_cols[levels_temp1]
# Plot PCA object
pca1 <- fviz_pca_ind(pca.temp1,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp1@pop,
             repel = TRUE,
             palette = temp1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.temp1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pca1
```

```{r}
temp2@pop <- factor(temp2@pop, levels = sort(unique(as.character(temp2@pop))))
levels_temp2 <- levels(temp2@pop)
sorted_cols <- herd_cols[sorted_levels]

temp2_cols <- herd_cols[levels_temp2]
# Plot PCA object
pca2 <- fviz_pca_ind(pca.temp2,
             axes = c(1, 2),
             geom = "blank",
             habillage = temp2@pop,
             repel = TRUE,
             palette = temp2_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = temp2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = temp2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.temp2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.temp2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pca2
```

```{r}
bison.genind@pop <- factor(bison.genind@pop, levels = sort(unique(as.character(bison.genind@pop))))
sorted_levels <- levels(bison.genind@pop)
sorted_cols <- herd_cols[sorted_levels]

global_cols <- herd_cols[sorted_levels]

# Plot PCA object
pcag <- fviz_pca_ind(pca.global,
             axes = c(1, 2),
             geom = "blank",
             habillage = bison.genind@pop,
             repel = TRUE,
             palette = global_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = bison.genind@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = bison.genind@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold")) + 
  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.global[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.global[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 2, shape = 19)))

```

### Arrange Plots together
```{r}
pca1_nl <- pca1 + theme(legend.position = "none")
pca2_nl <- pca2 + theme(legend.position = "none")

legend <- get_legend(pcag)

global_pcas <- grid.arrange(pca1_nl, legend, pca2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/global_pcas.svg",
       plot = global_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
# STRUCTURE Analysis 
```{r}
# Temp1 structure
temp1.str <- temp1
pop(temp1.str) <- as.numeric(as.factor(pop(temp1.str)))

genind_to_structure(temp1.str,
                    output = "STRUCTURE_files/temp1.w_plains.structure.txt")
# Temp2 structure
temp2.str <- temp2
pop(temp2.str) <- as.numeric(as.factor(pop(temp2.str)))

genind_to_structure(temp2.str,
                    output = "STRUCTURE_files/temp2.w_plains.structure.txt")
```


#### Load Data
```{r}
# List files
# Full datasets
sfiles_temp1 <- list.files("./STRUCTURE_files/results/temp1_w_plains/", full.names = TRUE)
sfiles_temp2 <- list.files("./STRUCTURE_files/results/temp2_w_plains/", full.names = TRUE)

# Qlist (pophelper)
slist_temp1 <- readQ(files = sfiles_temp1, filetype = "structure")
slist_temp2 <- readQ(files = sfiles_temp2, filetype = "structure")

# Summarise runs
struct_sum_temp1 <- summariseQ(tabulateQ(slist_temp1))
struct_sum_temp2 <- summariseQ(tabulateQ(slist_temp2))

# K plots
p1 <- evannoMethodStructure(data = struct_sum_temp1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "Temp1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- p1$grobs[[1]]
p_dk <- p1$grobs[[4]]

library(gridExtra)
library(grid)
# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("a)", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("b)", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_w1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/temp1.k.svg", 
       plot = evanno_w1, 
       width = 27, 
       height = 11)
```

```{r}

temp1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID))
# Align runs
slist_temp1 <- alignK(slist_temp1, "auto")

# Merge runs (average over 10 replicates)
temp1.mk2 <- mergeQ(slist_temp1[11:20])
temp1.mk3 <- mergeQ(slist_temp1[21:30])
temp1.mk4 <- mergeQ(slist_temp1[31:40])
temp1.mk5 <- mergeQ(slist_temp1[41:50])
temp1.mk6 <- mergeQ(slist_temp1[51:60])
temp1.mk7 <- mergeQ(slist_temp1[61:70])
temp1.mk8 <- mergeQ(slist_temp1[71:80])
temp1.mk9 <- mergeQ(slist_temp1[81:90])
temp1.mk10 <- mergeQ(slist_temp1[91:100])
temp1.mk11 <- mergeQ(slist_temp1[101:110])
temp1.mk12 <- mergeQ(slist_temp1[111:120])


temp1.merged.list <- as.qlist(list(temp1.mk2[[1]], temp1.mk3[[1]], temp1.mk4[[1]], 
                          temp1.mk5[[1]], temp1.mk6[[1]], temp1.mk7[[1]], temp1.mk8[[1]],
                          temp1.mk9[[1]], temp1.mk10[[1]], temp1.mk11[[1]], temp1.mk12[[1]]))

ind_label_temp1 <- readQ(files = sfiles_temp1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_temp1 <- rownames_to_column(ind_label_temp1, "ID")
locations_temp1 <- left_join(ind_label_temp1, temp1.data, by = "ID")

loc_labels_temp1 <- locations_temp1$HERD %>% as.data.frame()

herd_order_temp1 <- c("EINPP", "EINPW", "HZ", "MBS", "RL", "WAB", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

str_cols <- c("#0089B2", "#DF0101", "#77CE61", "#A945FF", "#FF9326", "#1D72F5","#FDF060")

plotQ(temp1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_temp1, 
            subsetgrp = herd_order_temp1, 
            clustercol = herd_palette,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8", "K = 9", "K = 10", "K = 11", "K = 12"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "pdf",
            outputfilename = "Temp1_w_plains",
            exportpath = "./plots/structure_plots/")
```
```



