---
title: "EINP and WBNP Temporal Comparisons"
author: "Alex Depew"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script works through the subquestion 1: the temporal comparison of genetic diversity and structure in Elk Island National Park (EINP) and Wood Buffalo National Park (WBNP).

## Load Required Packages
```{r}
rm(list = ls())
library(PopGenReport)
library(adegenet)
library(hierfstat)
library(strataG)
library(tidyverse)
library(poppr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026",
                       check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)

```

#### Subset data 
Data is subset into two temporal groups. Populations included in Wilson 1999 and their contemporary counterparts. 
```{r}
# Create datasets containing samples of interest for temporal comparisons
temp1 <- bison.genind[bison.genind@pop %in% c("WBNPGR",
                                              "WBNPPAD",
                                              "WBNPNY",
                                              "WBNPLB",
                                              "WBNPHC",
                                              "EINPP",
                                              "EINPW") & bison.genind@other$YEAR.ASSIGN <=1999,]

temp1.sum <- summary(temp1) # Create summary of genind object
temp1.sum

temp2 <- bison.genind[bison.genind@pop %in% c("WBNPGR",
                                              "WBNPPAD",
                                              "WBNPNY",
                                              "WBNPLB",
                                              "WBNPHC",
                                              "EINPP",
                                              "EINPW") & bison.genind@other$YEAR.ASSIGN >=2016, ]
temp2.sum <- summary(temp2)
```

# Descriptive Statistics
```{r}
# For use in strataG, genind objects must be converted to gtypes objects
temp1.g <- genind2gtypes(temp1)
temp2.g <- genind2gtypes(temp2)

# For use in external programs (HP-RARE), data must be formatted in GENEPOP format
library(graph4lg)
genind_to_genepop(temp1,
                  output = "genepop_files/temp1.genepop.txt")

genind_to_genepop(temp2,
                  output = "genepop_files/temp2.genepop.txt")

```



```{r}
# Number of samples per population
temp1.N = temp1.sum$n.by.pop
temp1.N

# Heterozygosity ----
# Expected heterozygosity
library(hierfstat)
temp1.stats <- basic.stats(temp1,
                           diploid = TRUE,
                           digits = 4)

temp1.Hexp <- round(colMeans(temp1.stats$Hs), 3)
temp1.Hexp

# Observed heterozygosity
temp1.Hobs <- round(colMeans(temp1.stats$Ho), 3)
temp1.Hobs

# Mean number of alleles ----
pf <- temp1.stats$pop.freq

# Count alleles per population per locus, then mean across loci
mean_Na <- sapply(pf, function(mat) {
  if(is.null(dim(mat))) mat <- matrix(mat, ncol = length(mat))  # handle monomorphic loci
  colSums(mat > 0)
})

# Take the mean across loci and round
temp1.Na <- round(rowMeans(mean_Na), 2)
temp1.Na

# Allelic Richness
library(hierfstat)
temp1.Ar <- allelic.richness(temp1, min.n = NULL, diploid = TRUE)
temp1.Ar.mean <- round(colMeans(temp1.Ar$Ar), 2)
temp1.Ar.mean

# Effective number of alleles
temp1.AE <- round(1/(1 - temp1.Hexp), 2)


# Private Alleles ----
temp1.NP <- privateAlleles(temp1.g)
temp1.NP
temp1.NP.sums <- data.frame(NP = colSums(temp1.NP)) %>% 
  rownames_to_column(var = "Population")

# Private Allelic Richness
temp1.PR <- read.table("temp1.hprare.txt", 
                       header = FALSE,
                       skip = 3)
colnames(temp1.PR) <- c("Population", "RP")
temp1.PR.short <- temp1.PR %>% 
  select(1:2) %>% 
  mutate(Population = str_remove(Population, "_.*")) # Note: rarefied to minimum genotypes = 18

# Join NP and PR by Herd
temp1.priv <- left_join(temp1.NP.sums, temp1.PR.short, by = "Population")

# Fis
temp1.Fis <- round(colMeans(temp1.stats$Fis, na.rm = TRUE), 3)
temp1.Fis

# Table
temp1.descr <- data.frame(
  Population = colnames(temp1.stats$Hs),
  n = temp1.N,
  HEXP = temp1.Hexp,
  HOBS = temp1.Hobs,
  Na = temp1.Na,
  AR = temp1.Ar.mean,
  AE = temp1.AE,
  FIS = temp1.Fis
) %>% 
left_join(temp1.priv)

temp1.descr

# Write to csv
write.csv(temp1.descr,
          "Descriptive Stats Temp1.csv")
```

```{r}
# Number of samples per population
temp2.N <- temp2.sum$n.by.pop
temp2.N

# Heterozygosity ----
# Expected heterozygosity
temp2.stats <- basic.stats(temp2,
                           diploid = TRUE,
                           digits = 4)

temp2.Hexp <- round(colMeans(temp2.stats$Hs), 3)
temp2.Hexp

# Observed heterozygosity
temp2.Hobs <- round(colMeans(temp2.stats$Ho), 3)
temp2.Hobs

# Mean number of alleles ----
pf2 <- temp2.stats$pop.freq

mean_Na2 <- sapply(pf2, function(mat) {
  if(is.null(dim(mat))) mat <- matrix(mat, ncol = length(mat))  # handle monomorphic loci
  colSums(mat > 0)
})

temp2.Na <- round(rowMeans(mean_Na2), 2)
temp2.Na

# Allelic richness
temp2.Ar <- allelic.richness(temp2, min.n = NULL, diploid = TRUE)
temp2.Ar.mean <- round(colMeans(temp2.Ar$Ar), 2)
temp2.Ar.mean

# Effective number of alleles
temp2.AE <- round(1/(1 - temp2.Hexp), 2)

# Private Alleles ----
temp2.NP <- privateAlleles(temp2.g)
temp2.NP
temp2.NP.sums <- data.frame(NP = colSums(temp2.NP)) %>% 
  rownames_to_column(var = "Population")

# Private Allelic Richness
temp2.PR <- read.table("temp2.hprare.txt", 
                       header = FALSE,
                       skip = 3)
colnames(temp2.PR) <- c("Population", "RP")
temp2.PR <- temp2.PR %>% 
  select(1:2) %>% 
  mutate(Population = str_remove(Population, "_.*")) # Note: rarefied to minimum genotypes = 10

# Join NP and PR by Herd
temp2.priv <- left_join(temp2.NP.sums, temp2.PR, by = "Population")


# Fis
temp2.Fis <- round(colMeans(temp2.stats$Fis, na.rm = TRUE), 3)
temp2.Fis

# Table
temp2.descr <- data.frame(
  Population = colnames(temp2.stats$Hs),
  n = temp2.N,
  HEXP = temp2.Hexp,
  HOBS = temp2.Hobs,
  Na = temp2.Na,
  AR = temp2.Ar.mean,
  AE = temp2.AE,
  FIS = temp2.Fis
  ) %>% 
  left_join(temp2.priv)

temp2.descr

# Write to csv
write.csv(temp2.descr,
          "Descriptive Stats Temp2.csv")
```


```{r}
# Compare genetic diversity between temporal groups
t.test(temp1.descr$HEXP, temp2.descr$HEXP, paired = TRUE)

# Paired t-test per herd for HEXP
pvals_HE <- mapply(function(i){
  t.test(
    x = temp1.descr$HEXP[i],
    y = temp2.descr$HEXP[i],
    paired = TRUE
  )$p.value
}, 1:nrow(temp1.descr))

# Combine results
results_HE <- data.frame(
  Herd = herds,
  p_value = pvals_HE,
  p_value_bonf = p.adjust(pvals_HE, method = "bonferroni")
)

results_HE
```


# Genetic Distance
Weir and Cockerham (1984) estimate of Fst is used to calculate genetic distances. 
```{r}
# Subset data to reduce computation time
temp1.sub <- popsub(temp1, sublist = c("WBNPGR",
                                        "WBNPPAD",
                                        "WBNPNY",
                                        "WBNPLB",
                                        "WBNPHC",
                                        "EINPP",
                                        "EINPW"))

temp2.sub <- popsub(temp2, sublist = c("WBNPGR",
                                        "WBNPPAD",
                                        "WBNPNY",
                                        "WBNPLB",
                                        "WBNPHC",
                                        "EINPP",
                                        "EINPW"))

# Calculate pairwise Fst
temp1.fst <- genet.dist(temp1.sub, method = "WC84") %>% 
  round(digits = 3)
temp1.fst

temp2.fst <- genet.dist(temp2.sub, method = "WC84") %>% 
  round(digits = 3)

# Visualize with heatmaps ----

# Label order
temp1.order <- c("WBNPGR",
                 "WBNPPAD",
                 "WBNPNY",
                 "WBNPLB",
                 "WBNPHC",
                 "EINPP",
                 "EINPW")

temp2.order <- c("WBNPGR",
                 "WBNPPAD",
                 "WBNPNY",
                 "WBNPLB",
                 "WBNPHC",
                 "EINPP",
                 "EINPW")

# Change order of rows and cols in matrices
temp1.fst.mat <- as.matrix(temp1.fst)
temp1.fst.mat.1 <- temp1.fst.mat[temp1.order, ]
temp1.fst.mat.2 <- temp1.fst.mat[ ,temp1.order]

temp2.fst.mat <- as.matrix(temp2.fst)
temp2.fst.mat.1 <- temp2.fst.mat[temp2.order, ]
temp2.fst.mat.2 <- temp2.fst.mat[ ,temp2.order]

# Df creation
temp1.ind <- which(upper.tri(temp1.fst.mat.2), arr.ind = TRUE)
temp1.fst.df <- data.frame(Site1 = dimnames(temp1.fst.mat.2)[[2]][temp1.ind[,2]],
                           Site2 = dimnames(temp1.fst.mat.2)[[1]][temp1.ind[,1]],
                           Fst = temp1.fst.mat.2[ temp1.ind ])

temp2.ind <- which(upper.tri(temp2.fst.mat.2), arr.ind = TRUE)
temp2.fst.df <- data.frame(Site1 = dimnames(temp2.fst.mat.2)[[2]][temp2.ind[,2]],
                           Site2 = dimnames(temp2.fst.mat.2)[[1]][temp2.ind[,1]],
                           Fst = temp2.fst.mat.2[ temp2.ind ])

# Keep order of levels in dataframe for plotting
temp1.fst.df$Site1 <- factor(temp1.fst.df$Site1, levels = unique(temp1.fst.df$Site1))
temp1.fst.df$Site2 <- factor(temp1.fst.df$Site2, levels = unique(temp1.fst.df$Site2))

temp2.fst.df$Site1 <- factor(temp2.fst.df$Site1, levels = unique(temp2.fst.df$Site1))
temp2.fst.df$Site2 <- factor(temp2.fst.df$Site2, levels = unique(temp2.fst.df$Site2))

# Make all negative values == 0
temp1.fst.df$Fst[temp1.fst.df$Fst < 0] <- 0

temp2.fst.df$Fst[temp2.fst.df$Fst < 0] <- 0

# Italic label
temp1.fst.label <-  expression(italic("F")[ST])

temp2.fst.label <-  expression(italic("F")[ST])

# Set middle Fst value to set gradient argument
temp1.mid = max(temp1.fst.df$Fst) / 2

temp2.mid = max(temp2.fst.df$Fst) / 2

# Plot ----
temp1.fst.plot <- ggplot(data = temp1.fst.df, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = Fst), color="black", size = 3)+
                  scale_fill_gradient2(low = "blue", mid = "pink", high = "red", midpoint = temp1.mid, name = temp1.fst.label, 
                  limits = c(0, max(temp1.fst.df$Fst)), breaks = c(0, 0.05, 0.10, 0.15))+
                  scale_x_discrete(expand = c(0,0))+
                  scale_y_discrete(expand = c(0,0), position = "right")+
                  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
                        axis.title = element_blank(),
                        panel.grid = element_blank(),
                        panel.background = element_blank(),
                        legend.position = "right",
                        legend.title = element_text(size = 14, face = "bold"),
                        legend.text = element_text(size = 10)
                        )
temp1.fst.plot


# Temporal group 2
temp2.fst.plot <- ggplot(data = temp2.fst.df, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = Fst), color="black", size = 3)+
                  scale_fill_gradient2(low = "blue", mid = "pink", high = "red", midpoint = temp2.mid, name = temp2.fst.label, 
                  limits = c(0, max(temp2.fst.df$Fst)), breaks = c(0, 0.05, 0.10, 0.15))+
                  scale_x_discrete(expand = c(0,0))+
                  scale_y_discrete(expand = c(0,0), position = "right")+
                  theme(axis.text = element_text(colour = "black", size = 10, face = "bold"),
                        axis.title = element_blank(),
                        panel.grid = element_blank(),
                        panel.background = element_blank(),
                        legend.position = "right",
                        legend.title = element_text(size = 14, face = "bold"),
                        legend.text = element_text(size = 10)
                        )

temp2.fst.plot
```
## Jost's D & G'st
```{r}
## Jost's D
# Temporal Group 1
library(mmod)
temp1.D.dist <- pairwise_D(temp1,
                         linearized = F,
                         hsht_mean = "arithmetic")
temp1.D.dist

# Temporal Group 2
temp2.D.dist <- pairwise_D(temp2,
                         linearized = F,
                         hsht_mean = "arithmetic")
temp2.D.dist

## Hedrick's G'st
# Temporal Group 1
temp1.Gst.dist <- pairwise_Gst_Hedrick(temp1, 
                                     linearized = F)

temp1.Gst.dist

# Temporal Group 2
temp2.Gst.dist <- pairwise_Gst_Hedrick(temp2, 
                                     linearized = F)

temp2.Gst.dist

# Plot Jost's D


temp1.D.mat
ggplot(data = temp1.D.mat, aes(x = Site1, y = Site2, fill = Fst))+
                  geom_tile(colour = "black")+
                  geom_text(aes(label = D), color="black", size = 3)+
                  scale_fill_gradient2(low = "yellow", mid = "orange", high = "red", midpoint = temp1.D.mid)

```


# Multivariate Analyses (PCA and DAPC)
```{r}
# Replace missing data
sum(is.na(temp1$tab)) # 42 missing data
sum(is.na(temp2$tab)) # 617 missing data

temp1.pca.mat <- scaleGen(temp1, NA.method = "mean")
temp2.pca.mat <- scaleGen(temp2, NA.method = "mean")

# Plot eigenvalues
temp1.pca1 <-dudi.pca(temp1.pca.mat,
                      cent = FALSE,
                      scale = FALSE,
                      scannf = FALSE,
                      nf = 3)

barplot(temp1.pca1$eig[1:50],
        main = "Temporal Group 1 PCA eigenvalues",
        col = heat.colors(50))



temp2.pca1 <-dudi.pca(temp2.pca.mat,
                      cent = FALSE,
                      scale = FALSE,
                      scannf = FALSE,
                      nf = 3)

barplot(temp2.pca1$eig[1:50],
        main = "Temporal Group 2 PCA eigenvalues",
        col = heat.colors(50))

# % of variance explained by eigenvalues
temp1.pc_percent <- 100 * temp1.pca1$eig / sum(temp1.pca1$eig)
round(temp1.pc_percent[1:2], 2) # 7.13% and 2.58%

temp2.pc_percent <- 100 * temp2.pca1$eig / sum(temp2.pca1$eig)
round(temp2.pc_percent[1:2], 2) # 6.39% and 2.08%

# PCA plot
col <- funky(15)

# Pre-1999
s.class(temp1.pca1$li,
        pop(temp1),
        xax = 1,
        yax = 2,
        col = transp(col,.6), 
        axesell = FALSE,
        cstar = 0,
        cpoint = 3,
        grid = FALSE,
        sub = "PCA of Temporal Group 1")

# Post-2016
s.class(temp2.pca1$li,
        pop(temp2),
        xax = 1,
        yax = 2,
        col = transp(col,.6), 
        axesell = FALSE,
        cstar = 0,
        cpoint = 3,
        grid = FALSE,
        sub = "PCA of Temporal Group 2")

```
From temporal group 2, it appears as if the WBNP samples show some linear separation in the PCA, consistent with the relative north–south positioning of the herds, suggesting isolation by distance which was not as clearly resolved in the first temporal group.

# Isolate WBNP samples
```{r}
# PCA with contemporary WBNP Samples
WBNP.temp2 <- temp2[grepl("^WBNP", pop(temp2)), ]
 # Subset to contain only WBNP samples

sum(is.na(WBNP.temp2$tab)) # 294 missing data

WBNP.temp2.pca.mat <- scaleGen(WBNP.temp2, NA.method = "mean")

# Plot eigenvalues
WBNP.temp2.pca1 <-dudi.pca(WBNP.temp2.pca.mat,
                      cent = FALSE,
                      scale = FALSE,
                      scannf = FALSE,
                      nf = 3)

barplot(WBNP.temp2.pca1$eig[1:50],
        main = "WBNP Temporal Group 2 PCA eigenvalues",
        col = heat.colors(50))

# % of variance explained by eigenvalues
WBNP.temp2.pc_percent <- 100 * WBNP.temp2.pca1$eig / sum(WBNP.temp2.pca1$eig)
round(WBNP.temp2.pc_percent[1:2], 2) # 2.78% and 2.24%

# PCA plot
col <- funky(15)
# Post-2016
s.class(WBNP.temp2.pca1$li,
        pop(WBNP.temp2),
        xax = 1,
        yax = 2,
        col = transp(col,.6), 
        axesell = FALSE,
        cstar = 0,
        cpoint = 3,
        grid = FALSE,
        sub = "PCA of WBNP from Temporal Group 2")


```
Note the extremely small variances in PC1 (2.78%) and PC2 (2.24%).

# STRUCTURE Bayesian Clustering

```{r}
temp1.struct.fix <- read.table("temp1.structure.txt",
           header = FALSE,
           skip = 1)

temp1.struct.fix <- temp1.struct.fix %>% 
  mutate(
    V2 = case_when(
    V2 == "EINPP"   ~ 1L,
    V2 == "EINPW"   ~ 2L,
    V2 == "WBNPPAD" ~ 3L,
    V2 == "WBNPGR" ~ 4L,
    V2 == "WBNPNY" ~ 5L,
    V2 == "WBNPLB" ~ 6L,
    V2 == "WBNPHC" ~ 7L
    )
  )
write.table(temp1.struct.fix, "temp1.struct.fix.txt", row.names = F)


```

```{r}
fastDivPart("genepop_files/temp1.genepop.txt", 
            "genepop_files/divout1.txt",
            gp = 3,
            pairwise = TRUE, 
            fst = TRUE, bs_locus = TRUE, bs_pairwise = TRUE,
            boots = 100, plot = FALSE, para = FALSE)
```

