---
title: "WBNP_PCA_STRUCT"
author: "Alex Depew"
date: "2026-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate the presence and extent of genetic substructure within Wood Buffalo National Park. 

Files used described below:

- `Revised_cleandata_07.03.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
library(ggplot2)
library(ggpubr)
library(grid)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_11.03.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
w1 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

w2 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.

```{r}
wbnp_subherds <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

wbnp_palette <- c("#cecece", 
                  "#a559aa", 
                  "#59a89c", 
                  "#f0c571", 
                  "#e02b35", 
                  "#082a54"
                  )

wbnp_cols <- setNames(as.character(wbnp_palette), wbnp_subherds)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.w1 <- tab(w1, freq = TRUE, NA.method = "mean")
pca.w1 <- dudi.pca(x.w1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w1 <- 100*pca.w1$eig/sum(pca.w1$eig)

x.w2 <- tab(w2, freq = TRUE, NA.method = "mean")
pca.w2 <- dudi.pca(x.w2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w2 <- 100*pca.w2$eig/sum(pca.w2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (W1: No EINP)
```{r}
w1@pop <- factor(w1@pop, levels = sort(unique(as.character(w1@pop))))
levels_w1 <- levels(w1@pop)
w1_cols <- wbnp_cols[levels_w1]


# Plot PCA object
pcaw1 <- fviz_pca_ind(pca.w1,
             axes = c(1, 2),
             geom = "blank",
             habillage = w1@pop,
             repel = TRUE,
             palette = w1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(colour = Groups), type = "norm", level = 0.68) + 
  geom_point(aes(colour = Groups),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.w1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaw1
```




```{r}
w2@pop <- factor(w2@pop, levels = sort(unique(as.character(w2@pop))))
levels_w2 <- levels(w2@pop)
w2_cols <- wbnp_cols[levels_w2]

# Plot PCA object
pcaw2 <- fviz_pca_ind(pca.w2,
             axes = c(1, 2),
             geom = "blank",
             habillage = w2@pop,
             repel = TRUE,
             palette = w2_cols,
             alpha.ind = 1) + 
  stat_ellipse(aes(color = w2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = w2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
  theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  ) + 

  labs(title = NULL,
       x = paste0("PC1 (", round(eig.perc.w2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

pcaw2
```

### Arrange Plots together
```{r}
pcaw1_nl <- pcaw1 + theme(legend.position = "none")
pcaw2_nl <- pcaw2 + theme(legend.position = "none")

legendw1 <- get_legend(pcaw1)

wbnp_pcas <- grid.arrange(pcaw1_nl, legendw1, pcaw2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_pcas.svg",
       plot = wbnp_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
## Add EINPW AND EINPP
#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
a1 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

a2 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.
```{r}
admix_subherds <- c("EINPP", "EINPW", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

admix_palette <- c("#FF00FF",
                   "#005000",
                   "#cecece", 
                   "#a559aa", 
                   "#59a89c", 
                   "#f0c571", 
                   "#e02b35", 
                   "#082a54"
                  )

admix_cols <- setNames(as.character(admix_palette), admix_subherds)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.a1 <- tab(a1, freq = TRUE, NA.method = "mean")
pca.a1 <- dudi.pca(x.a1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a1 <- 100*pca.a1$eig/sum(pca.a1$eig)

x.a2 <- tab(a2, freq = TRUE, NA.method = "mean")
pca.a2 <- dudi.pca(x.a2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a2 <- 100*pca.a2$eig/sum(pca.a2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (a1: With EINP)
```{r}
a1@pop <- factor(a1@pop, levels = sort(unique(as.character(a1@pop))))
levels_a1 <- levels(a1@pop)
a1_cols <- admix_cols[levels_a1]

pca.a1$li[ ,2] <- pca.a1$li[ ,2] * -1

# Plot PCA object
pcaa1 <- fviz_pca_ind(pca.a1,
             axes = c(1, 2),
             geom = "blank",
             habillage = a1@pop,
             repel = TRUE,
             palette = a1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa1
```


## PCA plot with ggplot
#### WBNP Temp1 (a2: With EINP)
```{r}
a2@pop <- factor(a2@pop, levels = sort(unique(as.character(a2@pop))))
levels_a2 <- levels(a2@pop)
a2_cols <- admix_cols[levels_a2]

pca.a2$li[ ,2] <- pca.a2$li[ ,2] * -1

# Plot PCA object
pcaa2 <- fviz_pca_ind(pca.a2,
             axes = c(1, 2),
             geom = "blank",
             habillage = a2@pop,
             repel = TRUE,
             palette = a2_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa2
```

### Arrange Plots together
```{r}
pcaa1_nl <- pcaa1 + theme(legend.position = "none")
pcaa2_nl <- pcaa2 + theme(legend.position = "none")

legenda1 <- get_legend(pcaa1)

admix_pcas <- grid.arrange(pcaa1_nl, legenda1, pcaa2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_admix_pcas.svg",
       plot = admix_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
# STRUCTURE Analysis
```{r}
# W1 structure (92 individuals, 6 pops)
w1.str <- w1 
pop(w1.str) <- as.numeric(as.factor(pop(w1.str)))

genind_to_structure(w1.str,
                    output = "STRUCTURE_files/w1.structure.txt")

# W2 structure (146 individuals, 6 pops)
w2.str <- w2
pop(w2.str) <- as.numeric(as.factor(pop(w2.str)))

genind_to_structure(w2.str,
                    output = "STRUCTURE_files/w2.structure.txt")

# A1 structure (161 individuals, 8 pops)
a1.str <- a1
pop(a1.str) <- as.numeric(as.factor(pop(a1.str)))

genind_to_structure(a1.str,
                    output = "STRUCTURE_files/a1.structure.txt")

# A2 structure (200 individuals, 8 pops)
a2.str <- a2
pop(a2.str) <- as.numeric(as.factor(pop(a2.str)))

genind_to_structure(a2.str,
                    output = "STRUCTURE_files/a2.structure.txt")

```
## Load Files
```{r}
# List files
# Full datasets
sfiles_w1 <- list.files("./STRUCTURE_files/results/wbnp1/", full.names = TRUE)
sfiles_w2 <- list.files("./STRUCTURE_files/results/wbnp2/", full.names = TRUE)

# Qlist (pophelper)
slist_w1 <- readQ(files = sfiles_w1, filetype = "structure")
slist_w2 <- readQ(files = sfiles_w2, filetype = "structure")

# Summarise runs
struct_sum_w1 <- summariseQ(tabulateQ(slist_w1))
struct_sum_w2 <- summariseQ(tabulateQ(slist_w2))

# And for admixed -----
# List files
# Full datasets
sfiles_a1 <- list.files("./STRUCTURE_files/results/admix1/", full.names = TRUE)
sfiles_a2 <- list.files("./STRUCTURE_files/results/admix2/", full.names = TRUE)

# Qlist (pophelper)
slist_a1 <- readQ(files = sfiles_a1, filetype = "structure")
slist_a2 <- readQ(files = sfiles_a2, filetype = "structure")

# Summarise runs
struct_sum_a1 <- summariseQ(tabulateQ(slist_a1))
struct_sum_a2 <- summariseQ(tabulateQ(slist_a2))
```


## Likelihood Plots
```{r}
## WBNP1 ----
# K plots
w1.k <- evannoMethodStructure(data = struct_sum_w1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w1.k$grobs[[1]]
p_dk <- w1.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_wbnp1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w1.k.svg", 
       plot = evanno_wbnp1, 
       width = 27, 
       height = 11)

# WBNP2 ----
# K plots
w2.k <- evannoMethodStructure(data = struct_sum_w2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w2.k$grobs[[1]]
p_dk <- w2.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_wbnp2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w2.k.svg", 
       plot = evanno_wbnp2, 
       width = 27, 
       height = 11)

## ADMIX1 ----
# K plots
a1.k <- evannoMethodStructure(data = struct_sum_a1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a1.k$grobs[[1]]
p_dk <- a1.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_admix1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a1.k.svg", 
       plot = evanno_admix1, 
       width = 27, 
       height = 11)

## ADMIX2 ----
# K plots
a2.k <- evannoMethodStructure(data = struct_sum_a2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a2.k$grobs[[1]]
p_dk <- a2.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_admix2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a2.k.svg", 
       plot = evanno_admix2, 
       width = 27, 
       height = 11)
```

#### WBNP 1
```{r}
w1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT"))
# Align runs
slist_w1 <- alignK(slist_w1, "auto")

# Merge runs (average over 10 replicates)
w1.mk2 <- mergeQ(slist_w1[11:20])
w1.mk3 <- mergeQ(slist_w1[21:30])
w1.mk4 <- mergeQ(slist_w1[31:40])
w1.mk5 <- mergeQ(slist_w1[41:50])
w1.mk6 <- mergeQ(slist_w1[51:60])

w1.merged.list <- as.qlist(list(w1.mk2[[1]], w1.mk3[[1]], w1.mk4[[1]], 
                          w1.mk5[[1]], w1.mk6[[1]]))

ind_label_w1 <- readQ(files = sfiles_w1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_w1 <- rownames_to_column(ind_label_w1, "ID")
locations_w1 <- left_join(ind_label_w1, w1.data, by = "ID")

loc_labels_w1 <- locations_w1$HERD %>% as.data.frame()

herd_order_w1 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_w1_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e")

plotQ(w1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_w1, 
            subsetgrp = herd_order_w1, 
            clustercol = struct_w1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "W1",
            exportpath = "./plots/structure_plots/")
```

#### WBNP 2
```{r}
w2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT"))
# Align runs
slist_w2 <- alignK(slist_w2, "auto")

# Merge runs (average over 10 replicates)
w2.mk2 <- mergeQ(slist_w2[11:20])
w2.mk3 <- mergeQ(slist_w2[21:30])
w2.mk4 <- mergeQ(slist_w2[31:40])
w2.mk5 <- mergeQ(slist_w2[41:50])
w2.mk6 <- mergeQ(slist_w2[51:60])

w2.merged.list <- as.qlist(list(w2.mk2[[1]], w2.mk3[[1]], w2.mk4[[1]], 
                          w2.mk5[[1]], w2.mk6[[1]]))

ind_label_w2 <- readQ(files = sfiles_w2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_w2 <- rownames_to_column(ind_label_w2, "ID")
locations_w2 <- left_join(ind_label_w2, w2.data, by = "ID")

loc_labels_w2 <- locations_w2$HERD %>% as.data.frame()

herd_order_w2 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_w2_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e")

plotQ(w2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_w2, 
            subsetgrp = herd_order_w2,
            clustercol = struct_w2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "W2",
            exportpath = "./plots/structure_plots/")
```

#### ADMIX 1
```{r}
a1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "EINPW", "EINPP"))
# Align runs
slist_a1 <- alignK(slist_a1, "auto")

# Merge runs (average over 10 replicates)
a1.mk2 <- mergeQ(slist_a1[11:20])
a1.mk3 <- mergeQ(slist_a1[21:30])
a1.mk4 <- mergeQ(slist_a1[31:40])
a1.mk5 <- mergeQ(slist_a1[41:50])
a1.mk6 <- mergeQ(slist_a1[51:60])
a1.mk7 <- mergeQ(slist_a1[61:70])
a1.mk8 <- mergeQ(slist_a1[71:80])


a1.merged.list <- as.qlist(list(a1.mk2[[1]], a1.mk3[[1]], a1.mk4[[1]], 
                          a1.mk5[[1]], a1.mk6[[1]], a1.mk7[[1]], a1.mk8[[1]]))

ind_label_a1 <- readQ(files = sfiles_a1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_a1 <- rownames_to_column(ind_label_a1, "ID")
locations_a1 <- left_join(ind_label_a1, a1.data, by = "ID")

loc_labels_a1 <- locations_a1$HERD %>% as.data.frame()

herd_order_a1 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT", "EINPW", "EINPP")
struct_a1_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e", "#f5df93","#858434")

plotQ(a1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_a1, 
            subsetgrp = herd_order_a1, 
            clustercol = struct_a1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "A1",
            exportpath = "./plots/structure_plots/")
```


#### ADMIX 2
```{r}
a2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "EINPW", "EINPP"))
# Align runs
slist_a2 <- alignK(slist_a2, "auto")

# Merge runs (average over 10 replicates)
a2.mk2 <- mergeQ(slist_a2[11:20])
a2.mk3 <- mergeQ(slist_a2[21:30])
a2.mk4 <- mergeQ(slist_a2[31:40])
a2.mk5 <- mergeQ(slist_a2[41:50])
a2.mk6 <- mergeQ(slist_a2[51:60])
a2.mk7 <- mergeQ(slist_a2[61:70])
a2.mk8 <- mergeQ(slist_a2[71:80])


a2.merged.list <- as.qlist(list(a2.mk2[[1]], a2.mk3[[1]], a2.mk4[[1]], 
                          a2.mk5[[1]], a2.mk6[[1]], a2.mk7[[1]], a2.mk8[[1]]))

ind_label_a2 <- readQ(files = sfiles_a2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_a2 <- rownames_to_column(ind_label_a2, "ID")
locations_a2 <- left_join(ind_label_a2, a2.data, by = "ID")

loc_labels_a2 <- locations_a2$HERD %>% as.data.frame()

herd_order_a2 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT", "EINPW", "EINPP")
struct_a2_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e", "#f5df93","#858434")

plotQ(a2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_a2, 
            subsetgrp = herd_order_a2, 
            clustercol = struct_a2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "A2",
            exportpath = "./plots/structure_plots/")
```



# Combined Plots
#### No EINP outgroups
```{r}
## WBNP - Wood bison only
plota1 <- plotQ(a1.mk3, 
            basesize = 10, 
            grplab = loc_labels_a1, 
            subsetgrp = herd_order_a1, 
            clustercol = struct_a1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 3",
            splabsize = 20,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "A1_temp",
            exportpath = "./STRUCTURE_files/")

# Pull STRUCTURE Plot from object
ta1 <- plota1$plot[[1]]


## Temp1 STRUCTURE
plota2 <- plotQ(a2.mk4, 
            basesize = 10, 
            grplab = loc_labels_a2, 
            subsetgrp = herd_order_a2, 
            clustercol = struct_a2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 4",
            splabsize = 20,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.15,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "A2_temp",
            exportpath = "./STRUCTURE_files/")

# Pull STRUCTURE Plot from object
ta2 <- plota2$plot[[1]]

m <- theme(plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm"))
pcaa1_tight <- pcaa1_nl + m
pcaa2_tight <- pcaa2_nl + m

pca.a2$li[ ,2] <- pca.a2[,2]$li * -1

# 2. Re-assemble the Bottom Section (PCAs)
# Using a slightly larger width for the legend to prevent crowding
bottom_section <- grid.arrange(
  pcaa1_tight, legenda1, pcaa2_tight, 
  ncol = 3, 
  widths = c(5, 1, 5)
)

# 3. Apply the 'A' Label
A_labeled <- arrangeGrob(bottom_section, 
                         top = textGrob("A", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

# 4. Apply 'B' and 'C' to STRUCTURE (Using your existing t1/t2 grobs)
B_labeled <- arrangeGrob(ta1, 
                         top = textGrob("B", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

C_labeled <- arrangeGrob(ta2, 
                         top = textGrob("C", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

# 5. Final Assembly with Reduced Vertical Gaps
# We combine Structure plots first to keep them tight together
top_section_labeled <- grid.arrange(B_labeled, C_labeled, ncol = 1, padding = unit(-2, "line"))

# Stack everything and use padding to pull A closer to B
WBNP_with_plains <- grid.arrange(A_labeled, top_section_labeled, 
                             ncol = 1, 
                             heights = c(1.3, 1.2), # Adjusts the ratio of PCA vs Structure
                             padding = unit(-6, "line"))

# 6. Save
ggsave("./plots/WBNP_with_plains_structure.svg", 
       plot = WBNP_with_plains, 
       width = 18, 
       height = 14, # Dropped height from 16 to 14 to force elements together
       units = "in")
```


#### WBNP No plains
```{r}
## WBNP - Wood bison only
plot1 <- plotQ(w1.mk2, 
            basesize = 10, 
            grplab = loc_labels_w1, 
            subsetgrp = herd_order_w1, 
            clustercol = struct_w1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 2",
            splabsize = 20,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "W1_temp",
            exportpath = "./STRUCTURE_files/")

# Pull STRUCTURE Plot from object
t1 <- plot1$plot[[1]]


## Temp1 STRUCTURE
plot2 <- plotQ(w2.mk2, 
            basesize = 10, 
            grplab = loc_labels_w2, 
            subsetgrp = herd_order_w2, 
            clustercol = struct_w2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = "K = 2",
            splabsize = 20,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.15,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "W2_temp",
            exportpath = "./STRUCTURE_files/")

# Pull STRUCTURE Plot from object
t2 <- plot2$plot[[1]]

m <- theme(plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm"))
pcaw1_tight <- pcaw1_nl + m
pcaw2_tight <- pcaw2_nl + m

# 2. Re-assemble the Bottom Section (PCAs)
# Using a slightly larger width for the legend to prevent crowding
bottom_section <- grid.arrange(
  pcaw1_tight, legendw1, pcaw2_tight, 
  ncol = 3, 
  widths = c(5, 1, 5)
)

# 3. Apply the 'A' Label
A_labeled <- arrangeGrob(bottom_section, 
                         top = textGrob("A", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

# 4. Apply 'B' and 'C' to STRUCTURE (Using your existing t1/t2 grobs)
B_labeled <- arrangeGrob(t1, 
                         top = textGrob("B", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

C_labeled <- arrangeGrob(t2, 
                         top = textGrob("C", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 28, fontface = "bold")))

# 5. Final Assembly with Reduced Vertical Gaps
# We combine Structure plots first to keep them tight together
top_section_labeled <- grid.arrange(B_labeled, C_labeled, ncol = 1, padding = unit(-2, "line"))

# Stack everything and use padding to pull A closer to B
WBNP_no_plains <- grid.arrange(A_labeled, top_section_labeled, 
                             ncol = 1, 
                             heights = c(1.3, 1.2), # Adjusts the ratio of PCA vs Structure
                             padding = unit(-6, "line"))

# 6. Save
ggsave("./plots/WBNP_noplains_structure.svg", 
       plot = WBNP_no_plains, 
       width = 18, 
       height = 14, # Dropped height from 16 to 14 to force elements together
       units = "in")
```

# DAPC

#### WBNP samples temp1
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp1 <- find.clusters(w1, max.n.clust = 8, n.pca = (nInd(w1) -1))

# Print BIC plot
svg(file = "./plots/WBNP1_v2_BIC.svg", width = 7, height = 5)
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 2015" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(w1), grp1$grp), col.lab=paste("inf", 1:6),
row.lab=paste("ori", 1:6))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx <- xvalDapc(tab(w1, NA.method = "mean"), pop(w1))

# Determine optimal number of PCs
numpc1 <- as.numeric(pramx$`Number of PCs Achieving Highest Mean Success`)

# DAPC with known parameters
dapc1 <- dapc(w1, pop(w1), n.pca = numpc1, n.da = (length(pop(w1))-1))
dapc1
```

#### WBNP Samples temp 2
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp2 <- find.clusters(w2, max.n.clust = 8, n.pca = (nInd(w2) -1))

# Print BIC plot
svg(file = "./plots/WBNP2_v2_BIC.svg", width = 7, height = 5)
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(w2), grp2$grp), col.lab=paste("inf", 1:6),
row.lab=paste("ori", 1:6))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx2 <- xvalDapc(tab(w2, NA.method = "mean"), pop(w2),
                   xval.plot = TRUE)

# Determine optimal number of PCs
numpc2 <- as.numeric(pramx2$`Number of PCs Achieving Highest Mean Success`)

# DAPC with known parameters
dapc2 <- dapc(w2, pop(w2), n.pca = numpc2, n.da = (length(pop(w2))-1))
dapc2
```

#### Plots BIC plots
```{r}
####### Plot ----
# Plot
svg(file = "./plots/wbnp_BIC_merged_v2.svg", width = 10, height = 5) 

# Set up  grid 1 row, 2 columns
par(mfrow = c(1, 2)) 

# Temp1
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 2015" ,
     col = "blue", pch = 1)

# Temp2
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)


dev.off()
```


```{r}
####### Plot ----
# Plot
svg(file = "./plots/wbnp_DAPC_stacked_v2.svg", width = 20, height = 10) 

# Set up  grid 1 row, 2 columns
par(mfrow = c(1, 2)) 

# Plot 1
scatter(dapc1,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, col=struct_w1_cols,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.13, csub = 2,
sub = paste0("DA1: ", round(dapc1$eig[1], 1), "% | DA2: ", round(dapc1$eig[2], 1), "%"))
mtext("A", side = 3, adj = -7.8, line = 2, cex = 2, font = 2)

# Plot 2
scatter(dapc2,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, col=struct_w1_cols,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.13, csub = 2,
sub = paste0("DA1: ", round(dapc2$eig[1], 1), "% | DA2: ", round(dapc2$eig[2], 1), "%"))
mtext("B", side = 3, adj = -7.8, line = 2, cex = 2, font = 2)

dev.off()
```


