---
title: "WBNP_PCA_STRUCT"
author: "Alex Depew"
date: "2026-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate population substructure in WBNP. 

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering

## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(dplyr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset Data
Two temporal groups of data exist for WBNP - those collected prior to 1999 and samples collected post-2016
```{r}
subherds <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")
W1 <- bison.genind[bison.genind@pop %in% subherds &
                     bison.genind@other$YEAR.ASSIGN <= 1999]

W2 <- bison.genind[bison.genind@pop %in% subherds &
                     bison.genind@other$YEAR.ASSIGN >= 2016]
```

## PCA

```{r}
clist <- list(
"shiny"=c("#1D72F5","#DF0101","#77CE61", "#FF9326","#A945FF","#0089B2","#FDF060","#FFA6B2","#BFF217","#60D5FD","#CC1577","#F2B950","#7FB21D","#EC496F","#326397","#B26314","#027368","#A4A4A4","#610B5E"),
"strong"=c("#11A4C8","#63C2C5","#1D4F9F","#0C516D","#2A2771","#396D35","#80C342","#725DA8","#B62025","#ED2224","#ED1943","#ED3995","#7E277C","#F7EC16","#F8941E","#8C2A1C","#808080"),
"oceanfive"=c("#00A0B0", "#6A4A3C", "#CC333F", "#EB6841", "#EDC951"),
"keeled"=c("#48B098", "#91CB62", "#FFEE3B", "#FB9013", "#FF3C28"),
"vintage"=c("#400F13", "#027368", "#A3BF3F", "#F2B950", "#D93A2B"),
"muted"=c("#46BDDD","#82DDCE","#F5F06A","#F5CC6A","#F57E6A"),
"teal"=c("#CFF09E","#A8DBA8","#79BD9A","#3B8686","#0B486B"),
"merry"=c("#5BC0EB","#FDE74C","#9BC53D","#E55934","#FA7921"),
"funky"=c("#A6CEE3", "#3F8EAA", "#79C360", "#E52829", "#FDB762","#ED8F47","#9471B4"),
"retro"=c("#01948E","#A9C4E2","#E23560","#01A7B3","#FDA963","#323665","#EC687D"),
"cb_paired"=c("#A6CEE3","#1F78B4","#B2DF8A","#33A02C","#FB9A99","#E31A1C","#FDBF6F","#FF7F00","#CAB2D6","#6A3D9A","#FFFF99","#B15928"),
"cb_set3"=c("#8DD3C7","#FFFFB3","#BEBADA","#FB8072","#80B1D3","#FDB462","#B3DE69","#FCCDE5","#D9D9D9","#BC80BD","#CCEBC5","#FFED6F"),
"morris"=c("#4D94CC","#34648A","#8B658A","#9ACD32","#CC95CC","#9ACD32","#8B3A39","#CD6601","#CC5C5B","#8A4500"),
"wong"=c("#000000","#E69F00","#56B4E9","#009E73","#F0E442","#006699","#D55E00","#CC79A7"),
"krzywinski"=c("#006E82","#8214A0","#005AC8","#00A0FA","#FA78FA","#14D2DC","#AA0A3C","#FA7850","#0AB45A","#F0F032","#A0FA82","#FAE6BE"))
```

```{r}
x.W1 <- tab(W1, freq = TRUE, NA.method = "mean")
pca.W1 <- dudi.pca(x.W1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.W1 <- 100*pca.W1$eig/sum(pca.W1$eig)

# Plot
png(filename = "./plots/pca_W1.png", width = 30, height = 30, units = "cm", res = 300)

s.class(pca.W1$li, fac = pop(W1), col = transp(clist$shiny, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.W1[1], 1), "% | PC2: ", round(eig.perc.W1[2], 1), "%"),)

add.scatter.eig(pca.W1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.W1$c1^2)
```

```{r}
x.W2 <- tab(W2, freq = TRUE, NA.method = "mean")
pca.W2 <- dudi.pca(x.W2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.W2 <- 100*pca.W2$eig/sum(pca.W2$eig)

# Plot
png(filename = "./plots/pca_W2.png", width = 30, height = 30, units = "cm", res = 300)

s.class(pca.W2$li, fac = pop(W2), col = transp(clist$shiny, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.W2[1], 1), "% | PC2: ", round(eig.perc.W2[2], 1), "%"),)

add.scatter.eig(pca.W2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.W2$c1^2)
```
## STRUCTURE 

# STRUCTURE Analyses
```{r}
# Populations need to be in unique integer format for input into STRUCTURE:
W1.struct.gen <- W1
pop(W1.struct.gen) <- as.numeric(as.factor(pop(W1.struct.gen)))
summary(W1.struct.gen)

W2.struct.gen <- W2
pop(W2.struct.gen) <- as.numeric(as.factor(pop(W2.struct.gen)))
summary(W2.struct.gen)


# Format in STRUCTURE format
library(graph4lg)
genind_to_structure(W1.struct.gen, output = "structure/W1.structure.txt")
genind_to_structure(W2.struct.gen, output = "structure/W2.structure.txt")
```

## Subset csv
The data csv file can be subset to match the samples included in each STRUCTURE run. This will be used to match the sample ID's for plotting on the STRUCTURE plot.
```{r}
W1.data <- bison.data %>%
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>%
  rename(ID = Spreadsheet_ID) %>%     
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% subherds)

W2.data <- temp2.data <- bison.data %>% 
  select(-ID) %>%
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% subherds)

```


### Temporal Group 1
Load Data
```{r}
sfiles_W1 <- list.files("./structure/STRUCT_W/W1", full.names = TRUE)
slist_W1 <- readQ(files = sfiles_W1, filetype = "structure")
```

Optimal K clusters
```{r}
library(gridExtra)
struct_sum_W1 <- summariseQ(tabulateQ(slist_W1))

p1 <- evannoMethodStructure(data = struct_sum_W1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "evannoMethodStructure_W1",
                           exportpath = "./plots/structure_plots/")
grid.arrange(p1)
```
```{r}
# Print summary of runs (Evanno method plot is based on these values)
summariseQ(tabulateQ(slist_temp1))

# Individual 
ind_label_W1 <- readQ(files = sfiles_W1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_W1 <- rownames_to_column(ind_label_W1, "ID")

locations_W1 <- left_join(ind_label_W1, W1.data, by = "ID")

loc_labels_W1 <- locations_W1$HERD %>% as.data.frame()
colnames(loc_labels_W1) <- "Region"

herd_order_W1 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD")

# Align and merge across K
m_k2 <- mergeQ(alignK(slist_W1[11:20])) # This aggregates all 10 replicates of K = 2

# Structure Plot
plotQ(m_k2,
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_W1, 
            subsetgrp = herd_order_W1, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "W1_K2",
            exportpath = "./plots/structure_plots/")

```

### Temporal Group 2
Load Data
```{r}
sfiles_W2 <- list.files("./structure/STRUCT_W/W2", full.names = TRUE)
slist_W2 <- readQ(files = sfiles_W2, filetype = "structure")
```

Optimal K clusters
```{r}
library(gridExtra)
struct_sum_W2 <- summariseQ(tabulateQ(slist_W2))

p2 <- evannoMethodStructure(data = struct_sum_W2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "evannoMethodStructure_W2",
                           exportpath = "./plots/structure_plots/")
grid.arrange(p2)
```
```{r}
# Print summary of runs (Evanno method plot is based on these values)
summariseQ(tabulateQ(slist_W2))

# Individual 
ind_label_W2 <- readQ(files = sfiles_W2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_W2 <- rownames_to_column(ind_label_W2, "ID")

locations_W2 <- left_join(ind_label_W2, W2.data, by = "ID")

loc_labels_W2 <- locations_W2$HERD %>% as.data.frame()
colnames(loc_labels_W2) <- "Region"

herd_order_W2 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD") # Sort North-South to identify any IBD patterns

# Align and merge across K
m2_k2 <- mergeQ(alignK(slist_W2[11:20])) # This aggregates all 10 replicates of K = 2

# Structure Plot
plotQ(m2_k2,
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_W2, 
            subsetgrp = herd_order_W2, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "W2_K2",
            exportpath = "./plots/structure_plots/")
```


# Admixture (Plains)
Tested with EINPP, EINPW, and WBNP subherds
```{r}
library(tidyverse)
# Create Structure files for testing admixture
admix_herds <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD")

admix_W1 <- bison.genind[bison.genind@pop %in% admix_herds &
                           bison.genind@other$YEAR.ASSIGN <= 1999]
pop(admix_W1) <- as.numeric(as.factor(pop(admix_W1)))

admix_W2 <- bison.genind[bison.genind@pop %in% admix_herds &
                           bison.genind@other$YEAR.ASSIGN >= 2016]
pop(admix_W2) <- as.numeric(as.factor(pop(admix_W2)))


# Write to STRUCTURE format
genind_to_structure(admix_W1,
                    output = "./structure/admix_W1.txt")

genind_to_structure(admix_W2,
                    output = "./structure/admix_W2.txt")

# Create Datasets subset from bison.data
T1.data <- bison.data %>%
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>%
  rename(ID = Spreadsheet_ID) %>%     
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% admix_herds)

T2.data <- bison.data %>% 
  select(-ID) %>%
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% admix_herds)
```







### Read Files
```{r}
sfiles_T1 <- list.files("./structure/STRUCT_W_ADMIX/T1", full.names = TRUE)
slist_T1 <- readQ(sfiles_T1, filetype = "structure")
```

### Find optimal K clusters
```{r}
struct_sum_T1 <- summariseQ(tabulateQ(slist_T1))

p_ad1 <- evannoMethodStructure(data = struct_sum_T1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "evannoMethodStructure_T1",
                           exportpath = "./plots/structure_plots/")
grid.arrange(p_ad1)
```
Ln(K) indicates optimal K clusters at 3, whereas deltaK indicating at 2 (likely high level structure ie. plains and wood)

```{r}
# Print summary of runs (Evanno method plot is based on these values)
summariseQ(tabulateQ(slist_T1))

# Individual 
ind_label_T1 <- readQ(files = sfiles_T1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_T1 <- rownames_to_column(ind_label_T1, "ID")

locations_T1 <- left_join(ind_label_T1, T1.data, by = "ID")

loc_labels_T1 <- locations_T1$HERD %>% as.data.frame()
colnames(loc_labels_T1) <- "Region"

herd_order_T1 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD") # In theory - northern herds should have more introgression

# Align and merge across K
mad1_k3 <- mergeQ(alignK(slist_T1[21:30])) # This aggregates all 10 replicates of K = 3

# Structure Plot
plotQ(mad1_k3,
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_T1, 
            subsetgrp = herd_order_T1, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "T1_K3",
            exportpath = "./plots/structure_plots/")
```

### Read Files
```{r}
sfiles_T2 <- list.files("./structure/STRUCT_W_ADMIX/T2", full.names = TRUE)
slist_T2 <- readQ(sfiles_T2, filetype = "structure")
```

### Find optimal K clusters
```{r}
struct_sum_T2 <- summariseQ(tabulateQ(slist_T2))

p_ad2 <- evannoMethodStructure(data = struct_sum_T2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "evannoMethodStructure_T2",
                           exportpath = "./plots/structure_plots/")
grid.arrange(p_ad2)
```

```{r}
# Print summary of runs (Evanno method plot is based on these values)
summariseQ(tabulateQ(slist_T2))

# Individual 
ind_label_T2 <- readQ(files = sfiles_T2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_T2 <- rownames_to_column(ind_label_T2, "ID")

locations_T2 <- left_join(ind_label_T2, T2.data, by = "ID")

loc_labels_T2 <- locations_T2$HERD %>% as.data.frame()
colnames(loc_labels_T2) <- "Region"

herd_order_T2 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPGR", "WBNPPAD") # In theory - northern herds should have more introgression

# Align and merge across K
mad2_k3 <- mergeQ(alignK(slist_T2[21:30])) # This aggregates all 10 replicates of K = 4

# Structure Plot
k3 <- plotQ(mad2_k3,
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_T2, 
            subsetgrp = herd_order_T2, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "T2_K3",
            exportpath = "./plots/structure_plots/")

# Align and merge across K
mad2_k4 <- mergeQ(alignK(slist_T2[31:40])) # This aggregates all 10 replicates of K = 4

k4 <- plotQ(mad2_k4,
            basesize = 10, 
            clustercol = clist$shiny, 
            grplab = loc_labels_T2, 
            subsetgrp = herd_order_T2, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 5, 
            grplabangle = 45,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            splabsize = 17,
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "T2_K4",
            exportpath = "./plots/structure_plots/")

```

```{r}
# Subset Genind
T1 <- bison.genind[bison.genind@pop %in% admix_herds &
                     bison.genind@other$YEAR.ASSIGN <= 1999]

x.T1 <- tab(T1, freq = TRUE, NA.method = "mean")
pca.T1 <- dudi.pca(x.T1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.T1 <- 100*pca.T1$eig/sum(pca.T1$eig)

# Plot
png(filename = "./plots/pca_T1.png", width = 30, height = 30, units = "cm", res = 300)

s.class(pca.T1$li, fac = pop(T1), col = transp(clist$shiny, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.T1[1], 1), "% | PC2: ", round(eig.perc.T1[2], 1), "%"),)

add.scatter.eig(pca.T1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.T1$c1^2)
```

```{r}
# Subset Genind
T2 <- bison.genind[bison.genind@pop %in% admix_herds &
                     bison.genind@other$YEAR.ASSIGN >= 2016]

x.T2 <- tab(T2, freq = TRUE, NA.method = "mean")
pca.T2 <- dudi.pca(x.T2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.T2 <- 100*pca.T2$eig/sum(pca.T2$eig)

# Plot
png(filename = "./plots/pca_T2.png", width = 30, height = 30, units = "cm", res = 300)

s.class(pca.T2$li, fac = pop(T2), col = transp(clist$shiny, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.T2[1], 1), "% | PC2: ", round(eig.perc.T2[2], 1), "%"),)

add.scatter.eig(pca.T2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.T2$c1^2)
```



