---
title: "WBNP_PCA_STRUCT"
author: "Alex Depew"
date: "2026-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate the presence and extent of genetic substructure within Wood Buffalo National Park. 

Files used described below:

- `Revised_cleandata_07.03.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
library(ggplot2)
library(ggpubr)
library(grid)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_07.03.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
w1 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

w2 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.

```{r}
wbnp_subherds <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

wbnp_palette <- c("#cecece", 
                  "#a559aa", 
                  "#59a89c", 
                  "#f0c571", 
                  "#e02b35", 
                  "#082a54"
                  )

wbnp_cols <- setNames(as.character(wbnp_palette), wbnp_subherds)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.w1 <- tab(w1, freq = TRUE, NA.method = "mean")
pca.w1 <- dudi.pca(x.w1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w1 <- 100*pca.w1$eig/sum(pca.w1$eig)

x.w2 <- tab(w2, freq = TRUE, NA.method = "mean")
pca.w2 <- dudi.pca(x.w2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w2 <- 100*pca.w2$eig/sum(pca.w2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (W1: No EINP)
```{r}
w1@pop <- factor(w1@pop, levels = sort(unique(as.character(w1@pop))))
levels_w1 <- levels(w1@pop)
w1_cols <- wbnp_cols[levels_w1]


# Plot PCA object
pcaw1 <- fviz_pca_ind(pca.w1,
             axes = c(1, 2),
             geom = "blank",
             habillage = w1@pop,
             repel = TRUE,
             palette = w1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = w1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = w1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.w1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaw1
```




```{r}
w2@pop <- factor(w2@pop, levels = sort(unique(as.character(w2@pop))))
levels_w2 <- levels(w2@pop)
w2_cols <- wbnp_cols[levels_w2]

# Plot PCA object
pcaw2 <- fviz_pca_ind(pca.w2,
             axes = c(1, 2),
             geom = "blank",
             habillage = w2@pop,
             repel = TRUE,
             palette = w2_cols,
             alpha.ind = 1) + 
  stat_ellipse(aes(color = w2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = w2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
  theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  ) + 

  labs(title = NULL,
       x = paste0("PC1 (", round(eig.perc.w2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

pcaw2
```

### Arrange Plots together
```{r}
pcaw1_nl <- pcaw1 + theme(legend.position = "none")
pcaw2_nl <- pcaw2 + theme(legend.position = "none")

legend <- get_legend(pcaw1)

wbnp_pcas <- grid.arrange(pcaw1_nl, legend, pcaw2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_pcas.svg",
       plot = wbnp_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
## Add EINPW AND EINPP
#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
a1 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

a2 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.
```{r}
admix_subherds <- c("EINPP", "EINPW", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

admix_palette <- c("#FF00FF",
                   "#005000",
                   "#cecece", 
                   "#a559aa", 
                   "#59a89c", 
                   "#f0c571", 
                   "#e02b35", 
                   "#082a54"
                  )

admix_cols <- setNames(as.character(admix_palette), admix_subherds)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.a1 <- tab(a1, freq = TRUE, NA.method = "mean")
pca.a1 <- dudi.pca(x.a1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a1 <- 100*pca.a1$eig/sum(pca.a1$eig)

x.a2 <- tab(a2, freq = TRUE, NA.method = "mean")
pca.a2 <- dudi.pca(x.a2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a2 <- 100*pca.a2$eig/sum(pca.a2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (a1: With EINP)
```{r}
a1@pop <- factor(a1@pop, levels = sort(unique(as.character(a1@pop))))
levels_a1 <- levels(a1@pop)
a1_cols <- admix_cols[levels_a1]

pca.a1$li[ ,2] <- pca.a1$li[ ,2] * -1

# Plot PCA object
pcaa1 <- fviz_pca_ind(pca.a1,
             axes = c(1, 2),
             geom = "blank",
             habillage = a1@pop,
             repel = TRUE,
             palette = a1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa1
```


## PCA plot with ggplot
#### WBNP Temp1 (a2: With EINP)
```{r}
a2@pop <- factor(a2@pop, levels = sort(unique(as.character(a2@pop))))
levels_a2 <- levels(a2@pop)
a2_cols <- admix_cols[levels_a2]

pca.a2$li[ ,2] <- pca.a2$li[ ,2] * -1

# Plot PCA object
pcaa2 <- fviz_pca_ind(pca.a2,
             axes = c(1, 2),
             geom = "blank",
             habillage = a2@pop,
             repel = TRUE,
             palette = a2_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold"), # Kept bold but removed blank
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa2
```

### Arrange Plots together
```{r}
pcaa1_nl <- pcaa1 + theme(legend.position = "none")
pcaa2_nl <- pcaa2 + theme(legend.position = "none")

legend <- get_legend(pcaa1)

admix_pcas <- grid.arrange(pcaa1_nl, legend, pcaa2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_admix_pcas.svg",
       plot = admix_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
# STRUCTURE Analysis
```{r}
# W1 structure (92 individuals, 6 pops)
w1.str <- w1 
pop(w1.str) <- as.numeric(as.factor(pop(w1.str)))

genind_to_structure(w1.str,
                    output = "STRUCTURE_files/w1.structure.txt")

# W2 structure (146 individuals, 6 pops)
w2.str <- w2
pop(w2.str) <- as.numeric(as.factor(pop(w2.str)))

genind_to_structure(w2.str,
                    output = "STRUCTURE_files/w2.structure.txt")

# A1 structure (161 individuals, 8 pops)
a1.str <- a1
pop(a1.str) <- as.numeric(as.factor(pop(a1.str)))

genind_to_structure(a1.str,
                    output = "STRUCTURE_files/a1.structure.txt")

# A2 structure (200 individuals, 8 pops)
a2.str <- a2
pop(a2.str) <- as.numeric(as.factor(pop(a2.str)))

genind_to_structure(a2.str,
                    output = "STRUCTURE_files/a2.structure.txt")

```
## Load Files
```{r}
# List files
# Full datasets
sfiles_w1 <- list.files("./STRUCTURE_files/results/wbnp1/", full.names = TRUE)
sfiles_w2 <- list.files("./STRUCTURE_files/results/wbnp2/", full.names = TRUE)

# Qlist (pophelper)
slist_w1 <- readQ(files = sfiles_w1, filetype = "structure")
slist_w2 <- readQ(files = sfiles_w2, filetype = "structure")

# Summarise runs
struct_sum_w1 <- summariseQ(tabulateQ(slist_w1))
struct_sum_w2 <- summariseQ(tabulateQ(slist_w2))

# And for admixed -----
# List files
# Full datasets
sfiles_a1 <- list.files("./STRUCTURE_files/results/admix1/", full.names = TRUE)
sfiles_a2 <- list.files("./STRUCTURE_files/results/admix2/", full.names = TRUE)

# Qlist (pophelper)
slist_a1 <- readQ(files = sfiles_a1, filetype = "structure")
slist_a2 <- readQ(files = sfiles_a2, filetype = "structure")

# Summarise runs
struct_sum_a1 <- summariseQ(tabulateQ(slist_a1))
struct_sum_a2 <- summariseQ(tabulateQ(slist_a2))
```


## Likelihood Plots
```{r}
## WBNP1 ----
# K plots
w1 <- evannoMethodStructure(data = struct_sum_w1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w1$grobs[[1]]
p_dk <- w1$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_wbnp1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w1.k.svg", 
       plot = evanno_wbnp1, 
       width = 27, 
       height = 11)

# WBNP2 ----
# K plots
w2 <- evannoMethodStructure(data = struct_sum_w2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w2$grobs[[1]]
p_dk <- w2$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_wbnp2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w2.k.svg", 
       plot = evanno_wbnp2, 
       width = 27, 
       height = 11)

## ADMIX1 ----
# K plots
a1 <- evannoMethodStructure(data = struct_sum_a1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a1$grobs[[1]]
p_dk <- a1$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_admix1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a1.k.svg", 
       plot = evanno_admix1, 
       width = 27, 
       height = 11)

## ADMIX2 ----
# K plots
a2 <- evannoMethodStructure(data = struct_sum_a2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a2$grobs[[1]]
p_dk <- a2$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(1, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 18, fontface = "bold")))

evanno_admix2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a2.k.svg", 
       plot = evanno_admix2, 
       width = 27, 
       height = 11)
```


