---
title: "WBNP_PCA_STRUCT"
author: "Alex Depew"
date: "2026-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate population substructure in WBNP. 

Files used described below:

- `Cleaned_samples.06.01.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering

## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(dplyr)
```

## Load Required Data
```{r}
bison.data <- read.csv("Cleaned_samples.06.01.2026.csv",
                       check.names = F)
```


#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=bison.data[,c(19:ncol(bison.data))],
                              sep = "/",
                              ind.names = bison.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = bison.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), bison.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- bison.data$YEAR.ASSIGN[idx]

# Remove loci with more than 20% missing data
bison.genind <- missingno(bison.genind, type = "loci", cutoff = 0.20)
```

#### Subset Data
Two temporal groups of data exist for WBNP - those collected prior to 1999 and samples collected post-2016
```{r}
subherds <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")
W1 <- bison.genind[bison.genind@pop %in% subherds &
                     bison.genind@other$YEAR.ASSIGN <= 1999]

W2 <- bison.genind[bison.genind@pop %in% subherds &
                     bison.genind@other$YEAR.ASSIGN >= 2016]
```

## PCA

```{r}
custom_palette <- c(
  "#E31A1C", "#1F78B4", "#33A02C", "#FF7F00", "#6A3D9A", 
  "#B15928", "#A6CEE3", "#B2DF8A", "#FB9A99", "#FDBF6F", 
  "#CAB2D6", "#FFFF99", "#000000", "#8DD3C7", "#BEBADA"  
)
```

```{r}
x.W1 <- tab(W1, freq = TRUE, NA.method = "mean")
pca.W1 <- dudi.pca(x.W1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.W1 <- 100*pca.W1$eig/sum(pca.W1$eig)

# Plot
pdf(file = "./plots/pca_W1.pdf", width = 10, height = 10)

s.class(pca.W1$li, fac = pop(W1), col = transp(custom_palette, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.W1[1], 1), "% | PC2: ", round(eig.perc.W1[2], 1), "%"),)

add.scatter.eig(pca.W1$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.W1$c1^2)
```

```{r}
x.W2 <- tab(W2, freq = TRUE, NA.method = "mean")
pca.W2 <- dudi.pca(x.W2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.W2 <- 100*pca.W2$eig/sum(pca.W2$eig)

# Plot
pdf(file = "./plots/pca_W2.pdf", width = 10, height = 10)

s.class(pca.W2$li, fac = pop(W2), col = transp(custom_palette, 0.75),
        axesell = F, cstar = 0, cpoint = 3,
        grid = F,
        sub = paste0("PC1: ", round(eig.perc.W2[1], 1), "% | PC2: ", round(eig.perc.W2[2], 1), "%"),)

add.scatter.eig(pca.W2$eig[1:20], 3, 1, 2, ratio = 0.15, "bottomright")
dev.off()



# Allele contributions to variance
loadingplot(pca.W2$c1^2)
```
## STRUCTURE 

#### Select Colour Palette for STRUCTURE Plotting
```{r}
colours <- c(
  "#E41A1C", 
  "#377EB8", 
  "#4DAF4A", 
  "#FF7F00", 
  "#984EA3", 
  "#FFFF33", 
  "#A65628", 
  "#F781BF", 
  "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#666666"
)
```

# STRUCTURE Analyses
```{r}
# Populations need to be in unique integer format for input into STRUCTURE:
W1.struct.gen <- W1
pop(W1.struct.gen) <- as.numeric(as.factor(pop(W1.struct.gen)))
summary(W1.struct.gen)

W2.struct.gen <- W2
pop(W2.struct.gen) <- as.numeric(as.factor(pop(W2.struct.gen)))
summary(W2.struct.gen)


# Format in STRUCTURE format
library(graph4lg)
genind_to_structure(W1.struct.gen, output = "structure/W1.structure.txt")
genind_to_structure(W2.struct.gen, output = "structure/W2.structure.txt")
```

## Subset csv
The data csv file can be subset to match the samples included in each STRUCTURE run. This will be used to match the sample ID's for plotting on the STRUCTURE plot.
```{r}
W1.data <- bison.data %>%
  select(-ID) %>% 
  filter(YEAR.ASSIGN <= 1999) %>%
  rename(ID = Spreadsheet_ID) %>%     
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% subherds)

W2.data <- temp2.data <- bison.data %>% 
  select(-ID) %>%
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% subherds)

```


### Temporal Group 1
Load Data
```{r}
sfiles_W1 <- list.files("./structure/STRUCT_W/W1", full.names = TRUE)
slist_W1 <- readQ(files = sfiles_W1, filetype = "structure")
```

Optimal K clusters
```{r}
library(gridExtra)
struct_sum_W1 <- summariseQ(tabulateQ(slist_W1))

p1 <- evannoMethodStructure(data = struct_sum_W1, exportplot = F,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7)
grid.arrange(p1)
```
### Plotting
```{r, results='hide'}
library(tidyverse)
ind_label_W1 <- readQ(files = sfiles_W1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_W1 <- rownames_to_column(ind_label_W1, "ID")

locations_W1 <- left_join(ind_label_W1, W1.data, by = "ID")

# 
slist <- slist_W1

W1_ind_order <- locations_W1
rownames(slist[[1]]) <- W1_ind_order$ID
if(length(unique(sapply(slist,nrow)))==1) slist <- lapply(slist,"rownames<-",W1_ind_order$ID)
lapply(slist, rownames)[1:50]

slist_W1 <- slist 
slist_align1 <- alignK(slist_W1, "within") # Align replicate runs
merged_W1 <- mergeQ(slist_W1[1:10]) # Merge K = 2 runs

loc_labels_W1 <- locations_W1$HERD %>% as.data.frame()
colnames(loc_labels_W1) <- "Region"

herd_order_W1 <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")


plotQ(slist_align1[c(5,15,25,35,45)], 
      imgoutput = "join",
            basesize = 10, 
            clustercol = colours, 
            grplab = loc_labels_W1, 
            subsetgrp = herd_order_W1, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 3, 
            grplabangle = 0,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "W1_K2",
            exportpath = "./plots/structure_plots/")

```

### Temporal Group 2
Load Data
```{r}
sfiles_W2 <- list.files("./structure/STRUCT_W/W2", full.names = TRUE)
slist_W2 <- readQ(files = sfiles_W2, filetype = "structure")
```

Optimal K clusters
```{r}
library(gridExtra)
struct_sum_W2 <- summariseQ(tabulateQ(slist_W2))

p2 <- evannoMethodStructure(data = struct_sum_W2, exportplot = F,
                           returnplot = T, returndata = F,
                           basesize = 12, linesize = 0.7)
grid.arrange(p2)
```
### Plotting
```{r, results='hide'}
library(tidyverse)
ind_label_W2 <- readQ(files = sfiles_W2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_W2 <- rownames_to_column(ind_label_W2, "ID")

locations_W2 <- left_join(ind_label_W2, W2.data, by = "ID")

# 
slist <- slist_W2

W2_ind_order <- locations_W2
rownames(slist[[1]]) <- W2_ind_order$ID
if(length(unique(sapply(slist,nrow)))==1) slist <- lapply(slist,"rownames<-",W2_ind_order$ID)
lapply(slist, rownames)[1:50]

slist_W2 <- slist 
slist_align2 <- alignK(slist_W2, "within") # Align replicate runs
merged_W2 <- mergeQ(slist_W2[11:20]) # Merge K = 2 runs

loc_labels_W2 <- locations_W2$HERD %>% as.data.frame()
colnames(loc_labels_W2) <- "Region"

herd_order_W2 <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD")


plotQ(slist_align2[c(5,15,25,35,45)], 
      imgoutput = "join",
            basesize = 10, 
            clustercol = colours, 
            grplab = loc_labels_W2, 
            subsetgrp = herd_order_W2, 
            ordergrp = TRUE,
            showindlab = TRUE, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,       
            grplabsize = 3, 
            grplabangle = 0,
            grplabpos = 0.65,
            grplabheight = 0.2,      
            grplabspacer = 0.1,
            linesize = 1,          
            pointsize = 4, 
            divcol = "white",
            divsize = 1, 
            useindlab = TRUE, 
            returndata = TRUE,
            height = 12,             
            width = 35,
            outputfilename = "W2_K2",
            exportpath = "./plots/structure_plots/")

```