---
title: "WBNP_PCA_STRUCT"
author: "Alex Depew"
date: "2026-03-08"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File information
This script is used to evaluate the presence and extent of genetic substructure within Wood Buffalo National Park. 

Files used described below:

- `Revised_cleandata_07.03.2026.csv`
   - Contains all samples for population genetic analyses, post-filtering
   
## Load Required Packages
```{r}
rm(list = ls())
library(adegenet)
library(graph4lg)
library(poppr)
library(hierfstat)
library(pophelper)
library(tidyverse)
library(gridExtra)
library(factoextra)
library(ggplot2)
library(ggpubr)
library(grid)
```

## Load Required Data
```{r}
cleaned.data <- read.csv("./datafiles/Revised_cleandata_11.03.2026.csv",
                         check.names = F)
```

#### Convert data to Genind object
```{r}
# Make genind object
bison.genind <- df2genind(X=cleaned.data[,c(19:ncol(cleaned.data))],
                              sep = "/",
                              ind.names = cleaned.data$Spreadsheet_ID,
                              loc.names = NULL,
                              pop = cleaned.data$HERD,
                              NA.char = "NA",
                              ploidy = 2,
                              type = "codom",
                              strata = NULL,
                              hierarchy = NULL)

# Attach collection year metadata
idx <- match(indNames(bison.genind), cleaned.data$Spreadsheet_ID)
stopifnot(!any(is.na(idx)))
bison.genind@other$YEAR.ASSIGN <- cleaned.data$YEAR.ASSIGN[idx]
```

#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
w1 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

w2 <- bison.genind[bison.genind@pop %in% c("WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.

```{r}
wbnp_subherds <- c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

wbnp_palette <- c("#cecece", 
                  "#a559aa", 
                  "#59a89c", 
                  "#f0c571", 
                  "#e02b35", 
                  "#082a54"
                  )

wbnp_cols <- setNames(as.character(wbnp_palette), wbnp_subherds)

# STRUCTURE colours
master_cols <- c("#ee964c", "#3b96cf", "#7d2a7c", "#e39bc1", "#e73f4e", 
                 "#4b9b46", "#f5df93", "#858434", "#ba552b", "#81c7c7", 
                 "#2d4a5e", "#44c29a", "#542e44", "#ff7f32", "#607d8b", "#1b4332")

```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.w1 <- tab(w1, freq = TRUE, NA.method = "mean")
pca.w1 <- dudi.pca(x.w1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w1 <- 100*pca.w1$eig/sum(pca.w1$eig)

x.w2 <- tab(w2, freq = TRUE, NA.method = "mean")
pca.w2 <- dudi.pca(x.w2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.w2 <- 100*pca.w2$eig/sum(pca.w2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (W1: No EINP)
```{r}
w1@pop <- factor(w1@pop, levels = sort(unique(as.character(w1@pop))))
levels_w1 <- levels(w1@pop)
w1_cols <- wbnp_cols[levels_w1]


# Plot PCA object
pcaw1 <- fviz_pca_ind(pca.w1,
             axes = c(1, 2),
             geom = "blank",
             habillage = w1@pop,
             repel = TRUE,
             palette = w1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(colour = Groups), type = "norm", level = 0.68) + 
  geom_point(aes(colour = Groups),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 

  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.w1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaw1
```




```{r}
w2@pop <- factor(w2@pop, levels = sort(unique(as.character(w2@pop))))
levels_w2 <- levels(w2@pop)
w2_cols <- wbnp_cols[levels_w2]

# Plot PCA object
pcaw2 <- fviz_pca_ind(pca.w2,
             axes = c(1, 2),
             geom = "blank",
             habillage = w2@pop,
             repel = TRUE,
             palette = w2_cols,
             alpha.ind = 1) + 
  stat_ellipse(aes(color = w2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = w2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
  coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 


  labs(title = NULL,
       x = paste0("PC1 (", round(eig.perc.w2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.w2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

pcaw2
```

### Arrange Plots together
```{r}
pcaw1_nl <- pcaw1 + theme(legend.position = "none")
pcaw2_nl <- pcaw2 + theme(legend.position = "none")

legendw1 <- get_legend(pcaw1)

wbnp_pcas <- grid.arrange(pcaw1_nl, legendw1, pcaw2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_pcas.svg",
       plot = wbnp_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
## Add EINPW AND EINPP
#### Subset data 
Data is subset into two temporal groups. Temporal groups are: 1: pre-2016, and 3: post-2016
```{r}
# Create datasets containing samples of interest for temporal comparisons
a1 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN <=2015,]

a2 <- bison.genind[bison.genind@pop %in% c("EINPP", "EINPW", "WBNPGR", "WBNPHC", 
                                           "WBNPLB", "WBNPNY", "WBNPPAD", "WENT") &
                   bison.genind@other$YEAR.ASSIGN >=2016,]
```

# Principal Coordinate Analysis (PCA)
Visualize population structure without assumptions of HWE or LD, to identify clusters in the data.
```{r}
admix_subherds <- c("EINPP", "EINPW", "WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT")

admix_palette <- c("#FF00FF",
                   "#005000",
                   "#cecece", 
                   "#a559aa", 
                   "#59a89c", 
                   "#f0c571", 
                   "#e02b35", 
                   "#082a54"
                  )

admix_cols <- setNames(as.character(admix_palette), admix_subherds)
```

# PCA
### Perform PCA on both datasets
```{r}
# Perform PCA
x.a1 <- tab(a1, freq = TRUE, NA.method = "mean")
pca.a1 <- dudi.pca(x.a1, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a1 <- 100*pca.a1$eig/sum(pca.a1$eig)

x.a2 <- tab(a2, freq = TRUE, NA.method = "mean")
pca.a2 <- dudi.pca(x.a2, center = T, scale = F, scannf = F, nf = 2)

# Axes % Variance
eig.perc.a2 <- 100*pca.a2$eig/sum(pca.a2$eig)

#### axis limits
xlims <- c(-2.5, 3.5)
ylims <- c(-2, 2.1)
```

## PCA plot with ggplot
#### WBNP Temp1 (a1: With EINP)
```{r}
a1@pop <- factor(a1@pop, levels = sort(unique(as.character(a1@pop))))
levels_a1 <- levels(a1@pop)
a1_cols <- admix_cols[levels_a1]

pca.a1$li[ ,2] <- pca.a1$li[ ,2] * -1

# Plot PCA object
pcaa1 <- fviz_pca_ind(pca.a1,
             axes = c(1, 2),
             geom = "blank",
             habillage = a1@pop,
             repel = TRUE,
             palette = a1_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a1@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a1@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 


  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a1[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a1[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa1
```


## PCA plot with ggplot
#### WBNP Temp1 (a2: With EINP)
```{r}
a2@pop <- factor(a2@pop, levels = sort(unique(as.character(a2@pop))))
levels_a2 <- levels(a2@pop)
a2_cols <- admix_cols[levels_a2]

pca.a2$li[ ,2] <- pca.a2$li[ ,2] * -1

# Plot PCA object
pcaa2 <- fviz_pca_ind(pca.a2,
             axes = c(1, 2),
             geom = "blank",
             habillage = a2@pop,
             repel = TRUE,
             palette = a2_cols,
             alpha.ind = 1,) + 
  stat_ellipse(aes(color = a2@pop), type = "norm", level = 0.68) + 
  geom_point(aes(colour = a2@pop),
             shape = 19,
             alpha = 0.75,
             size = 1.2) + 
coord_fixed(ratio = 1, xlim = xlims, ylim = ylims) +
  
theme(
    plot.title = element_blank(),
    legend.title = element_text(face = "bold", size = 15),
    legend.text = element_text(size = 15),
    axis.title = element_text(face = "bold", size = 20),
    text = element_text(size = 20),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),  ) + 


  labs(title = NULL,
       # Fixed the placement of closing parentheses and commas below
       x = paste0("PC1 (", round(eig.perc.a2[1], 2), "%)"),
       y = paste0("PC2 (", round(eig.perc.a2[2], 2), "%)"),
       colour = "Herd") + 
    guides(colour = guide_legend(override.aes = list(size = 4, shape = 19)))

       

pcaa2
```

### Arrange Plots together
```{r}
pcaa1_nl <- pcaa1 + theme(legend.position = "none")
pcaa2_nl <- pcaa2 + theme(legend.position = "none")

legenda1 <- get_legend(pcaa1)

admix_pcas <- grid.arrange(pcaa1_nl, legenda1, pcaa2_nl, # Set out grid
                            ncol = 3, 
                            widths = c(10, 2.5, 10)
                            )
                            
# Save plot
library(svglite)
ggsave("./plots/wbnp_admix_pcas.svg",
       plot = admix_pcas, 
       device = "svg", 
       width = 16, 
       height = 12, 
       units = "in")
```
# STRUCTURE Analysis
```{r}
# W1 structure (92 individuals, 6 pops)
w1.str <- w1 
pop(w1.str) <- as.numeric(as.factor(pop(w1.str)))

genind_to_structure(w1.str,
                    output = "STRUCTURE_files/w1.structure.txt")

# W2 structure (146 individuals, 6 pops)
w2.str <- w2
pop(w2.str) <- as.numeric(as.factor(pop(w2.str)))

genind_to_structure(w2.str,
                    output = "STRUCTURE_files/w2.structure.txt")

# A1 structure (161 individuals, 8 pops)
a1.str <- a1
pop(a1.str) <- as.numeric(as.factor(pop(a1.str)))

genind_to_structure(a1.str,
                    output = "STRUCTURE_files/a1.structure.txt")

# A2 structure (200 individuals, 8 pops)
a2.str <- a2
pop(a2.str) <- as.numeric(as.factor(pop(a2.str)))

genind_to_structure(a2.str,
                    output = "STRUCTURE_files/a2.structure.txt")

```
## Load Files
```{r}
# List files
# Full datasets
sfiles_w1 <- list.files("./STRUCTURE_files/results/wbnp1/", full.names = TRUE)
sfiles_w2 <- list.files("./STRUCTURE_files/results/wbnp2/", full.names = TRUE)

# Qlist (pophelper)
slist_w1 <- readQ(files = sfiles_w1, filetype = "structure")
slist_w2 <- readQ(files = sfiles_w2, filetype = "structure")

# Summarise runs
struct_sum_w1 <- summariseQ(tabulateQ(slist_w1))
struct_sum_w2 <- summariseQ(tabulateQ(slist_w2))

# And for admixed -----
# List files
# Full datasets
sfiles_a1 <- list.files("./STRUCTURE_files/results/admix1/", full.names = TRUE)
sfiles_a2 <- list.files("./STRUCTURE_files/results/admix2/", full.names = TRUE)

# Qlist (pophelper)
slist_a1 <- readQ(files = sfiles_a1, filetype = "structure")
slist_a2 <- readQ(files = sfiles_a2, filetype = "structure")

# Summarise runs
struct_sum_a1 <- summariseQ(tabulateQ(slist_a1))
struct_sum_a2 <- summariseQ(tabulateQ(slist_a2))
```


## Likelihood Plots
```{r}
## WBNP1 ----
# K plots
w1.k <- evannoMethodStructure(data = struct_sum_w1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w1.k$grobs[[1]]
p_dk <- w1.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_wbnp1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w1.k.svg", 
       plot = evanno_wbnp1, 
       width = 27, 
       height = 11)

# WBNP2 ----
# K plots
w2.k <- evannoMethodStructure(data = struct_sum_w2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "w2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- w2.k$grobs[[1]]
p_dk <- w2.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_wbnp2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/w2.k.svg", 
       plot = evanno_wbnp2, 
       width = 27, 
       height = 11)

## ADMIX1 ----
# K plots
a1.k <- evannoMethodStructure(data = struct_sum_a1, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a1_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a1.k$grobs[[1]]
p_dk <- a1.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_admix1 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a1.k.svg", 
       plot = evanno_admix1, 
       width = 27, 
       height = 11)

## ADMIX2 ----
# K plots
a2.k <- evannoMethodStructure(data = struct_sum_a2, exportplot = T,
                           returnplot = T, returndata = F,
                           basesize = 25, linesize = 0.7,
                           height = 25, width = 25, units = "cm",
                           outputfilename = "a2_K",
                           exportpath = "./STRUCTURE_files/",
                           imgtype = "pdf")

p_lik <- a2.k$grobs[[1]]
p_dk <- a2.k$grobs[[4]]


# Label Likelihood as 'a'
panel_a <- arrangeGrob(p_lik, top = textGrob("A", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

# Label Delta K as 'b'
panel_b <- arrangeGrob(p_dk, top = textGrob("B", x = unit(0.5, "line"), y = unit(0.8, "npc"), 
                                             just = c("left", "top"), 
                                             gp = gpar(fontsize = 35, fontface = "bold")))

evanno_admix2 <- grid.arrange(panel_a, panel_b, 
                         ncol = 2, 
                         widths = c(1, 1))

ggsave("./STRUCTURE_files/a2.k.svg", 
       plot = evanno_admix2, 
       width = 27, 
       height = 11)
```

#### WBNP 1
```{r}
w1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT"))
# Align runs
slist_w1 <- alignK(slist_w1, "auto")

# Merge runs (average over 10 replicates)
w1.mk2 <- mergeQ(slist_w1[11:20])
w1.mk3 <- mergeQ(slist_w1[21:30])
w1.mk4 <- mergeQ(slist_w1[31:40])
w1.mk5 <- mergeQ(slist_w1[41:50])
w1.mk6 <- mergeQ(slist_w1[51:60])

w1.merged.list <- as.qlist(list(w1.mk2[[1]], w1.mk3[[1]], w1.mk4[[1]], 
                          w1.mk5[[1]], w1.mk6[[1]]))

ind_label_w1 <- readQ(files = sfiles_w1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_w1 <- rownames_to_column(ind_label_w1, "ID")
locations_w1 <- left_join(ind_label_w1, w1.data, by = "ID")

loc_labels_w1 <- locations_w1$HERD %>% as.data.frame()

herd_order_w1 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_w1_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e")

plotQ(w1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_w1, 
            subsetgrp = herd_order_w1, 
            clustercol = struct_w1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "W1",
            exportpath = "./plots/structure_plots/")
```

#### WBNP 2
```{r}
w2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT"))
# Align runs
slist_w2 <- alignK(slist_w2, "auto")

# Merge runs (average over 10 replicates)
w2.mk2 <- mergeQ(slist_w2[11:20])
w2.mk3 <- mergeQ(slist_w2[21:30])
w2.mk4 <- mergeQ(slist_w2[31:40])
w2.mk5 <- mergeQ(slist_w2[41:50])
w2.mk6 <- mergeQ(slist_w2[51:60])

w2.merged.list <- as.qlist(list(w2.mk2[[1]], w2.mk3[[1]], w2.mk4[[1]], 
                          w2.mk5[[1]], w2.mk6[[1]]))

ind_label_w2 <- readQ(files = sfiles_w2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_w2 <- rownames_to_column(ind_label_w2, "ID")
locations_w2 <- left_join(ind_label_w2, w2.data, by = "ID")

loc_labels_w2 <- locations_w2$HERD %>% as.data.frame()

herd_order_w2 <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_w2_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e")

plotQ(w2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_w2, 
            subsetgrp = herd_order_w2,
            clustercol = struct_w2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "W2",
            exportpath = "./plots/structure_plots/")
```

#### ADMIX 1
```{r}
a1.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN <=2015) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "EINPW", "EINPP"))
# Align runs
slist_a1 <- alignK(slist_a1, "auto")

# Merge runs (average over 10 replicates)
a1.mk2 <- mergeQ(slist_a1[11:20])
a1.mk3 <- mergeQ(slist_a1[21:30])
a1.mk4 <- mergeQ(slist_a1[31:40])
a1.mk5 <- mergeQ(slist_a1[41:50])
a1.mk6 <- mergeQ(slist_a1[51:60])
a1.mk7 <- mergeQ(slist_a1[61:70])
a1.mk8 <- mergeQ(slist_a1[71:80])


a1.merged.list <- as.qlist(list(a1.mk2[[1]], a1.mk3[[1]], a1.mk4[[1]], 
                          a1.mk5[[1]], a1.mk6[[1]], a1.mk7[[1]], a1.mk8[[1]]))

ind_label_a1 <- readQ(files = sfiles_a1[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_a1 <- rownames_to_column(ind_label_a1, "ID")
locations_a1 <- left_join(ind_label_a1, a1.data, by = "ID")

loc_labels_a1 <- locations_a1$HERD %>% as.data.frame()

herd_order_a1 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_a1_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e", "#f5df93","#858434")

plotQ(a1.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_a1, 
            subsetgrp = herd_order_a1, 
            clustercol = struct_a1_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "A1",
            exportpath = "./plots/structure_plots/")
```


#### ADMIX 2
```{r}
a2.data <- cleaned.data %>% 
  select(-ID) %>% 
  filter(YEAR.ASSIGN >=2016) %>%
  rename(ID = Spreadsheet_ID) %>% 
  mutate(ID = as.character(ID)) %>% 
  filter(HERD %in% c("WBNPGR", "WBNPHC", "WBNPLB", "WBNPNY", "WBNPPAD", "WENT", "EINPW", "EINPP"))
# Align runs
slist_a2 <- alignK(slist_a2, "auto")

# Merge runs (average over 10 replicates)
a2.mk2 <- mergeQ(slist_a2[11:20])
a2.mk3 <- mergeQ(slist_a2[21:30])
a2.mk4 <- mergeQ(slist_a2[31:40])
a2.mk5 <- mergeQ(slist_a2[41:50])
a2.mk6 <- mergeQ(slist_a2[51:60])
a2.mk7 <- mergeQ(slist_a2[61:70])
a2.mk8 <- mergeQ(slist_a2[71:80])


a2.merged.list <- as.qlist(list(a2.mk2[[1]], a2.mk3[[1]], a2.mk4[[1]], 
                          a2.mk5[[1]], a2.mk6[[1]], a2.mk7[[1]], a2.mk8[[1]]))

ind_label_a2 <- readQ(files = sfiles_a2[1], indlabfrom = TRUE) %>% as.data.frame()
ind_label_a2 <- rownames_to_column(ind_label_a2, "ID")
locations_a2 <- left_join(ind_label_a2, a2.data, by = "ID")

loc_labels_a2 <- locations_a2$HERD %>% as.data.frame()

herd_order_a2 <- c("EINPP", "EINPW", "WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT")
struct_a2_cols <- c("#ee964c", "#3b96cf", "#4b9b46", "#7d2a7c", "#e39bc1", "#e73f4e", "#f5df93","#858434")

plotQ(a2.merged.list, 
      imgoutput = "join",
            basesize = 10, 
            grplab = loc_labels_a2, 
            subsetgrp = herd_order_a2, 
            clustercol = struct_a2_cols,
            ordergrp = TRUE,
      
      # Side "K" Labels
            splab = c("K = 2", "K = 3", "K = 4", "K = 5", "K = 6", "K = 7", "K = 8"),
            splabsize = 35,
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
            # Individual Label Settings
            showindlab = F, 
            indlabsize = 4,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 10, 
            grplabangle =45,
            grplabpos = 0.65,
            grplabheight = 1,      
            grplabspacer = 0.1,
            grplabcol = "black",
      
            linesize = 2,          
            pointsize = 15,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 3, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Plot settings
            height = 7,             
            width = 50,
            panelspacer = 1,
            panelratio = c(1,1),
      
      
      # File settings
            imgtype = "tiff",
            outputfilename = "A2",
            exportpath = "./plots/structure_plots/")
```



# Combined Plots
```{r}
### PCA w/ plains bison outgroup

# STRUCTURE A1 w/ plains
plota1 <- plotQ(c(a1.mk2, a1.mk3, a1.mk4, a1.mk5, a1.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_a1, 
            subsetgrp = herd_order_a1, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3*", "K=4", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plota1.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
ta1 <- plota1$plot[[1]]

# Combine Global w/ plains T1
row_1 <- grid.arrange(pcaa1_nl, ta1,
                      ncol = 2,
                      widths = c(10, 15))


# STRUCTURE T1 w/ plains
plota2 <- plotQ(c(a2.mk2, a2.mk3, a2.mk4, a2.mk5, a2.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_a2, 
            subsetgrp = herd_order_a2, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4*", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plota2.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
ta2 <- plota2$plot[[1]]


# Combine Global w/ plains T1
row_2 <- grid.arrange(pcaa2_nl, ta2,
                      ncol = 2,
                      widths = c(10, 15))

################ Without Plains from Elk Island

# STRUCTURE W1 w/o plains
plotw1 <- plotQ(c(w1.mk2, w1.mk3, w1.mk4, w1.mk5, w1.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_w1, 
            subsetgrp = herd_order_w1, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3", "K=4", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plotw1.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
tw1 <- plotw1$plot[[1]]

# Combine Global w/ plains T1
row_3 <- grid.arrange(pcaw1_nl, tw1,
                      ncol = 2,
                      widths = c(10, 15))

# STRUCTURE W2 w/o plains
plotw2<- plotQ(c(w2.mk2, w2.mk3, w2.mk4, w2.mk5, w2.mk6),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_w2, 
            subsetgrp = herd_order_w2, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2*", "K=3", "K=4", "K=5", "K=6"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plotw2.k2-6",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
tw2<- plotw2$plot[[1]]

# Combine Global w/ plains T1
row_4 <- grid.arrange(pcaw2_nl, tw2,
                      ncol = 2,
                      widths = c(10, 15))


#### Label Sections
# 3. Apply the 'A' Label
A_labeled <- arrangeGrob(row_1, 
                         top = textGrob("A (T1: 1986 - 1999)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

B_labeled <- arrangeGrob(row_2, 
                         top = textGrob("B (T2: 2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

top_half <- arrangeGrob(A_labeled, 
                        rectGrob(gp = gpar(col = NA)), # Spacer between rows
                        B_labeled, 
                        ncol = 1, 
                        heights = c(10, 1.5, 10),
                        top = textGrob("Including EINP (Wood and Plains) Outgroup", 
                                       gp = gpar(fontsize = 35, fontface = "bold")))

C_labeled <- arrangeGrob(row_3, 
                         top = textGrob("C (T1: 1986 - 1999)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

D_labeled <- arrangeGrob(row_4, 
                         top = textGrob("D (T2: 2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold"))) 

bottom_half <- arrangeGrob(C_labeled, 
                           rectGrob(gp = gpar(col = NA)), # Spacer between rows
                           D_labeled, 
                           ncol = 1, 
                           heights = c(10, 1.5, 10), 
                           top = textGrob("Excluding EINP (Wood and Plains) Outgroup", 
                                          gp = gpar(fontsize = 35, fontface = "bold")))

#### Combine PCAs and STRUCTURE barplots
comb <- arrangeGrob(top_half, 
                    rectGrob(gp = gpar(col = NA)), # Blank space
                    bottom_half,
                    ncol = 1, 
                    heights = c(20, 2, 20)) # Give the middle some space to account for the separating line


#### Pull legend from global PCA
pcaa1_horizontal <- pcaa1 + 
  theme(legend.direction = "horizontal", 
        legend.position = "top",

       legend.justification = "left", 
        legend.box.just = "top",
        legend.text = element_text(size = 20), 
        legend.title = element_text(size = 25, face = "bold"),
        legend.margin = margin(l = 0)) + 
  guides(color = guide_legend(
    title = "Herd", 
    nrow = 3, 
    byrow = TRUE,
    override.aes = list(size = 7, alpha = 1) 
  ))

# Isolate legend
legend_bottom <- get_legend(pcaa1_horizontal)

# Legend move to left
legend_shifted <- arrangeGrob(legend_bottom, 
                              vp = viewport(x = 0.08, just = "left", width = 0.8)) 

# Combine Structuring and legend
comb_leg <- grid.arrange(
  comb, 
  legend_shifted,
  ncol = 1, 
  heights = c(46, 4), 
  padding = unit(0, "line")
)

comb_leg_sep <-grobTree(comb_leg, linesGrob(x = unit(c(0, 1), "npc"), 
                                       y = unit(0.543, "npc"), 
                                       gp = gpar(col = "black", lwd = 2)))
 

# Save plots
ggsave("./plots/WBNP_leg_sep.svg",
       plot = comb_leg_sep,
       height = 25,
       width = 20,
       bg = "white",
       units = "in")
```

# DAPC

#### WBNP samples temp1
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp1 <- find.clusters(w1, max.n.clust = 8, n.pca = (nInd(w1) -1))

# Print BIC plot
svg(file = "./plots/WBNP1_v2_BIC.svg", width = 7, height = 5)
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 2015" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(w1), grp1$grp), col.lab=paste("inf", 1:6),
row.lab=paste("ori", 1:6))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx <- xvalDapc(tab(w1, NA.method = "mean"), pop(w1))

# Determine optimal number of PCs
numpc1 <- as.numeric(pramx$`Number of PCs Achieving Highest Mean Success`)

# DAPC with known parameters
dapc1 <- dapc(w1, pop(w1), n.pca = numpc1, n.da = (length(pop(w1))-1))
dapc1
```

#### WBNP Samples temp 2
```{r echo=TRUE}
## Unsupervised K means clustering
# Identify K Clusters 
grp2 <- find.clusters(w2, max.n.clust = 8, n.pca = (nInd(w2) -1))

# Print BIC plot
svg(file = "./plots/WBNP2_v2_BIC.svg", width = 7, height = 5)
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)
dev.off()

# BIC minimum value is at K =1
table.value(table(pop(w2), grp2$grp), col.lab=paste("inf", 1:6),
row.lab=paste("ori", 1:6))

## Supervised DAPC
# TO prevent overfitting of the model, find optimal PCs to retain (Max = n-1 individuals)
set.seed(999)
pramx2 <- xvalDapc(tab(w2, NA.method = "mean"), pop(w2),
                   xval.plot = TRUE)

# Determine optimal number of PCs
numpc2 <- as.numeric(pramx2$`Number of PCs Achieving Highest Mean Success`)

# DAPC with known parameters
dapc2 <- dapc(w2, pop(w2), n.pca = numpc2, n.da = (length(pop(w2))-1))
dapc2
```

#### Plots BIC plots
```{r}
####### Plot ----
# Plot
svg(file = "./plots/wbnp_BIC_merged_v2.svg", width = 10, height = 5) 

# Set up  grid 1 row, 2 columns
par(mfrow = c(1, 2)) 

# Temp1
plot(grp1$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 1986 - 2015" ,
     col = "blue", pch = 1)

# Temp2
plot(grp2$Kstat, type = "o", xlab = "Number of clusters (K)", 
     ylab = "BIC", main = "Value of BIC \n versus number of clusters \n WBNP: 2016 - 2025" ,
     col = "blue", pch = 1)


dev.off()
```


```{r}
####### Plot ----
dapc_palette <- c("#cecece", 
                  "#a559aa", 
                  "#59a89c", 
                  "#f0c571", 
                  "#e02b35", 
                  "#082a54"
                  )

# Plot
svg(file = "./plots/wbnp_DAPC_stacked_v2.svg", width = 20, height = 10) 

# Set up  grid 1 row, 2 columns
par(mfrow = c(1, 2)) 

# Plot 1
scatter(dapc1,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, cex = 2, col=dapc_palette,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.13, csub = 2,
sub = paste0("DA1: ", round(dapc1$eig[1], 1), "% | DA2: ", round(dapc1$eig[2], 1), "%"))
mtext("A", side = 3, adj = -7.8, line = 2, cex = 2, font = 2)

# Plot 2
scatter(dapc2,posi.da="bottomright", bg="white",
pch=17:22,cstar=0, cex = 2,col=dapc_palette,scree.pca=TRUE,
posi.pca = "topright", ratio.da = 0.20, ratio.pca = 0.13, csub = 2,
sub = paste0("DA1: ", round(dapc2$eig[1], 1), "% | DA2: ", round(dapc2$eig[2], 1), "%"))
mtext("B", side = 3, adj = -7.8, line = 2, cex = 2, font = 2)

dev.off()
```
# Plains Admixture
#### Assess Q-values (membership probabilities of assigning to clusters)
```{r}
a1.order <- data.frame(plota1$data$grplab[[1]]) %>% 
  rownames_to_column("Ind")

a1.k2 <- data.frame(a1.mk2) %>% 
  rownames_to_column("Ind") %>% 
  left_join(a1.order, by = "Ind") %>% 
  rename(Herd = ".",
        Wood = X2.Cluster2,
         Plains = X2.Cluster1) %>% 
  select(-Ind) %>% 
  relocate(Herd, .before = Plains)
  
a1.Q.sum <- a1.k2 %>% 
  group_by(Herd) %>% 
  summarise(
    QWood = paste0(round(mean(Wood), 3), "±", round(sd(Wood), 3)),
    QPlains = paste0(round(mean(Plains), 3), "±", round(sd(Plains), 3))
  )

# Summary per herd of Q assignments
a1.Q.sum

# Just WBNP
a1.wbnp.Q.sum <-a1.Q.sum[3:8, ] 
mean(a1.wbnp.Q.sum$Plains_mean)

write.csv(a1.wbnp.Q.sum, "./tables/A1.Q.plains_assign.csv")


#-------------------


a2.order <- data.frame(plota2$data$grplab[[1]]) %>% 
  rownames_to_column("Ind")

a2.k2 <- data.frame(a2.mk2) %>% 
  rownames_to_column("Ind") %>% 
  left_join(a2.order, by = "Ind") %>% 
  rename(Herd = ".",
         Wood = X2.Cluster1, # Swapped from previous
         Plains = X2.Cluster2) %>% 
  select(-Ind) %>% 
  relocate(Herd, .before = Wood)
  
a2.Q.sum <- a2.k2 %>% 
  group_by(Herd) %>% 
  summarise(
    QWood = paste0(round(mean(Wood), 3), "±", round(sd(Wood), 3)),
    QPlains = paste0(round(mean(Plains), 3), "±", round(sd(Plains), 3))
  )

# Summary per herd of Q assignments
a2.Q.sum

a2.wbnp.Q.sum <-a2.Q.sum[3:8, ] 
mean(a2.wbnp.Q.sum$Plains_mean)


write.csv(a2.wbnp.Q.sum, "./tables/A2.Q.plains_assign.csv")


```
#### JUST WBNP Plains (but grouped as all herds)
```{r}
# Identify the names of the WBNP herds based on your subsetting (rows 3 to 8)
wbnp_herds <- a1.Q.sum$Herd[3:8]

# Filter the original individual data for these herds and calculate the overall stats
a1.wbnp.overall_stats <- a1.k2 %>%
  filter(Herd %in% wbnp_herds) %>%
  summarise(
    Overall_Mean_Plains = mean(Plains, na.rm = TRUE),
    Overall_SD_Plains = sd(Plains, na.rm = TRUE)
  )

# View the result
a1.wbnp.overall_stats


# Filter the original individual data for these herds and calculate the overall stats
a2.wbnp.overall_stats <- a2.k2 %>%
  filter(Herd %in% wbnp_herds) %>%
  summarise(
    Overall_Mean_Plains = mean(Plains, na.rm = TRUE),
    Overall_SD_Plains = sd(Plains, na.rm = TRUE)
  )

# View the result
a2.wbnp.overall_stats
```




# WBNP Split
```{r}
w2.order <- data.frame(plotw2$data$grplab[[2]]) %>% 
  rownames_to_column("Ind")

w2.k2 <- data.frame(w2.mk2) %>% 
  rownames_to_column("Ind") %>% 
  left_join(w2.order, by = "Ind") %>% 
  rename(Herd = ".",
        North = X2.Cluster2,
         Central = X2.Cluster1) %>% 
  select(-Ind) %>% 
  relocate(Herd, .before = North)
  
w2.Q.sum <- w2.k2 %>% 
  group_by(Herd) %>% 
  summarise(
    QC1 = paste0(round(mean(North), 3), "±", round(sd(North), 3)),
    QC2 = paste0(round(mean(Central), 3), "±", round(sd(Central), 3))
  )

# Summary per herd of Q assignments
w2.Q.sum

write.csv(w2.Q.sum, "./tables/WBNP.Cluster.assignment.csv")
```

# Combined Plots
```{r}
### PCA w/ plains bison outgroup

herd_order <- c("WBNPLB", "WBNPNY", "WBNPHC", "WBNPPAD", "WBNPGR","WENT","EINPP", "EINPW")
master_cols <- c("#ee964c", "#3b96cf", "#7d2a7c","#4b9b46", "#e39bc1", "#e73f4e")
master_cols2 <- c("#ee964c", "#3b96cf", "#7d2a7c", "#4b9b46", "#e39bc1", "#e73f4e")

# Flip clusters so they plot correctly
a1.mk2$`2` <- a1.mk2$`2`[, c("Cluster2", "Cluster1")]
colnames(a1.mk2$`2`) <- c("Cluster1", "Cluster2")

# STRUCTURE A1 w/ plains
plota1 <- plotQ(a1.mk2,
            basesize = 10, 
            grplab = loc_labels_a1, 
            subsetgrp = herd_order, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = "K=2",
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plota1.k2",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
ta1 <- plota1$plot[[1]]

# Combine Global w/ plains T1
row_1 <- grid.arrange(pcaa1_nl, ta1,
                      ncol = 2,
                      widths = c(10, 15))


# STRUCTURE T1 w/ plains
plota2 <- plotQ(a2.mk2,
            basesize = 10, 
            grplab = loc_labels_a2, 
            subsetgrp = herd_order, 
            clustercol = master_cols2,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = "K=2",
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plota2.k2",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
ta2 <- plota2$plot[[1]]


# Combine Global w/ plains T1
row_2 <- grid.arrange(pcaa2_nl, ta2,
                      ncol = 2,
                      widths = c(10, 15))

################ Without Plains from Elk Island

# STRUCTURE W1 w/o plains
plotw1 <- plotQ(c(w1.mk2, w1.mk3),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_w1, 
            subsetgrp = herd_order, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2", "K=3"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plotw1.k2-3",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
tw1 <- plotw1$plot[[1]]

# Combine Global w/ plains T1
row_3 <- grid.arrange(pcaw1_nl, tw1,
                      ncol = 2,
                      widths = c(10, 15))

# STRUCTURE W2 w/o plains
plotw2<- plotQ(c(w2.mk2, w2.mk3),
            imgout = "join",
            basesize = 10, 
            grplab = loc_labels_w2, 
            subsetgrp = herd_order, 
            clustercol = master_cols,
            ordergrp = TRUE,

      # Side "K" Labels
            splab = c("K=2*", "K=3"),
            splabsize = 20,
            splabface = "bold",
            sppos = "left",
            splabcol = "black",
            splabangle = 0,
      
      # Individual Label Settings
            showindlab = F, 
            indlabsize = 0,        
            indlabangle = 90,
            indlabheight = 0,
      
      # Group Label settings
            grplabsize = 6.2, 
            grplabangle = 45,
            grplabpos = 0.85,
            grplabheight = 0.6,      
            grplabspacer = 0,
            grplabcol = "black",
      
            linesize = 1,          
            pointsize = 8,
            linepos = 0.95,
      
      # Divider Line Settings
            divcol = "white",
            divsize = 1, 
            divtype = NULL,
      
            useindlab = TRUE, 
            returndata = TRUE,
      
      # Add title
           showtitle = TRUE,
           titlelab = NULL,
           titlehjust = 0.5,
           titlevjust = 0.5,
           titlesize = 15,
           titlecol = "black",
      titleface = "bold",
      
      # Plot settings
            height = 5,             
            width = 50,
            panelspacer = 0.2,
            panelratio = c(1,2),
      
      # File settings
            imgtype = "tiff",
            returnplot = TRUE,
            outputfilename = "plotw2.k2-3",
            exportpath = "./plots/temp")

# Pull STRUCTURE Plot from object
tw2<- plotw2$plot[[1]]

# Combine Global w/ plains T1
row_4 <- grid.arrange(pcaw2_nl, tw2,
                      ncol = 2,
                      widths = c(10, 15))


#### Label Sections
# 3. Apply the 'A' Label
A_labeled <- arrangeGrob(row_1, 
                         top = textGrob("A (T1: 1986 - 1999)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

B_labeled <- arrangeGrob(row_2, 
                         top = textGrob("B (T2: 2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

top_half <- arrangeGrob(A_labeled, 
                        rectGrob(gp = gpar(col = NA)), # Spacer between rows
                        B_labeled, 
                        ncol = 1, 
                        heights = c(10, 1.5, 10),
                        top = textGrob("Including EINP (Wood and Plains) Outgroup", 
                                       gp = gpar(fontsize = 35, fontface = "bold")))

C_labeled <- arrangeGrob(row_3, 
                         top = textGrob("C (T1: 1986 - 1999)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold")))

D_labeled <- arrangeGrob(row_4, 
                         top = textGrob("D (T2: 2016 - 2025)", x = unit(0.02, "npc"), just = "left", 
                                        gp = gpar(fontsize = 35, fontface = "bold"))) 

bottom_half <- arrangeGrob(C_labeled, 
                           rectGrob(gp = gpar(col = NA)), # Spacer between rows
                           D_labeled, 
                           ncol = 1, 
                           heights = c(10, 1.5, 10), 
                           top = textGrob("Excluding EINP (Wood and Plains) Outgroup", 
                                          gp = gpar(fontsize = 35, fontface = "bold")))

#### Combine PCAs and STRUCTURE barplots
comb <- arrangeGrob(top_half, 
                    rectGrob(gp = gpar(col = NA)), # Blank space
                    bottom_half,
                    ncol = 1, 
                    heights = c(20, 2, 20)) # Give the middle some space to account for the separating line


#### Pull legend from global PCA
pcaa1_horizontal <- pcaa1 + 
  theme(legend.direction = "horizontal", 
        legend.position = "top",

       legend.justification = "left", 
        legend.box.just = "top",
        legend.text = element_text(size = 20), 
        legend.title = element_text(size = 25, face = "bold"),
        legend.margin = margin(l = 0)) + 
  guides(color = guide_legend(
    title = "Herd", 
    nrow = 3, 
    byrow = TRUE,
    override.aes = list(size = 7, alpha = 1) 
  ))

# Isolate legend
legend_bottom <- get_legend(pcaa1_horizontal)

# Legend move to left
legend_shifted <- arrangeGrob(legend_bottom, 
                              vp = viewport(x = 0.08, just = "left", width = 0.8)) 

# Combine Structuring and legend
comb_leg <- grid.arrange(
  comb, 
  legend_shifted,
  ncol = 1, 
  heights = c(46, 4), 
  padding = unit(0, "line")
)

comb_leg_sep <-grobTree(comb_leg, linesGrob(x = unit(c(0, 1), "npc"), 
                                       y = unit(0.543, "npc"), 
                                       gp = gpar(col = "black", lwd = 2)))
 

# Save plots
ggsave("./plots/WBNP_leg_sep_revised.svg",
       plot = comb_leg_sep,
       height = 25,
       width = 20,
       bg = "white",
       units = "in")
```
